---
title: "Daily measurement plotting"
author: "Zoe Dellaert "
date: "5/25/2025"
output:
  html_document: default
  pdf_document: default
---

This script plots daily measurements from the experiment, and is based on the Putnam Lab script available here: https://github.com/Putnam-Lab/CBLS_Wetlab/tree/main

## Load packages 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate) # used for converting 8 digit date into datetime format for R
library(RColorBrewer)
library(rmarkdown)
library(tinytex)

## If seacarb needs to be downloaded:
#packageurl <- "https://cran.r-project.org/src/contrib/Archive/seacarb/seacarb_3.2.tar.gz"
#install.packages(packageurl, repos = NULL, type = "source")
#install.packages("seacarb")
library(seacarb) 
```

## Load data 
```{r}
## Read in data
daily <- read.csv("../data/water_chemistry/DMs.csv")
head(daily)
tail(daily) # check to make sure data from today is there

daily$Date <- as.Date(as.character(daily$Date), format = "%Y%m%d")
daily$tris.date <- as.character(daily$tris.date)
daily$Tank_ID <- as.character(daily$Tank_ID)

daily$DateTime <- as.POSIXct(paste(daily$Date, daily$Time), format="%Y-%m-%d %H:%M")

```

```{r}
daily.probe1 <- daily %>% filter(Probe.Set == "Probe1")  %>% filter(Salinity_ATC_Mode=="MAN")

range(na.omit(daily.probe1$Temperature_C))
range(na.omit(daily.probe1$pH_mv))
range(na.omit(daily.probe1$Salinity_psu))
```

## Calculate total pH from Probe Set 1 

Calculate the calibration curve from the Tris calibration and calculate pH on the total scale from pH.mV. 
```{r}
pHcalib <- read_csv("../data/water_chemistry/Tris_Calibration.csv")
pHcalib$tris.date<-as.character(pHcalib$tris.date)

pHSlope <- pHcalib %>%
  group_by(tris.date) %>%
  nest() %>%
  mutate(fitpH = map(data, ~ lm(mVTris ~ TTris, data = .x))) %>%
  mutate(tidy_fit = map(fitpH, broom::tidy)) %>%
  unnest(tidy_fit) %>%
  select(tris.date, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  left_join(daily.probe1, ., by = "tris.date") %>%
  mutate(mVTris = Temperature_C * TTris + `(Intercept)`)

pHSlope <- pHSlope %>%
  mutate(pH.total = seacarb::pH(Ex = pH_mv, Etris = mVTris, S=Salinity_psu, T=Temperature_C))
```

Convert date to ymd for plotting 

```{r}
pHSlope$Date <- ymd(pHSlope$Date) # convert 8 digit date into datetime format

pHSlope <- pHSlope %>% relocate("pH.total", .after = Salinity_psu) %>%
  relocate(pH_mv, .after = Probe.Set)  %>%
  relocate(Salinity_ATC_Mode, .after = Probe.Set)
```

## Change to long format

Change data format to long format 
```{r}
pHSlope.long <-pHSlope %>% pivot_longer(cols=Temperature_C:TA_dKH,
  names_to = "metric",
  values_to = "value")
```

## Plot

Make a list of dataframes, each containing a horizontal line that will correspond to the upper and lower threshold of each parameter (temperature, salinity, pH total)
```{r}
hlines_data <- list(
  data.frame(yintercept = 25.0, metric = "Temperature_C"), # lower threshold for temperature in C°
  data.frame(yintercept = 31, metric = "Temperature_C"), # upper threshold for temperature in C°
  data.frame(yintercept = 34.5, metric = "Salinity_psu"), # lower threshold for salinity in psu
  data.frame(yintercept = 35.5, metric = "Salinity_psu"), # upper threshold for salinity in psu
  data.frame(yintercept = 8.05, metric = "pH.total"), # lower threshold for total pH
  data.frame(yintercept = 8.2, metric = "pH.total") # upper threshold for total pH
    )
```

```{r}
facet_labels <- c(unique(pHSlope.long$metric), unique(pHSlope.long$Treatment))
names(facet_labels) = facet_labels
facet_labels <- replace(facet_labels, which(facet_labels == "TA_dKH"), "TA (dKH)")
facet_labels <- replace(facet_labels, which(facet_labels == "Calcium_ppm"), "Calcium (ppm)")
facet_labels <- replace(facet_labels, which(facet_labels == "pH.total"), "pH")
facet_labels <- replace(facet_labels, which(facet_labels == "Salinity_psu"), "Salinity (psu)")
facet_labels <- replace(facet_labels, which(facet_labels == "Conductivity_mScm"), "Conductivity (mS/cm)")
facet_labels <- replace(facet_labels, which(facet_labels == "Temperature_C"), "Temperature (ºC)")

daily_tank<-pHSlope.long %>% filter(Treatment !=  "Ramp") %>%
  ggplot(aes(x=DateTime, y=value, colour=Tank_ID))+
  geom_point(size=2)+
  xlab("Date")+
  facet_grid(factor(metric,c("pH.total","Salinity_psu","Conductivity_mScm","Temperature_C")) ~ Treatment, scales = "free", labeller = as_labeller(facet_labels))+
  geom_hline(data = hlines_data[[1]], aes(yintercept = yintercept), linetype = "dashed") +    
  geom_hline(data = hlines_data[[2]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[3]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[4]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[5]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[6]], aes(yintercept = yintercept), linetype = "dashed") +
  theme_bw() +
  theme(text = element_text(size = 14)); daily_tank


# Save plot 
ggsave("../output/Daily_Measurements.pdf", daily_tank, width = 10, height = 10, units = c("in"))
```

```{r}
daily_tank<-pHSlope.long %>% filter(Treatment !=  "AcclimationRecovery" & Treatment !=  "Ramp") %>%
  ggplot(aes(x=DateTime, y=value, colour=Tank_ID))+
  geom_point(size=2)+
  xlab("Date")+
  facet_grid(factor(metric,c("pH.total","Salinity_psu","Conductivity_mScm","Temperature_C")) ~ Treatment, scales = "free", labeller = as_labeller(facet_labels))+
  geom_hline(data = hlines_data[[1]], aes(yintercept = yintercept), linetype = "dashed") +    
  geom_hline(data = hlines_data[[2]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[3]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[4]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[5]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[6]], aes(yintercept = yintercept), linetype = "dashed") +
  theme_bw() +
  theme(text = element_text(size = 14)); daily_tank

# Save plot 
ggsave("../output/Daily_Measurements_Exp.pdf", daily_tank, width = 10, height = 10, units = c("in"))
```



Summarize daily measurements during the heat stress experiment
```{r}
daily_exp <- pHSlope %>% 
  filter(Treatment != "AcclimationRecovery" & Treatment !=  "Ramp")

summary <- daily_exp%>%
  group_by(Tank_ID)%>%
  select(Temperature_C:TA_dKH) %>%
  summarise(across(everything(), list(mean = mean, sd = sd), na.rm = TRUE)); summary
```

## HOBO Temps, based on [Jill's Script](https://github.com/JillAshey/Astrangia_repo/blob/0041652d5b2a01145c1c049f10dbc53a8513cb86/scripts/Hobo_Temps.Rmd#L27)

```{r setup, include=FALSE}
#Install and Load necessary libraries
if("lubridate" %in% rownames(installed.packages()) == 'FALSE') install.packages('lubridate') 
if("tidyverse" %in% rownames(installed.packages()) == 'FALSE') install.packages('tidyverse') 
if("emmeans" %in% rownames(installed.packages()) == 'FALSE') install.packages('emmeans') 
if("gridExtra" %in% rownames(installed.packages()) == 'FALSE') install.packages('gridExtra') 

library(lubridate)
library(tidyverse)
library(emmeans)
library(gridExtra)
```

```{r }
##### CONTINUOUS TEMPERATURE DATA FROM HOBO LOGGERS #####

##### QC Hobo Files #####

# Read in temp data 
Tank1 <- read.csv("../data/LoggerData/Tank1_9893752.csv", sep=",", skip=c(1), header=TRUE, na.strings = "NA")[ ,2:4]
Tank2 <- read.csv("../data/LoggerData/Tank2_10655123.csv", sep=",", skip=c(1), header=TRUE, na.strings = "NA")[ ,2:4]
Tank3 <- read.csv("../data/LoggerData/Tank3_10655130.csv", sep=",", skip=c(1), header=TRUE, na.strings = "NA")[ ,2:4]
Tank4 <- read.csv("../data/LoggerData/Tank4_10655129.csv", sep=",", skip=c(1), header=TRUE, na.strings = "NA")[ ,2:4]
Tank5 <- read.csv("../data/LoggerData/Tank5_10655120.csv", sep=",", skip=c(1), header=TRUE, na.strings = "NA")[ ,2:4]
Tank6 <- read.csv("../data/LoggerData/Tank6_10655122.csv", sep=",", skip=c(1), header=TRUE, na.strings = "NA")[ ,2:4]

col_names <- c("DateTimeGMT10","TempC","IntensityLux")

# combine all dataframes into list
Tanks <- list(Tank1 = Tank1,
              Tank2 = Tank2,
              Tank3 = Tank3,
              Tank4 = Tank4,
              Tank5 = Tank5,
              Tank6 = Tank6)

# find dataframe with the fewest number of rows
min_rows <- min(sapply(Tanks, nrow))

# trim them all to be this length
Tanks <- lapply(Tanks, function(df) {df[1:min_rows, ]
                                    colnames(df) <- col_names
                                    return(df)
})

Tank1 <- Tanks$Tank1
Tank2 <- Tanks$Tank2
Tank3 <- Tanks$Tank3
Tank4 <- Tanks$Tank4
Tank5 <- Tanks$Tank5
Tank6 <- Tanks$Tank6

Tank1$Tank <- "Tank1"
Tank2$Tank <- "Tank2"
Tank3$Tank <- "Tank3"
Tank4$Tank <- "Tank4"
Tank5$Tank <- "Tank5"
Tank6$Tank <- "Tank6"

Tank1$Treatment <- "Heated"
Tank2$Treatment <- "Heated"
Tank3$Treatment <- "Heated"
Tank4$Treatment <- "Control"
Tank5$Treatment <- "Control"
Tank6$Treatment <- "Control"

tank_df <- rbind(Tank1, Tank2, Tank3, Tank4, Tank5, Tank6)

tank_df$DateTimeGMT10 <- parse_date_time(tank_df$DateTimeGMT10, "%m/%d/%y %I:%M:%S %p", tz = "Etc/GMT-10")

#this time feels incorrect

write.csv(tank_df,file="../output/All_Tank_HoboTempLight_data.csv")

Temps <- tank_df %>% ggplot(aes(x=DateTimeGMT10, y=TempC)) +
  geom_line(aes(color = Treatment), size = 1)+
  #ylim(5,15)+
  ylab("Temperature °C")
Temps
ggsave("../output/All_Tank_HoboTemp.pdf", plot = last_plot(), width = 8, height = 4)

#remove commas from light data and make numeric
tank_df$IntensityLux <- as.numeric(gsub(",", "", tank_df$IntensityLux))

Light <- tank_df %>% ggplot(aes(x=DateTimeGMT10, y=IntensityLux)) +
  geom_line(aes(color = Treatment), size = 1)+
  #ylim(5,15)+
  ylab("Light lux")

Light
ggsave("../output/All_Tank_HoboLight.pdf", plot = last_plot(), width = 8, height = 4)
```