---
title: "ColorScore"
author: "Zoe Dellaert"
date: "2026-02-18"
output:
  github_document: default
---

## NOTES:

- POC Outlier Images:
  - IMG_4656 --> IMG_4657 
  - IMG_4573 --> reanalyze
  - IMG_4598 --> reanalyze
  - IMG_4755 --> check with Pooja what area she selected -- in picture much of the tissue has sloughed off
    - POC_PAM_1059 was identified as a PAM outlier at 120hrs, which is also when it was photographed for color score
  - IMG_4749 --> reanalyze

## Color Score Analysis of Time Series Image Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Basing this on [Emma Strand's script](https://github.com/emmastrand/EmmaStrand_Notebook/blob/master/Dani_colorscore/Colorscore.md)

Look at the Hackerott et al repo [here](https://github.com/eelabfiu/CoralColorScore) for further analysis ideas

### Load libraries 

```{r}
library(plyr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(vegan)
library(Rmisc)
library(ggpubr)
library(factoextra)
library(MASS)
```

```{r}
save_ggplot <- function(plot, filename, width = 10, height = 7, units = "in", dpi = 300,bg=NULL) {
  png_path <- file.path(outdir, paste0(filename, ".png"))
  pdf_dir <- file.path(outdir, "pdf_figs")
  pdf_path <- file.path(pdf_dir, paste0(filename, ".pdf"))
  
  # Ensure the pdf_figs directory exists
  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)
  
  # Save plots
  ggsave(filename = png_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
  ggsave(filename = pdf_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
}

treat_colors <- c("Control" = "lightblue4", "Heat" = "#D55E00")
time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- c("0", "1", "3", "12", "24", "72", "120")
outdir <- "../output_experimental/ColorScore"
```

## *Pocillopora acuta* 

### Read in data

```{r}
raw <- read.csv("../../3-Pacu/data/Images/ColorScore.csv",
                colClasses=c("Date"="character",
                             "Timepoint"="character",
                             "Plug"="character",
                             "Date"="character"))  %>%
  dplyr::rename(Sample=FileName)

raw <- raw %>% mutate(Treatment=str_replace(Treatment,"heat","Heat"),
                      Treatment=str_replace(Treatment,"control","Control"))
```

### Check raw values

Identify any outliers in the standard values -- any red flags in the pictures/analysis?

These don't get removed here, we remove outliers after they have been normalized to their picture standard values. But outliers here might mean taking a second look at the picture.

```{r}
for (column in c("Red.Standard", "Green.Standard","Blue.Standard")){
  # IQR method 
  Q1 <- quantile(raw[[column]], 0.25)
  Q3 <- quantile(raw[[column]], 0.75)
  IQR_val <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  outliers <- raw[raw[[column]] < lower_bound | raw[[column]] > upper_bound, ]
  print(paste0(column,": IQR = ", Q1,"-",Q3,"; ",nrow(outliers)," outliers"))
  print(outliers)
}
```

### Normalize data

```{r}
proc <- raw %>% mutate(
    Red.Norm.Coral = Red.Coral/Red.Standard,
    Green.Norm.Coral = Green.Coral/Green.Standard,
    Blue.Norm.Coral = Blue.Coral/Blue.Standard
)

rownames(proc) <- proc$Sample
```

#### Identify outliers via IQR

Identify outliers within treatment. Because we do not expect the distributions of the heat and control colors to be the same, we look for outliers separately in these treatments.

```{r}
norm_outliers <- c()

for (column in c("Red.Norm.Coral","Green.Norm.Coral", "Blue.Norm.Coral")){
    stats_temp <- proc %>%
        group_by(across(all_of(c("Treatment")))) %>%
        mutate(Q1 = quantile(get(column), 0.25),
               Q3 = quantile(get(column), 0.75),
               IQR_val = Q3 - Q1,
               lower_bound = Q1 - 1.5 * IQR_val,
               upper_bound = Q3 + 1.5 * IQR_val,
               is_outlier = get(column) < lower_bound | get(column) > upper_bound) %>%
        ungroup()
    
    outliers_heat <- stats_temp %>% filter(Treatment =="Heat" & is_outlier==TRUE)
    outliers_control <- stats_temp %>% filter(Treatment =="Control" & is_outlier==TRUE)
    
    print(paste0(column,", Heat Samples: IQR = ", round(outliers_heat$Q1,2),"-",round(outliers_heat$Q3,2),"; ",nrow(outliers_heat)," outlier(s)"))
    print(outliers_heat %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,column,lower_bound,upper_bound)))
    
    print(paste0(column,", Control Samples: IQR = ", round(outliers_control$Q1,2),"-",round(outliers_control$Q3,2),"; ",nrow(outliers_control)," outlier(s)"))
    print(outliers_control %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,column,lower_bound,upper_bound)))

    norm_outliers <- c(norm_outliers,outliers_heat$Sample,outliers_control$Sample)
}

unique(norm_outliers)
```

#### Remove outliers

Go back and assess the images. Why are these flagging as outliers? 

```{r}
proc_out_rm <- proc %>% filter(!(Sample %in% unique(norm_outliers)))
proc <- proc_out_rm
```

### Calculate color Score

```{r}
#create matrix with rownames as the samples
proc_matrix <- as.matrix(cbind(proc$Red.Norm.Coral,
                                   proc$Green.Norm.Coral,
                                   proc$Blue.Norm.Coral))

rownames(proc_matrix) <- proc$Sample

#calculate distance
dist <- vegdist(proc_matrix, method="euclidean") 

#calculate PCA
PCA <- princomp(dist)
```

```{r}
#extract PC1 as color score
colorscore <- as.data.frame(-PCA$scores[,1]) #extract PC1
colorscore$Sample <- rownames(proc_matrix)

colorscore <- colorscore %>% dplyr::rename(., ColorScore = `-PCA$scores[, 1]`)

final <- left_join(proc, colorscore, by = "Sample") 

# make time and treatment factors
final$Timepoint <- factor(final$Timepoint, levels = as.character(sort(unique(as.numeric(final$Timepoint)))))
final$Treatment <- factor(final$Treatment)

range(final$ColorScore)
```

#### Identify outliers via IQR

At this point, there should not be outliers in the color score values if the normalized values had ouliers removed. But check if there are and examine the data.

```{r}
final <- final %>%
    group_by(across(all_of(c("Treatment")))) %>%
    mutate(Q1 = quantile(ColorScore, 0.25),
           Q3 = quantile(ColorScore, 0.75),
           IQR_val = Q3 - Q1,
           lower_bound = Q1 - 1.5 * IQR_val,
           upper_bound = Q3 + 1.5 * IQR_val,
           is_outlier = ColorScore < lower_bound | ColorScore > upper_bound) %>%
    ungroup()

outliers_heat <- final %>% filter(Treatment =="Heat" & is_outlier==TRUE)
outliers_control <- final %>% filter(Treatment =="Control" & is_outlier==TRUE)

print(paste0("ColorScore, Heat Samples: IQR = ", round(outliers_heat$Q1,2),"-",round(outliers_heat$Q3,2),"; ",nrow(outliers_heat)," outlier(s)"))
print(outliers_heat %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,ColorScore,lower_bound,upper_bound)))

print(paste0("ColorScore, Control Samples: IQR = ", round(outliers_control$Q1,2),"-",round(outliers_control$Q3,2),"; ",nrow(outliers_control)," outlier(s)"))
print(outliers_control %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,ColorScore,lower_bound,upper_bound)))

color_outliers <- c(outliers_heat$Sample,outliers_control$Sample)
```

#### Remove outliers

See note above. Ideally there are no outliers at this stage.

```{r}
final_filtered_IQR <- final %>% filter(!(Sample %in% unique(color_outliers)))
final <- final_filtered_IQR
```

### Plot

```{r}
final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Treatment), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, aes(group = Treatment)) +
  geom_line(stat = "summary", fun = mean, size = 1.2, aes(group = Treatment)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pacu")

final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment, shape=Tank_ID)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Tank_ID), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1, size = 0.4, aes(group = Tank_ID)) +
  geom_line(stat = "summary", fun = mean, size = 0.4, aes(group = Tank_ID)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pacu_by_tank")

final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment, shape=Tank_ID)) +
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data = mean_se, geom = "errorbar", position = position_dodge(width = 0.5), width = 0.2, alpha = 0.5) +
  stat_summary(fun.data = mean_se, geom = "point", position = position_dodge(width = 0.5), alpha = 0.5,size=2.5) +
  stat_summary(fun = mean, geom = "point", aes(group = Treatment), size = 2.5) +
  stat_summary(fun = mean, geom = "line", aes(group = Treatment), size = 1.2) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pacu_tank_means")
```

### Color Score Hackerott

#### Run PCA
```{r}
Color.PCA<-prcomp(proc[,c("Red.Norm.Coral","Green.Norm.Coral","Blue.Norm.Coral")], scale.=TRUE) 

#Initial plot
fviz_pca_ind(Color.PCA)

#Check Variance Explained by Components
summary(Color.PCA)

#Visualize the importance of each principal component
fviz_eig(Color.PCA, addlabels = TRUE) 
```

PC1 Explains 81% of the variance in the color data.

```{r}
##Extract % Variance PC 1
PCA.PC1<-sprintf("%1.2f",get_eigenvalue(Color.PCA)$variance.percent[1])
PCA.PC1

##Extract % Variance PC 2
PCA.PC2<-sprintf("%1.2f",get_eigenvalue(Color.PCA)$variance.percent[2])
PCA.PC2

##Extract Individual Sample Scores
Color.PCA.scores <- as.data.frame(Color.PCA$x)
Color.PCA.scores$Sample<-rownames(Color.PCA.scores)

#Prepare for Plotting
Color.PCA.scores<-left_join(proc, Color.PCA.scores, by = "Sample") 

Color.PCA.scores$Timepoint <- factor(Color.PCA.scores$Timepoint, levels = as.character(sort(unique(as.numeric(Color.PCA.scores$Timepoint)))))
```

#### Plot PCA 
```{r}
#Plot PCA
ggplot(data = Color.PCA.scores, aes(x = PC1, y = PC2)) + 
  geom_point(aes(colour = Treatment),  alpha = 0.8) + 
  scale_colour_manual(values =treat_colors)+
  theme_classic()+
  scale_x_continuous(limits = c(-3, 5.5))+
  scale_y_continuous(limits = c(-4.25, 4.25))+
  labs(x=paste0('PC 1 (',PCA.PC1,"%)"), y=paste0('PC 2 (',PCA.PC2,"%)"))+
  ggtitle("Color Score PCA")

#Plot PCA
ggplot(data = Color.PCA.scores, aes(x = PC1, y = PC2)) + 
  geom_point(aes(colour = Timepoint),  alpha = 0.8) + 
  scale_colour_manual(values =time_colors)+
  theme_classic()+
  scale_x_continuous(limits = c(-3, 5.5))+
  scale_y_continuous(limits = c(-4.25, 4.25))+
  labs(x=paste0('PC 1 (',PCA.PC1,"%)"), y=paste0('PC 2 (',PCA.PC2,"%)"))+
  ggtitle("Color Score PCA")
```

#### LDA on PCA Scores
```{r}
##LDA on PCA Scores
Color.LDA_PCA<-lda(Treatment~PC1+PC2+PC3, data=Color.PCA.scores)

##Predict
Color.pLDA_PCA<-predict(object=Color.LDA_PCA, newdata=Color.PCA.scores)

##Save LD1 Scores
Color.LDA_PCA.scores<-data.frame(Sample=Color.PCA.scores$Sample, LD1=Color.pLDA_PCA$x)

#Prepare for Plotting
ColorData<-left_join(proc, Color.LDA_PCA.scores, by = "Sample") 

# make time and treatment factors
ColorData$Timepoint <- factor(ColorData$Timepoint, levels = as.character(sort(unique(as.numeric(ColorData$Timepoint)))))
ColorData$Treatment <- factor(ColorData$Treatment)
```

#### LD1 Density Plot

```{r}
ggplot(data = ColorData, aes(x = LD1)) + 
  geom_density(aes(fill = Treatment), alpha = 0.6) + 
  scale_fill_manual(values =treat_colors)+
  theme_classic()+
  scale_x_continuous(limits = c(-3, 5.5))+
  labs(x="LD1", y= "Density")+
  ggtitle("Full Dataset")

ggplot(data = ColorData, aes(x = LD1)) + 
  geom_density(aes(fill = Timepoint), alpha = 0.6) + 
  scale_fill_manual(values =time_colors)+
  theme_classic()+
  scale_x_continuous(limits = c(-3, 5.5))+
  labs(x="LD1", y= "Density")+
  ggtitle("Full Dataset")
```

#### Extract Color Score
```{r}
##Invert signs for Control > Heated
ColorData$LD1<-ColorData$LD1*(-1)

#Adding 10 to make all score values positive 
#ColorData$LD1<- ColorData$LD1 +10

ColorData %>%
  ggplot(., aes(x=Timepoint, y=LD1, color=Treatment)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Treatment), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, aes(group = Treatment)) +
  geom_line(stat = "summary", fun = mean, size = 1.2, aes(group = Treatment)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pacu_Hackerott")

ColorData_Full <- ColorData
```



#### by timepoint

```{r}
for (timepoint in unique(raw$Timepoint)){
  proc_timepoint <- proc %>% filter(Timepoint == timepoint)
  
  proc_tp <- proc_timepoint #%>% filter(Sample!="POC_PAM_1059")
  Color.PCA<-prcomp(proc_tp[,c("Red.Norm.Coral","Green.Norm.Coral","Blue.Norm.Coral")], scale.=TRUE) 

  ##Extract % Variance PC 1 & 2
  PCA.PC1<-sprintf("%1.2f",get_eigenvalue(Color.PCA)$variance.percent[1])
  PCA.PC2<-sprintf("%1.2f",get_eigenvalue(Color.PCA)$variance.percent[2])

  ##Extract Individual Sample Scores
  Color.PCA.scores <- as.data.frame(Color.PCA$x)
  Color.PCA.scores$Sample<-rownames(Color.PCA.scores)
  
  #Prepare for Plotting
  Color.PCA.scores<-left_join(proc_tp, Color.PCA.scores, by = "Sample") 
  
  timepoint_pca <- ggplot(data = Color.PCA.scores, aes(x = PC1, y = PC2)) + 
    geom_point(aes(colour = Treatment),  alpha = 0.8) + 
    scale_colour_manual(values =treat_colors)+
    theme_classic()+
    scale_x_continuous(limits = c(-3, 5.5))+
    scale_y_continuous(limits = c(-4.25, 4.25))+
    labs(x=paste0('PC 1 (',PCA.PC1,"%)"), y=paste0('PC 2 (',PCA.PC2,"%)"))+
    ggtitle(paste0("Color Score PCA: ",timepoint,"hrs"))
  
  print(timepoint_pca)
  
  ##LDA on PCA Scores
  Color.LDA_PCA<-lda(Treatment~PC1+PC2+PC3, data=Color.PCA.scores)
  
  ##Predict
  Color.pLDA_PCA<-predict(object=Color.LDA_PCA, newdata=Color.PCA.scores)
  
  ##Save LD1 Scores
  Color.LDA_PCA.scores<-data.frame(Sample=Color.PCA.scores$Sample, LD1=Color.pLDA_PCA$x)
  
  #Prepare for Plotting
  ColorData<-left_join(proc_tp, Color.LDA_PCA.scores, by = "Sample") 
  
  # make time and treatment factors
  ColorData$Timepoint <- factor(ColorData$Timepoint, levels = as.character(sort(unique(as.numeric(ColorData$Timepoint)))))
  ColorData$Treatment <- factor(ColorData$Treatment)
  
  timepoint_density <- ggplot(data = ColorData, aes(x = LD1)) + 
  geom_density(aes(fill = Treatment), alpha = 0.6) + 
  scale_fill_manual(values =treat_colors)+
  theme_classic()+
  scale_x_continuous(limits = c(-3, 5.5))+
  labs(x="LD1", y= "Density") +
    ggtitle(paste0("Density plot: ",timepoint,"hrs"))
  
  print(timepoint_density)

  ##Invert signs for Control > Heated
  ColorData$LD1<-ColorData$LD1*(-1)
  
  #Adding 10 to make all score values positive 
  #ColorData$LD1<- ColorData$LD1 +10

  plot <- ColorData %>%
    ggplot(., aes(x=Timepoint, y=LD1, color=Treatment)) +
    geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
    theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
    scale_color_manual(values = treat_colors) +
    geom_point(stat = "summary", fun = mean, aes(group = Treatment), size=3) +
    stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, aes(group = Treatment)) +
    geom_line(stat = "summary", fun = mean, size = 1.2, aes(group = Treatment)) +
    stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)
  
  print(plot)
  
  assign(paste0("ColorData_TP_",timepoint),ColorData)
}

```
#### combine full and by timepoint data
```{r}
paste0("ColorData_TP_",unique(raw$Timepoint))

##Combine results from individual timepoints
ColorData.TP<-dplyr::bind_rows(mget(paste0("ColorData_TP_", unique(raw$Timepoint))))

##Retain PC1 as Color Score
ColorData.TP$Score_TP<-ColorData.TP$LD1

##Initial Visual Check
ggplot(ColorData.TP, aes(x=Timepoint, y=Score_TP, color=Treatment)) + 
  geom_boxplot(alpha=0.5, shape=2, outlier.shape = NA)+
  geom_jitter(shape=16, position=position_jitter(0.1))+
  theme(axis.text.x = element_text(angle = 90))

##Plot by Treatment
ggplot(ColorData.TP, aes(x=Treatment, y=Score_TP)) + 
  geom_boxplot(alpha=0.5, shape=2, outlier.shape = NA)+
  geom_jitter(shape=16, position=position_jitter(0.1))+
  theme(axis.text.x = element_text(angle = 90))

##Merge with Color Data
names(ColorData.TP)
ColorData<-merge(ColorData_Full, ColorData.TP[,c("Sample", "Score_TP")])
ColorData <- ColorData %>% dplyr::rename(Score_Full=LD1)
```

#### Correlation
```{r}
cor.test(ColorData$Score_Full, ColorData$Score_TP, method="spearman")
plot(ColorData$Score_Full, ColorData$Score_TP)
```

## *Porites compressa* 
### Read in datafiles 

```{r}
raw <- read.csv("../../1-Pcom/data/Images/ColorScore.csv", colClasses=c("Date"="character", 
                                                                               "Timepoint"="character",
                                                                               "Plug"="character",
                                                                               "Date"="character"))  %>%
  dplyr::rename(Sample=FileName)
```

### Check raw values

Identify any outliers in the standard values -- any red flags in the pictures/analysis?

These don't get removed here, we remove outliers after they have been normalized to their picture standard values. But outliers here might mean taking a second look at the picture.

```{r}
for (column in c("Red.Standard", "Green.Standard","Blue.Standard")){
  # IQR method 
  Q1 <- quantile(raw[[column]], 0.25)
  Q3 <- quantile(raw[[column]], 0.75)
  IQR_val <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  outliers <- raw[raw[[column]] < lower_bound | raw[[column]] > upper_bound, ]
  print(paste0(column,": IQR = ", Q1,"-",Q3,"; ",nrow(outliers)," outliers"))
  print(outliers)
}
```

### Normalize data

```{r}
proc <- raw %>% mutate(
    Red.Norm.Coral = Red.Coral/Red.Standard,
    Green.Norm.Coral = Green.Coral/Green.Standard,
    Blue.Norm.Coral = Blue.Coral/Blue.Standard
)

rownames(proc) <- proc$Sample
```

#### Identify outliers via IQR

Identify outliers within treatment. Because we do not expect the distributions of the heat and control colors to be the same, we look for outliers separately in these treatments.

```{r}
norm_outliers <- c()

for (column in c("Red.Norm.Coral","Green.Norm.Coral", "Blue.Norm.Coral")){
    stats_temp <- proc %>%
        group_by(across(all_of(c("Treatment")))) %>%
        mutate(Q1 = quantile(get(column), 0.25),
               Q3 = quantile(get(column), 0.75),
               IQR_val = Q3 - Q1,
               lower_bound = Q1 - 1.5 * IQR_val,
               upper_bound = Q3 + 1.5 * IQR_val,
               is_outlier = get(column) < lower_bound | get(column) > upper_bound) %>%
        ungroup()
    
    outliers_heat <- stats_temp %>% filter(Treatment =="Heat" & is_outlier==TRUE)
    outliers_control <- stats_temp %>% filter(Treatment =="Control" & is_outlier==TRUE)
    
    print(paste0(column,", Heat Samples: IQR = ", round(outliers_heat$Q1,2),"-",round(outliers_heat$Q3,2),"; ",nrow(outliers_heat)," outlier(s)"))
    print(outliers_heat %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,column,lower_bound,upper_bound)))
    
    print(paste0(column,", Control Samples: IQR = ", round(outliers_control$Q1,2),"-",round(outliers_control$Q3,2),"; ",nrow(outliers_control)," outlier(s)"))
    print(outliers_control %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,column,lower_bound,upper_bound)))

    norm_outliers <- c(norm_outliers,outliers_heat$Sample,outliers_control$Sample)
}

unique(norm_outliers)
```

#### Remove outliers

Go back and assess the images. Why are these flagging as outliers? 

```{r}
proc_out_rm <- proc %>% filter(!(Sample %in% unique(norm_outliers)))
proc <- proc_out_rm
```

### Calculate color Score

```{r}
#create matrix with rownames as the samples
proc_matrix <- as.matrix(cbind(proc$Red.Norm.Coral,
                                   proc$Green.Norm.Coral,
                                   proc$Blue.Norm.Coral))

rownames(proc_matrix) <- proc$Sample

#calculate distance
dist <- vegdist(proc_matrix, method="euclidean") 

#calculate PCA
PCA <- princomp(dist)
```

```{r}
#extract PC1 as color score
colorscore <- as.data.frame(-PCA$scores[,1]) #extract PC1
colorscore$Sample <- rownames(proc_matrix)

colorscore <- colorscore %>% dplyr::rename(., ColorScore = `-PCA$scores[, 1]`)

final <- left_join(proc, colorscore, by = "Sample") 

# make time and treatment factors
final$Timepoint <- factor(final$Timepoint, levels = as.character(sort(unique(as.numeric(final$Timepoint)))))
final$Treatment <- factor(final$Treatment)

range(final$ColorScore)
```

#### Identify outliers via IQR

At this point, there should not be outliers in the color score values if the normalized values had ouliers removed. But check if there are and examine the data.

```{r}
final <- final %>%
    group_by(across(all_of(c("Treatment")))) %>%
    mutate(Q1 = quantile(ColorScore, 0.25),
           Q3 = quantile(ColorScore, 0.75),
           IQR_val = Q3 - Q1,
           lower_bound = Q1 - 1.5 * IQR_val,
           upper_bound = Q3 + 1.5 * IQR_val,
           is_outlier = ColorScore < lower_bound | ColorScore > upper_bound) %>%
    ungroup()

outliers_heat <- final %>% filter(Treatment =="Heat" & is_outlier==TRUE)
outliers_control <- final %>% filter(Treatment =="Control" & is_outlier==TRUE)

print(paste0("ColorScore, Heat Samples: IQR = ", round(outliers_heat$Q1,2),"-",round(outliers_heat$Q3,2),"; ",nrow(outliers_heat)," outlier(s)"))
print(outliers_heat %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,ColorScore,lower_bound,upper_bound)))

print(paste0("ColorScore, Control Samples: IQR = ", round(outliers_control$Q1,2),"-",round(outliers_control$Q3,2),"; ",nrow(outliers_control)," outlier(s)"))
print(outliers_control %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,ColorScore,lower_bound,upper_bound)))

color_outliers <- c(outliers_heat$Sample,outliers_control$Sample)
```

#### Remove outliers

See note above. Ideally there are no outliers at this stage.

```{r}
final_filtered_IQR <- final %>% filter(!(Sample %in% unique(color_outliers)))
final <- final_filtered_IQR
```

### Plot

```{r}
final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Treatment), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, aes(group = Treatment)) +
  geom_line(stat = "summary", fun = mean, size = 1.2, aes(group = Treatment)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pcom")

final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment, shape=Tank_ID)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Tank_ID), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1, size = 0.4, aes(group = Tank_ID)) +
  geom_line(stat = "summary", fun = mean, size = 0.4, aes(group = Tank_ID)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pcom_by_tank")

final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment, shape=Tank_ID)) +
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data = mean_se, geom = "errorbar", position = position_dodge(width = 0.5), width = 0.2, alpha = 0.5) +
  stat_summary(fun.data = mean_se, geom = "point", position = position_dodge(width = 0.5), alpha = 0.5,size=2.5) +
  stat_summary(fun = mean, geom = "point", aes(group = Treatment), size = 2.5) +
  stat_summary(fun = mean, geom = "line", aes(group = Treatment), size = 1.2) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pcom_tank_means")
```


## *Montipora capitata* 
### Read in datafiles 

```{r}
raw <- read.csv("../../2-Mcap/data/Images/ColorScore.csv", colClasses=c("Date"="character", 
                                                                               "Timepoint"="character",
                                                                               "Plug"="character",
                                                                               "Date"="character"))  %>%
  dplyr::rename(Sample=FileName)

raw <- raw %>% mutate(Treatment=str_replace(Treatment,"heat","Heat"),
                      Treatment=str_replace(Treatment,"control","Control"))
```

### Check raw values

Identify any outliers in the standard values -- any red flags in the pictures/analysis?

These don't get removed here, we remove outliers after they have been normalized to their picture standard values. But outliers here might mean taking a second look at the picture.

```{r}
for (column in c("Red.Standard", "Green.Standard","Blue.Standard")){
  # IQR method 
  Q1 <- quantile(raw[[column]], 0.25)
  Q3 <- quantile(raw[[column]], 0.75)
  IQR_val <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  outliers <- raw[raw[[column]] < lower_bound | raw[[column]] > upper_bound, ]
  print(paste0(column,": IQR = ", Q1,"-",Q3,"; ",nrow(outliers)," outliers"))
  print(outliers)
}
```

### Normalize data

```{r}
proc <- raw %>% mutate(
    Red.Norm.Coral = Red.Coral/Red.Standard,
    Green.Norm.Coral = Green.Coral/Green.Standard,
    Blue.Norm.Coral = Blue.Coral/Blue.Standard
)

rownames(proc) <- proc$Sample
```

#### Identify outliers via IQR

Identify outliers within treatment. Because we do not expect the distributions of the heat and control colors to be the same, we look for outliers separately in these treatments.

```{r}
norm_outliers <- c()

for (column in c("Red.Norm.Coral","Green.Norm.Coral", "Blue.Norm.Coral")){
    stats_temp <- proc %>%
        group_by(across(all_of(c("Treatment")))) %>%
        mutate(Q1 = quantile(get(column), 0.25),
               Q3 = quantile(get(column), 0.75),
               IQR_val = Q3 - Q1,
               lower_bound = Q1 - 1.5 * IQR_val,
               upper_bound = Q3 + 1.5 * IQR_val,
               is_outlier = get(column) < lower_bound | get(column) > upper_bound) %>%
        ungroup()
    
    outliers_heat <- stats_temp %>% filter(Treatment =="Heat" & is_outlier==TRUE)
    outliers_control <- stats_temp %>% filter(Treatment =="Control" & is_outlier==TRUE)
    
    print(paste0(column,", Heat Samples: IQR = ", round(outliers_heat$Q1,2),"-",round(outliers_heat$Q3,2),"; ",nrow(outliers_heat)," outlier(s)"))
    print(outliers_heat %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,column,lower_bound,upper_bound)))
    
    print(paste0(column,", Control Samples: IQR = ", round(outliers_control$Q1,2),"-",round(outliers_control$Q3,2),"; ",nrow(outliers_control)," outlier(s)"))
    print(outliers_control %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,column,lower_bound,upper_bound)))

    norm_outliers <- c(norm_outliers,outliers_heat$Sample,outliers_control$Sample)
}

unique(norm_outliers)
```

#### Remove outliers

Go back and assess the images. Why are these flagging as outliers? 

```{r}
proc_out_rm <- proc %>% filter(!(Sample %in% unique(norm_outliers)))
proc <- proc_out_rm
```

### Calculate color Score

```{r}
#create matrix with rownames as the samples
proc_matrix <- as.matrix(cbind(proc$Red.Norm.Coral,
                                   proc$Green.Norm.Coral,
                                   proc$Blue.Norm.Coral))

rownames(proc_matrix) <- proc$Sample

#calculate distance
dist <- vegdist(proc_matrix, method="euclidean") 

#calculate PCA
PCA <- princomp(dist)
```

```{r}
#extract PC1 as color score
colorscore <- as.data.frame(-PCA$scores[,1]) #extract PC1
colorscore$Sample <- rownames(proc_matrix)

colorscore <- colorscore %>% dplyr::rename(., ColorScore = `-PCA$scores[, 1]`)

final <- left_join(proc, colorscore, by = "Sample") 

# make time and treatment factors
final$Timepoint <- factor(final$Timepoint, levels = as.character(sort(unique(as.numeric(final$Timepoint)))))
final$Treatment <- factor(final$Treatment)

range(final$ColorScore)
```

#### Identify outliers via IQR

At this point, there should not be outliers in the color score values if the normalized values had ouliers removed. But check if there are and examine the data.

```{r}
final <- final %>%
    group_by(across(all_of(c("Treatment")))) %>%
    mutate(Q1 = quantile(ColorScore, 0.25),
           Q3 = quantile(ColorScore, 0.75),
           IQR_val = Q3 - Q1,
           lower_bound = Q1 - 1.5 * IQR_val,
           upper_bound = Q3 + 1.5 * IQR_val,
           is_outlier = ColorScore < lower_bound | ColorScore > upper_bound) %>%
    ungroup()

outliers_heat <- final %>% filter(Treatment =="Heat" & is_outlier==TRUE)
outliers_control <- final %>% filter(Treatment =="Control" & is_outlier==TRUE)

print(paste0("ColorScore, Heat Samples: IQR = ", round(outliers_heat$Q1,2),"-",round(outliers_heat$Q3,2),"; ",nrow(outliers_heat)," outlier(s)"))
print(outliers_heat %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,ColorScore,lower_bound,upper_bound)))

print(paste0("ColorScore, Control Samples: IQR = ", round(outliers_control$Q1,2),"-",round(outliers_control$Q3,2),"; ",nrow(outliers_control)," outlier(s)"))
print(outliers_control %>% dplyr::select(c(FileName_Orig,Sample,Timepoint,Treatment,Notes,ColorScore,lower_bound,upper_bound)))

color_outliers <- c(outliers_heat$Sample,outliers_control$Sample)
```

#### Remove outliers

See note above. Ideally there are no outliers at this stage.

```{r}
final_filtered_IQR <- final %>% filter(!(Sample %in% unique(color_outliers)))
final <- final_filtered_IQR
```

### Plot

```{r}
final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Treatment), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, aes(group = Treatment)) +
  geom_line(stat = "summary", fun = mean, size = 1.2, aes(group = Treatment)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Mcap")

final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment, shape=Tank_ID)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Tank_ID), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1, size = 0.4, aes(group = Tank_ID)) +
  geom_line(stat = "summary", fun = mean, size = 0.4, aes(group = Tank_ID)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Mcap_by_tank")

final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment, shape=Tank_ID)) +
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data = mean_se, geom = "errorbar", position = position_dodge(width = 0.5), width = 0.2, alpha = 0.5) +
  stat_summary(fun.data = mean_se, geom = "point", position = position_dodge(width = 0.5), alpha = 0.5,size=2.5) +
  stat_summary(fun = mean, geom = "point", aes(group = Treatment), size = 2.5) +
  stat_summary(fun = mean, geom = "line", aes(group = Treatment), size = 1.2) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Mcap_tank_means")
```
