---
title: "ColorScore"
author: "Zoe Dellaert"
date: "2026-02-18"
output:
  github_document: default
---

## Color Score Analysis of Time Series Image Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

Basing this on [Emma Strand's script](https://github.com/emmastrand/Ploidy_methylation/blob/main/scripts/07-Colorscore.Rmd)

### Load libraries 

```{r}
library(plyr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(vegan)
library(Rmisc)
library(ggpubr)
```

```{r}
save_ggplot <- function(plot, filename, width = 10, height = 7, units = "in", dpi = 300,bg=NULL) {
  png_path <- file.path(outdir, paste0(filename, ".png"))
  pdf_dir <- file.path(outdir, "pdf_figs")
  pdf_path <- file.path(pdf_dir, paste0(filename, ".pdf"))
  
  # Ensure the pdf_figs directory exists
  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)
  
  # Save plots
  ggsave(filename = png_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
  ggsave(filename = pdf_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
}

treat_colors <- c("Control" = "lightblue4", "Heat" = "#D55E00")
outdir <- "../output/ColorScore"
```

## *Pocillopora acuta* 
### Read in datafiles 

```{r}
raw <- read.csv("../../3-Pacu/data/Images/ColorScore.csv", colClasses=c("Date"="character", 
                                                                               "Timepoint"="character",
                                                                               "Plug"="character",
                                                                               "PhotoDate"="character"))  %>%
  dplyr::rename(Sample=FileName)

raw <- raw %>% mutate(Treatment=str_replace(Treatment,"heat","Heat"),
                      Treatment=str_replace(Treatment,"control","Control"))
```

Calculate normalized coral color values based on standard color values for that image.

```{r}
proc <- raw %>% mutate(
    Red.Norm.Coral = Red.Coral/Red.Standard,
    Green.Norm.Coral = Green.Coral/Green.Standard,
    Blue.Norm.Coral = Blue.Coral/Blue.Standard
)
```

```{r}
proc_matrix <- as.matrix(cbind(proc$Red.Norm.Coral,
                                   proc$Green.Norm.Coral,
                                   proc$Blue.Norm.Coral)) #create matrix

rownames(proc_matrix) <- proc$Sample #name columns in dataframe

dist <- vegdist(proc_matrix, method="euclidean") 

PCA <- princomp(dist)
```

```{r}
#extract PC1 as color score
colorscore <- as.data.frame(-PCA$scores[,1]) #extract PC1
colorscore$Sample <- rownames(proc_matrix)

colorscore <- colorscore %>% dplyr::rename(., ColorScore = `-PCA$scores[, 1]`)

final <- left_join(proc, colorscore, by = "Sample") 

# make time and treatment factors
final$Timepoint <- factor(final$Timepoint, levels = as.character(sort(unique(as.numeric(final$Timepoint)))))
final$Treatment <- factor(final$Treatment)
```


### Analyze
```{r}
range(final$ColorScore)
hist(final$ColorScore)
```

## Plot

```{r}
final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Treatment), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, aes(group = Treatment)) +
  geom_line(stat = "summary", fun = mean, size = 1.2, aes(group = Treatment)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pacu")

final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment, shape=Tank_ID)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Tank_ID), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1, size = 0.4, aes(group = Tank_ID)) +
  geom_line(stat = "summary", fun = mean, size = 0.4, aes(group = Tank_ID)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pacu_by_tank")
```

## *Porites compressa* 
### Read in datafiles 

```{r}
raw <- read.csv("../../1-Pcom/data/Images/ColorScore.csv", colClasses=c("Date"="character", 
                                                                               "Timepoint"="character",
                                                                               "Plug"="character",
                                                                               "PhotoDate"="character"))  %>%
  dplyr::rename(Sample=FileName)
```

Calculate normalized coral color values based on standard color values for that image.

```{r}
proc <- raw %>% mutate(
    Red.Norm.Coral = Red.Coral/Red.Standard,
    Green.Norm.Coral = Green.Coral/Green.Standard,
    Blue.Norm.Coral = Blue.Coral/Blue.Standard
)
```

```{r}
proc_matrix <- as.matrix(cbind(proc$Red.Norm.Coral,
                                   proc$Green.Norm.Coral,
                                   proc$Blue.Norm.Coral)) #create matrix

rownames(proc_matrix) <- proc$Sample #name columns in dataframe

dist <- vegdist(proc_matrix, method="euclidean") 

PCA <- princomp(dist)
```

```{r}
#extract PC1 as color score
colorscore <- as.data.frame(-PCA$scores[,1]) #extract PC1
colorscore$Sample <- rownames(proc_matrix)

colorscore <- colorscore %>% dplyr::rename(., ColorScore = `-PCA$scores[, 1]`)

final <- left_join(proc, colorscore, by = "Sample") 

# make time and treatment factors
final$Timepoint <- factor(final$Timepoint, levels = as.character(sort(unique(as.numeric(final$Timepoint)))))
final$Treatment <- factor(final$Treatment)
```


### Analyze
```{r}
range(final$ColorScore)
hist(final$ColorScore)
```

## Plot

```{r}
final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Treatment), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, aes(group = Treatment)) +
  geom_line(stat = "summary", fun = mean, size = 1.2, aes(group = Treatment)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pcom")

final %>%
  ggplot(., aes(x=Timepoint, y=ColorScore, color=Treatment, shape=Tank_ID)) +
  geom_jitter(alpha=0.25, size=1.5, width = 0.22) + 
  theme_minimal() + labs(x = "Timepoint",y = "Color Score") +
  scale_color_manual(values = treat_colors) +
  geom_point(stat = "summary", fun = mean, aes(group = Tank_ID), size=3) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.1, size = 0.4, aes(group = Tank_ID)) +
  geom_line(stat = "summary", fun = mean, size = 0.4, aes(group = Tank_ID)) +
  stat_compare_means(aes(group = Treatment),method = "anova",label = "p.format",size = 3)

save_ggplot(last_plot(), "Pcom_by_tank")
```

