---
title: "Genome Annotation"
author: "Zoe Dellaert"
date: "2025-12-07"
output:
  github_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Genome annotation

### SwissProt of ALL genes

Swissprot annotation method based on [E5 Annotations](https://github.com/urol-e5/deep-dive/blob/main/D-Apul/code/20-Apul-gene-annotation.Rmd) and
[here](https://github.com/urol-e5/deep-dive/blob/main/F-Pmea/code/20-Pmea-gene-annotation.Rmd) and Steven Robert's [notebook post here](https://sr320.github.io/tumbling-oysters/posts/sr320-27-go/)

#### Step 1: Gather protein fasta files

```{bash, eval=FALSE}
cd ../references

# copy all genome fasta files here
cp  /work/pi_hputnam_uri_edu/HI_Genomes/*/*pep.faa .
```

#### Step 2: Download swissprot database

```{bash, eval=FALSE}
mkdir blast_dbs
cd blast_dbs

# download swissprot database
curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/uniprot_sprot.fasta.gz
curl -O https://ftp.uniprot.org/pub/databases/uniprot/current_release/knowledgebase/complete/reldate.txt

gunzip uniprot_sprot.fasta.gz

cat reldate.txt
```

UniProt Knowledgebase Release 2025_04 consists of:

- UniProtKB/Swiss-Prot Release 2025_04 of 08-Oct-2025
- UniProtKB/TrEMBL Release 2025_04 of 08-Oct-2025

#### Step 3: Rename file with release date info and confirm contents

```{bash, eval=FALSE}
mv uniprot_sprot.fasta uniprot_sprot_r2025_10_08.fasta

head uniprot_sprot_r2025_10_08.fasta
echo "Number of Sequences"
grep -c ">" uniprot_sprot_r2025_10_08.fasta
# 573661 sequences
```

#### Step 4: Create BLAST protein database from swissprot fasta file

```{bash, eval=FALSE}
module load blast-plus/2.14.1

makeblastdb \
-in uniprot_sprot_r2025_10_08.fasta \
-dbtype prot \
-out uniprot_sprot_r2025_10_08
```

#### Step 5: Tun BLAST of geneome protein FASTAs against swissprot BLAST database

```{bash, eval=FALSE}
cd ../scripts
nano 00_blastp_SwissProt.sh
```

```{bash eval=FALSE}
#!/bin/bash
#SBATCH -t 18:00:00
#SBATCH --nodes=1
#SBATCH --ntasks=1
#SBATCH --cpus-per-task=48
#SBATCH --mem=500GB
#SBATCH --export=NONE
#SBATCH --error=outs_errs/"%x_error.%j" #write out slurm error reports
#SBATCH --output=outs_errs/"%x_output.%j" #write out any program outpus
#SBATCH --mail-type=BEGIN,END,FAIL #email you when job starts, stops and/or fails

module load blast-plus/2.14.1

cd ../references/
mkdir -p annotation

for fasta in *.pep.faa; do
    base=$(basename "$fasta" .pep.faa)

    blastp \
    -query "$fasta" \
    -db blast_dbs/uniprot_sprot_r2024_10_02 \
    -out "annotation/${base}_SwissProt_out.tab" \
    -evalue 1E-05 \
    -num_threads 48 \
    -max_target_seqs 1 \
    -max_hsps 1 \
    -outfmt 6
done

echo "Blast complete!" $(date)
```

#### Step 6: Rename results files

```{bash, eval=FALSE}
cd ../references/annotation/

ls

#Montipora_capitata_HIv3.genes_SwissProt_out.tab
#Porites_compressa_HIv1.genes_SwissProt_out.tab
#Pocillopora_acuta_HIv2.genes_SwissProt_out.tab

tr '|' '\t' <  Montipora_capitata_HIv3.genes_SwissProt_out.tab >  Montipora_capitata_HIv3_SwissProt_out_sep.tab
tr '|' '\t' <  Pocillopora_acuta_HIv2.genes_SwissProt_out.tab > Pocillopora_acuta_HIv2_SwissProt_out_sep.tab
tr '|' '\t' <  Porites_compressa_HIv1.genes_SwissProt_out.tab >  Porites_compressa_HIv1_SwissProt_out_sep.tab

rm *_SwissProt_out.tab
```

#### Step 7: Download Swissprot GO annotation information

- Previous file was just the sequences + accession numbers, this contains the metadata and other information we need about each protein

```{bash, eval=FALSE}
curl -H "Accept: text/plain; format=tsv" "https://rest.uniprot.org/uniprotkb/stream?fields=accession%2Creviewed%2Cid%2Cprotein_name%2Cgene_names%2Corganism_name%2Clength%2Cgo_p%2Cgo%2Cgo_id%2Cgo_c%2Cgo_f&format=tsv&query=%28reviewed%3Atrue%29" -o SwissProt-Annot-GO_20251207.tsv

wc -l SwissProt-Annot-GO_20251207.tsv
# 573662
```

#### Step 8: Join Swissprot GO annotations to BLAST results

```{r load-packages}
library(dplyr)
library(stringr)
library(readr)
```

```{r annotate, eval=FALSE}
# load in GO annotations
spgo <- read.csv("../references/annotation/SwissProt-Annot-GO_20251207.tsv", sep = '\t', header = TRUE)

genomes <- list.files("../references", pattern = "*.pep.faa", full.names = FALSE, ignore.case = TRUE)
genomes <- str_remove(genomes, ".genes.pep.faa")

for (genome in genomes){
  bltabl <- read.csv(paste0("../references/annotation/",genome,"_SwissProt_out_sep.tab"), sep = '\t', header = FALSE)

  annot_tab <- left_join(bltabl, spgo, by = c("V3" = "Entry")) %>%
    select(
      query = V1,
      blast_hit = V3,
      evalue = V13,
      ProteinNames = Protein.names,
      BiologicalProcess = Gene.Ontology..biological.process.,
      GeneOntologyIDs = Gene.Ontology.IDs,
      CellularComponent = Gene.Ontology..cellular.component.,
      MolecularFunction = Gene.Ontology..molecular.function.,
    )

  write.table(annot_tab, 
            file = paste0("../references/annotation/",genome,"_Swissprot_GO.tsv"), 
            sep = "\t", 
            row.names = FALSE, 
            quote = FALSE)
}

rm(spgo)
rm(bltabl)
rm(annot_tab)
```

```{r}
species <- c("Mcap","Pacuta","Pcomp")

Mcap_SwissP <- read.delim("../references/annotation/Montipora_capitata_HIv3_Swissprot_GO.tsv") %>% dplyr::rename(GOs = GeneOntologyIDs)
Pacuta_SwissP <- read.delim("../references/annotation/Pocillopora_acuta_HIv2_Swissprot_GO.tsv") %>% dplyr::rename(GOs = GeneOntologyIDs)
Pcomp_SwissP <- read.delim("../references/annotation/Porites_compressa_HIv1_Swissprot_GO.tsv") %>% dplyr::rename(GOs = GeneOntologyIDs)
```

## Heat Stress Genes


```{r}
for (i in 1:length(species)){
  
HeatStressGenes <- read_csv(paste0("/project/pi_hputnam_uri_edu/zdellaert/snRNA_analysis/multi-sp-snRNA/reference_genes/genes_of_interest/HeatStressGenes_", species[i] ,".csv")) %>%
  dplyr::select(-1) %>%
  dplyr::rename(query = paste0(species[i],"_gene")) %>%
  dplyr::select(query,everything())

assign(paste0(species[i],"_HeatStressGenes"),HeatStressGenes)

HeatStressGenes_unique <- HeatStressGenes %>%
  group_by(query) %>%
  summarize(gene_id = paste(unique(gene_id), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")
            ) 

assign(paste0(species[i],"_HeatStressGenes_unique"),HeatStressGenes_unique)

  rm(HeatStressGenes)
  rm(HeatStressGenes_unique)
}


```


