---
title: "DE"
author: "Zoe Dellaert"
date: "2025-11-20"
output:
  github_document: default
---

## Differential expression analysis of Time Series Bulk RNA Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Thoughts and Notes

- [This paper](https://academic.oup.com/bib/article/20/1/288/4364840) led me to the package ImpulseDE2, which outperforms pairwise timepoint-treatment comparisons typical of DESeq2 workflows.
  - Daniel Spies, Peter F Renz, Tobias A Beyer, Constance Ciaudo, Comparative analysis of differential gene expression tools for RNA sequencing time course data, Briefings in Bioinformatics, Volume 20, Issue 1, January 2019, Pages 288–298, https://doi.org/10.1093/bib/bbx115

## Load packages

```{r packages}
library("genefilter")
library("DESeq2")
library("ggplot2")
library("pheatmap")
library("ggnewscale")
library("RColorBrewer")
library("tidyverse")

sessionInfo() #provides list of loaded packages and version of R.
```

```{r}
save_ggplot <- function(plot, filename, width = 10, height = 7, units = "in", dpi = 300,bg=NULL) {
  png_path <- file.path(outdir, paste0(filename, ".png"))
  pdf_dir <- file.path(outdir, "pdf_figs")
  pdf_path <- file.path(pdf_dir, paste0(filename, ".pdf"))
  
  # Ensure the pdf_figs directory exists
  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)
  
  # Save plots
  ggsave(filename = png_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
  ggsave(filename = pdf_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
}

treat_colors <- c("C" = "lightblue4", "H" = "#D55E00")

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- c("0", "1", "3", "12", "24", "72", "120")
```

## *Pocillopora acuta*

### Preproccessing

#### 1. Read in raw count data

```{r}
# set standard output directory for figures
outdir <- "../output_RNA/differential_expression/POC_PacutaV2"
species <- "Pacuta"

# load in data
counts_raw <- read.csv("../output_RNA/count_matrices/POC_PacutaV2_gene_count_matrix.csv", row.names = 1)

# make list of samples 
samples <- colnames(counts_raw)

# read in SwissProt annotation
SwissProt <- read.delim("../references/annotation/Pocillopora_acuta_HIv2_Swissprot_GO.tsv")
```

#### 2. Extract metadata from sample names

```{r}
# create metadata dataframe from sample names
meta <- data.frame(
  sample = samples, 
  species = str_split(samples, "_", simplify = TRUE)[,1], #extract first part of sample name to get species
  time = str_replace(str_split(samples, "_", simplify = TRUE)[,2],"R", ""), #extract "R##" part to get timepoint then remove R
  replicate = str_split(samples, "_", simplify = TRUE)[,3], #extract "R##" part to get timepoint then remove R
  treatment = str_replace(str_split(samples, "_", simplify = TRUE)[,3],"\\d", "")
)

# add rownames
rownames(meta) <- meta$sample

# make time and treatment factors
meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
meta$treatment <- factor(meta$treatment)

# save metadata
meta <- meta %>% arrange(time, treatment)
write.csv(meta, paste0(outdir,"/RNA_seq_metadata.csv"))
```

#### 3. Confirm that sample names in metadata and count matrix match and are in the same order

```{r}
counts_raw <- counts_raw[, meta$sample]

stopifnot(all(meta$sample %in% colnames(counts_raw))) #are all of the sample names in the metadata column names in the gene count matrix?
stopifnot(all(meta$sample == colnames(counts_raw))) #are they the same in the same order?
```

#### 4. pOverA filtering to reduce dataset

##### Note to self: maybe replace this with treatment-specific filtering. To get genes expressed only at one timepoint in one treatment

```{r}
ffun<-filterfun(pOverA(0.07,10))  # Keep genes expressed at 10+ counts in at least 7% of samples - expressed in all 3 samples at one timepoint from one treatment
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter

paste0("Number of genes after filtering: ", sum(counts_filt_poa))

paste0("% of genes kept: ", round(100*(sum(counts_filt_poa)/nrow(counts_raw)),digits=2),"%")

write.csv(filtered_counts, file = file.path(outdir, "filtered_counts.csv"))
```

### Visualization of overall patterns, outlier detection

We will use [DESeq2](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) to normalize and visualize data

#### 1. Create DESeq object and run DESeq2

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds)
```

#### 2. Check size factors

```{r}
SF.dds <- estimateSizeFactors(dds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.dds)) #View size factors

all(sizeFactors(SF.dds)) < 4
```

#### 3. VST-Transforming count data for visualization 

```{r}
vsd <- vst(dds, blind=FALSE)

#save the vsd transformation
vsd_mat <- assay(vsd)
write.csv(vsd_mat, file = file.path(outdir, "vsd_expression_matrix.csv"))
```

#### 4. Heatmap of the sample-to-sample distances

```{r}
sampleDists <- dist(t(vsd_mat))

sampleDistMatrix <- as.matrix(sampleDists)
colnames(sampleDistMatrix) <- NULL

pheatmap(sampleDistMatrix,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))
```

#### 5. PCA

```{r}
pcaData <- plotPCA(vsd, intgroup=c("time", "treatment"), returnData=TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))
PCA <- ggplot() +
  geom_point(data = subset(pcaData, treatment == "C"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Blues"), name = "Time (hrs) - Control") +
  
  #start new scale
  ggnewscale::new_scale_color() +
  geom_point(data = subset(pcaData, treatment == "H"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Oranges"), name = "Time (hrs) - Heat") +

  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()

save_ggplot(PCA, "PCA_POC")
```

#### 6. Heatmap of variable genes

```{r}
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing=TRUE), 500)

pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

### Heat stress genes

```{r}
HeatStressGenes <- read_csv(paste0("/project/pi_hputnam_uri_edu/zdellaert/snRNA_analysis/multi-sp-snRNA/reference_genes/genes_of_interest/HeatStressGenes_", species ,".csv")) %>% dplyr::select(-1) %>% dplyr::rename(query = paste0(species,"_gene")) %>% dplyr::select(query,everything()) #%>% filter(ref_first_author =="Majerova")

HeatStressGenes_unique <- HeatStressGenes %>% group_by(query) %>%
  summarize(gene_id = paste(unique(gene_id), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")
            ) 

HeatStressGenes_unique <- HeatStressGenes_unique %>% filter(query %in% rownames(vsd_mat))
 
stress_genes_ids <- unique(HeatStressGenes_unique$query) 
stress_genes_vsd <- vsd_mat[stress_genes_ids, ]

plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_unique)

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("Type1", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 1 Expressed Response genes")

save_ggplot(last_plot(), "All_Type1")

plot_df %>% filter(grepl("Type2", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 2 Expressed Response genes")

save_ggplot(last_plot(), "All_Type2")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Pocillopora_acuta_HIv2___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

#### DESeq LRT Test

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds, test = "LRT", reduced = ~ treatment + time)

res <- results(dds)
sig_genes <- subset(res, padj < 0.05)
lrt_res <- as.data.frame(res)

DE_05 <- lrt_res[rownames(lrt_res %>% filter(padj<0.05)),]

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- levels(meta$time)

top_500_DE_genes <- DE_05 %>% arrange(padj) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(log2FoldChange) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(desc(log2FoldChange)) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```


#### DE Heat stress genes

```{r}
plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_unique) %>% left_join(DE_05 %>% rownames_to_column(var="query")) %>%
  filter(!is.na(padj))

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("Type1", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "DE_Type1")

plot_df %>% filter(grepl("Type2", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 2 DE (LRT) Response genes")

save_ggplot(last_plot(), "DE_Type2")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Pocillopora_acuta_HIv2___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Selected Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type1")

plot_df_norm <- plot_df %>%
  group_by(query, time) %>%
  mutate(delta = expression - mean(expression[treatment == "C"]))

plot_df_norm %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=delta, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Pocillopora_acuta_HIv2___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Norm to control -- Selected Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type1_norm_control")

plot_df_norm <- plot_df %>%
  group_by(query, treatment) %>%
  mutate(baseline = mean(expression[time == "0"])) %>%
  mutate(delta = expression - baseline) %>%
  ungroup()

plot_df_norm %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>%
  ggplot(aes(x=time, y=delta, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Pocillopora_acuta_HIv2___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Norm to T0 -- Selected Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type1_norm_T0")

plot_df %>% filter(grepl("GDH",gene_id)|grepl("GS",gene_id)|grepl("AMT1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Pocillopora_acuta_HIv2___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Selected Type 2 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type2")

plot_df %>% filter(grepl("DNaj",gene_id)|grepl("ALOX",gene_id)|grepl("AMT1",gene_id)|grepl("ST1C2",gene_id)|grepl("GDH",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Pocillopora_acuta_HIv2___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Type 1 & 2 DE (LRT) Response genes also DE in Oral Epidermis by LCM")

save_ggplot(last_plot(), "highlighted_tissue_DE")
```

#### Heat stress genes not ID'd above (working on fixing/completing ever-growing list)

NEED_TO_FIND_GENE_ID
  response_type                     category gene_id                           gene_name species species_gene_ID
1         Type1              Innate immunity   NF-KB                               NF-KB    Ahya            <NA>
2         Type1              Innate immunity    AP-1           Transcription factor AP-1    Ahya            <NA>
3         Type1 Stress transcription factors   Elk-3 ETS domain-containing protein Elk-3    Ahya            <NA>
4         Type1 Stress transcription factors    FosB                                FosB    Ahya            <NA>
5         Type1                          UPR   HSP70                               HSP70    Ahya            <NA>
6         Type1                    Apoptosis    Casp                             Caspase    Spis            <NA>
7         Type1                 ROS response     CAT                            Catalase    <NA>            <NA>
8         Type1                 ROS response     SOD                Superoxide dismutase    <NA>            <NA>
  key_timepoint_upregulation key_timepoint_downregulation        timepoint_notes ref_first_author         reference
1                      2.5hr                         <NA>                   <NA>  Traylor Knowles    10.1086/692717
2                        3hr                         <NA>                   <NA>  Traylor Knowles    10.1086/692717
3                        4hr                         <NA>                   <NA>  Traylor Knowles    10.1086/692717
4                        4hr                         <NA>                   <NA>  Traylor Knowles    10.1086/692717
5                        1hr                         <NA>                   <NA>  Traylor Knowles    10.1086/692717
6                        6hr                         <NA> 1st timepoint after T0            Kvitt 10.1038/srep30359
7                       <NA>                         <NA>                   <NA>             <NA>              <NA>
8                       <NA>                         <NA>                   <NA>             <NA>              <NA>

```{r}
plot_df <- as.data.frame(t(vsd_mat)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>% 
  pivot_longer(cols = all_of(rownames(vsd_mat)), names_to="query", values_to="expression") %>%
  mutate(is_DE = query %in% rownames(DE_05))

list <- list(NFKB = c("Pocillopora_acuta_HIv2___RNAseq.g25363.t1b", "Pocillopora_acuta_HIv2___RNAseq.g25813.t1"),
             AP1 = c("Pocillopora_acuta_HIv2___TS.g21530.t1", "Pocillopora_acuta_HIv2___TS.g13419.t1"),
             caspases = c("Pocillopora_acuta_HIv2___RNAseq.g26337.t1","Pocillopora_acuta_HIv2___TS.g22465.t1","Pocillopora_acuta_HIv2___RNAseq.g26338.t1","Pocillopora_acuta_HIv2___RNAseq.g6753.t1","Pocillopora_acuta_HIv2___RNAseq.g19199.t1","Pocillopora_acuta_HIv2___RNAseq.20932_t","Pocillopora_acuta_HIv2___RNAseq.g1378.t1","Pocillopora_acuta_HIv2___RNAseq.g6752.t2","Pocillopora_acuta_HIv2___TS.g18537.t1","Pocillopora_acuta_HIv2___RNAseq.g26123.t1","Pocillopora_acuta_HIv2___RNAseq.g1342.t1"),
             catalase = c("Pocillopora_acuta_HIv2___RNAseq.g11210.t1","Pocillopora_acuta_HIv2___TS.g29361.t1"),
             superoxide_dismutase = c("Pocillopora_acuta_HIv2___TS.g398.t1", "Pocillopora_acuta_HIv2___RNAseq.g10688.t2",
          "Pocillopora_acuta_HIv2___TS.g26014.t1a", "Pocillopora_acuta_HIv2___RNAseq.g208.t1","Pocillopora_acuta_HIv2___TS.g29776.t1",
          "Pocillopora_acuta_HIv2___TS.g29777.t1","Pocillopora_acuta_HIv2___RNAseq.g4525.t1"),
          ELK3 = c("Pocillopora_acuta_HIv2___RNAseq.g6733.t1","Pocillopora_acuta_HIv2___TS.g23786.t1","Pocillopora_acuta_HIv2___RNAseq.24170_t","Pocillopora_acuta_HIv2___RNAseq.g21406.t1","Pocillopora_acuta_HIv2___RNAseq.g6737.t1","Pocillopora_acuta_HIv2___TS.g23792.t4","Pocillopora_acuta_HIv2___RNAseq.g6730.t1","Pocillopora_acuta_HIv2___RNAseq.g6734.t1"),
          FOSB = c("Pocillopora_acuta_HIv2___RNAseq.g25126.t1"),
          TRPA = c("Pocillopora_acuta_HIv2___RNAseq.g9466.t1","Pocillopora_acuta_HIv2___RNAseq.g11852.t1","Pocillopora_acuta_HIv2___RNAseq.10010_t","Pocillopora_acuta_HIv2___RNAseq.g28464.t1","Pocillopora_acuta_HIv2___RNAseq.g7752.t1b","Pocillopora_acuta_HIv2___RNAseq.g11857.t1","Pocillopora_acuta_HIv2___RNAseq.g10676.t1","Pocillopora_acuta_HIv2___RNAseq.g15121.t1")
             )

for (genelist in names(list)){
  genes <- list[[genelist]]
  
  plot <- plot_df %>% filter(query %in% genes) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(paste0(str_replace(query,"Pocillopora_acuta_HIv2___",""), ": ", genelist)~ ~paste0("DE by LRT: ",is_DE),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
  print(plot)
}
```

#### Mechanosensing and other channels

```{r}
membrane_channels <- read_csv(paste0("../references/annotation/",species,"_membrane_channels.csv"))

membrane_channels <- membrane_channels %>% filter(query %in% rownames(vsd_mat))

plot_df <- as.data.frame(t(vsd_mat)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>% 
  pivot_longer(cols = all_of(rownames(vsd_mat)), names_to="query", values_to="expression") %>%
  mutate(is_DE = query %in% rownames(DE_05)) %>% right_join(membrane_channels)

membrane_list <- c("aquaporin","TRP",  "Mechanosensory" ,"calcium",  "ER_calcium" ,"Golgi_calcium" , "SLC",  "PMCA",  "Sodium_calcium_exchanger","Bhattacharya2016")

for (genelist in membrane_list){
  plot <- plot_df %>% filter(grepl(genelist,gene_set)) %>% filter(is_DE == FALSE) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(short_name~paste0("DE by LRT: ",is_DE),scales="free_y") +
  theme_bw() +
  theme(strip.text = element_text(size = 6)) +
  labs(y="VST expression", x="Timepoint")
  print(plot)
}

for (genelist in membrane_list){
  plot_df_filtered <- plot_df %>% filter(grepl(genelist,gene_set)) %>% filter(is_DE == TRUE)
  if (nrow(plot_df_filtered)>1){
    plot <- plot_df_filtered  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
    stat_summary(fun="mean", geom="line") +
    scale_color_manual(values = treat_colors) +
    stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
    facet_wrap(short_name~paste0("DE by LRT: ",is_DE),scales="free_y") +
    theme_bw() +
    theme(strip.text = element_text(size = 6)) +
    labs(y="VST expression", x="Timepoint")
    print(plot)
  }
}
```

### ImpulseDE2

#### Background info

Based on [this paper](https://academic.oup.com/bib/article/20/1/288/4364840#130283262), this is the best package to use other than comparing each time point against each other individually.

Repo here: https://github.com/YosefLab/ImpulseDE2

Tutorial here: http://bioconductor.statistik.tu-dortmund.de/packages/3.11/bioc/vignettes/ImpulseDE2/inst/doc/ImpulseDE2_Tutorial.html , I followed closely with the section "Case-control differential expression analysis"

Read the ImpulseDE2 paper [here](https://academic.oup.com/nar/article/46/20/e119/5068248)

David S Fischer, Fabian J Theis, Nir Yosef, Impulse model-based differential expression analysis of time course sequencing data, Nucleic Acids Research, Volume 46, Issue 20, 16 November 2018, Page e119, https://doi.org/10.1093/nar/gky675

#### 1. Install and load

```{r}
#library(devtools)
#install_github("YosefLab/ImpulseDE2")

library(ImpulseDE2)
library(ComplexHeatmap)
```

#### 2. Metadata formatting

First, reformat our metadata table to match the column names used in the ImpulseDE2 vignette.

```{r}
meta_impulse <- meta %>%
  dplyr::rename(Sample = sample, Time = time, Batch = replicate) %>% 
  mutate(Time = as.numeric(as.character(Time)),
         Condition = str_replace(treatment, "C", "control"),
         Condition = str_replace(Condition, "H", "case")
         ) %>%
  select(-c(species,treatment))
```

#### 3. Then, run ImpulseDE2

This takes a ton of time and memory, so I run it once then save as an RDS.

```{r, eval=FALSE} 
objectImpulseDE2 <- runImpulseDE2(
  matCountData    = as.matrix(counts_raw), #or use filtered_counts 
  dfAnnotation    = meta_impulse,
  boolCaseCtrl    = TRUE,
  vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
  boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
  scaNProc        = 18 )

saveRDS(objectImpulseDE2, file = paste0(outdir, "/objectImpulseDE2.rds"))
```

#### 4. View and save results

```{r}
objectImpulseDE2 <- readRDS(paste0(outdir, "/objectImpulseDE2.rds"))

impulse_results <- objectImpulseDE2$dfImpulseDE2Results
impulse_results_annot <- impulse_results %>% left_join(SwissProt, by = join_by("Gene"=="query")) %>% filter(!is.na(Gene))
write.table(impulse_results,file.path(outdir, "ImpulseDE2_Results.txt"),row.names=F,quote=F,sep="\t")

# Genes with significant treatment effect on temporal trajectory
impulse_sig_genes <- impulse_results %>% filter(padj < 0.05) 

#preview top DE genes 
impulse_sig_genes %>% arrange(padj) %>% head(5) %>% dplyr::select(!contains("converge"))

cat("\nTotal significant genes:", nrow(impulse_sig_genes), "\n")
cat("\nResponse patterns:\n")
cat("Transient:", sum(impulse_sig_genes$isTransient), "\n")
cat("Monotonous:", sum(impulse_sig_genes$isMonotonous), "\n")
cat("Complex:", sum(!impulse_sig_genes$isTransient & !impulse_sig_genes$isMonotonous), "\n")
```

#### 5. Heatmap of transient and non-transiently heat-affected genes

```{r}
lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "control",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.05)
draw(lsHeatmaps$complexHeatmapRaw) 
draw(lsHeatmaps$complexHeatmapFit) 

lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "combined",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.05)
draw(lsHeatmaps$complexHeatmapRaw) 
draw(lsHeatmaps$complexHeatmapFit) 

lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "case",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.05)
draw(lsHeatmaps$complexHeatmapRaw) 
draw(lsHeatmaps$complexHeatmapFit) 

png(paste0(outdir,"/ImpulseDE/ImpulseDE2_heatmap.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapRaw)
dev.off()

png(paste0(outdir,"/ImpulseDE/ImpulseDE2_heatmap_fit.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapFit)
dev.off()

str(lsHeatmaps$lsvecGeneGroups)
```

#### 6. Plot trajectories of top impulseDE genes and specific genes of interest

```{r}
# Plot top 10 differentially expressed (by q-value) genes
lsgplotsGenes <- plotGenes(
  vecGeneIDs       = NULL,
  scaNTopIDs       = 10,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

lsgplotsGenes
```

```{r}
# Majerova 2021 key qPCR genes
majerova_genes <- HeatStressGenes %>% filter(ref_first_author =="Majerova")
stress_genes_ids <- unique(majerova_genes$query)
plot_stress_genes <- stress_genes_ids[stress_genes_ids %in% rownames(objectImpulseDE2@matCountDataProc)] 

impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_unique, by = join_by(Gene==query))

heatgenes <- plotGenes(
  vecGeneIDs       = plot_stress_genes,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  strFileName = "stress_genes_Majerova.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

heatgenes

# HSP genes
HSPS <- impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_unique, by = join_by(Gene==query)) %>% filter(grepl("HSP",gene_id)) %>% pull(Gene)

HSPs <- plotGenes(
  vecGeneIDs       = HSPS,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  strFileName = "HSPs.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

HSPs
```

#### 7. Heatmap of top 500 impulseDE genes

```{r}
top_500_DE_genes <- impulse_results %>% arrange(padj) %>% head(500) %>% rownames()

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

#### 7.5. Heatmap of membrane channel genes

```{r}
membrane_DE_genes <- impulse_results %>%
  right_join(membrane_channels, join_by("Gene"=="query")) %>% 
  filter(gene_set != "SLC") %>%
  filter(padj < 0.05) %>% 
  column_to_rownames("Gene")

plot_df <- as.data.frame(t(vsd_mat)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>% 
  pivot_longer(cols = all_of(rownames(vsd_mat)), names_to="query", values_to="expression") %>%
  mutate(is_DE = query %in% rownames(DE_05)) %>% right_join(membrane_DE_genes %>% rownames_to_column("query"))

membrane_list <- c("aquaporin","TRP",  "Mechanosensory" ,"calcium",  "ER_calcium" ,"Golgi_calcium" , "SLC",  "PMCA",  "Sodium_calcium_exchanger","Bhattacharya2016")

pdf(paste0(outdir,"/ImpulseDE/membrane_sig_plots.pdf"), width = 14, height = 14)

pheatmap(assay(vsd)[rownames(membrane_DE_genes), ], cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=FALSE,
         labels_row=membrane_DE_genes$short_name,
         annotation_row = membrane_DE_genes %>% dplyr::select(gene_set),
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[rownames(membrane_DE_genes), ], cluster_rows=TRUE, show_rownames=TRUE,
          cluster_cols=TRUE, cutree_cols = 2,
         fontsize = 5,
         labels_row=membrane_DE_genes$short_name,
         annotation_row = membrane_DE_genes %>% dplyr::select(gene_set),
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))


for (genelist in unique(plot_df$gene_set)){
  plot_df_filtered <- plot_df %>% filter(grepl(genelist,gene_set))# %>% filter(is_DE == TRUE)
  if (nrow(plot_df_filtered)>1){
    plot <- plot_df_filtered  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
    stat_summary(fun="mean", geom="line") +
    scale_color_manual(values = treat_colors) +
    stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
    facet_wrap(short_name~query,scales="free_y") +
    theme_bw() +
    theme(strip.text = element_text(size = 6)) +
    labs(y="VST expression", x="Timepoint",title=paste(genelist))
    print(plot)
  }
}

dev.off()
```

#### 8. Cluster ImpulseDE2-significant genes by trajectory

For this we will use the package [Mfuzz](https://bioconductor.org/packages/release/bioc/html/Mfuzz.html), [vignette here](https://bioconductor.org/packages/release/bioc/vignettes/Mfuzz/inst/doc/Mfuzz.pdf).

```{r}
#BiocManager::install("Mfuzz")
library(Mfuzz)
```

```{r}
# analyze which of our ImpulseDE2 significant genes are in our vsd matrix
sum(impulse_sig_genes$Gene %in% rownames(vsd_mat))
length(impulse_sig_genes$Gene)

# which ones are missing? 
missing_genes <- impulse_sig_genes$Gene[!(impulse_sig_genes$Gene %in% rownames(vsd_mat))]

# 8 are missing and it is because they were filtered out during pOverA filtering -- as seen with rowSums below, each has fewer than 3 samples with a count >10

counts_raw[missing_genes,]
impulse_sig_genes[missing_genes,]
rowSums(counts_raw[missing_genes,] > 10)
```

```{r}
impulse_sig_genes_transient <- impulse_sig_genes %>% filter(isTransient==TRUE)
impulse_sig_genes_mat <- vsd_mat[rownames(vsd_mat) %in% impulse_sig_genes_transient$Gene,]

heat <- impulse_sig_genes_mat %>% as.data.frame %>% select(contains("_H"))

# average values together across replicates
heat_avg <- heat %>%
  rowwise() %>%
  mutate(
    R0 = mean(c_across(starts_with("POC_R0"))),
    R1 = mean(c_across(starts_with("POC_R1"))),
    R3 = mean(c_across(starts_with("POC_R3"))),
    R12 = mean(c_across(starts_with("POC_R12"))),
    R24 = mean(c_across(starts_with("POC_R24"))),
    R72 = mean(c_across(starts_with("POC_R72"))),
    R120 = mean(c_across(starts_with("POC_R120")))
  ) %>%
  select(R0, R1, R3, R12, R24, R72, R120)

rownames(heat_avg) <- rownames(heat)

heat_eset <- ExpressionSet(assayData = as.matrix(heat_avg))
heat_eset <- standardise(heat_eset) 
```

For fuzzy c-means clustering, the fuzzifier m and the number of clusters c has to be chosen in advance

```{r}
# Determine fuzzifier
m <- mestimate(heat_eset)

# Choose optimal cluster number
#Dmin(heat_eset, m = m, repeats = 3)
optimal_c <- 6

# Run Mfuzz clustering
mfuzz_clusters <- mfuzz(heat_eset, c = optimal_c, m = m)

mfuzz.plot(heat_eset, cl = mfuzz_clusters, new.window =FALSE, mfrow = c(3,3), time.labels =  c(0,1,3,12,24,72,120))

# Visualize clusters
pdf(paste0(outdir,"/temporal_clusters.pdf"), width = 12, height = 10)
mfuzz.plot2(heat_eset, cl = mfuzz_clusters, mfrow = c(3, 4), 
            time.labels = c("0", "1", "3", "12", "24", "72", "120"),
            xlab = "Time (hours)",x11=FALSE)
dev.off()
```

```{r}
cluster2 <- plotGenes(
  vecGeneIDs       = head(names(mfuzz_clusters$cluster)[mfuzz_clusters$cluster==2],5),
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  strFileName = "cluster2.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

cluster2
```

```{r}
cluster3 <- plotGenes(
  vecGeneIDs       = head(names(mfuzz_clusters$cluster)[mfuzz_clusters$cluster==3],5),
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  strFileName = "cluster3.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

cluster3
```

```{r}
# Extract cluster assignments
cluster_assignments <- data.frame(
  gene = names(mfuzz_clusters$cluster),
  cluster = mfuzz_clusters$cluster,
  membership = apply(mfuzz_clusters$membership, 1, max)
)

# Get cluster centers (average trajectory)
cluster_centers <- mfuzz_clusters$centers

# Identify peak timepoint for each cluster
peak_times <- apply(cluster_centers, 1, which.max)
timepoints <- c(0, 1, 3, 12, 24, 72, 120)
cluster_peaks <- data.frame(
  cluster = 1:optimal_c,
  peak_time = timepoints[peak_times])

print(cluster_peaks)
```

##### View clustered genes of interest

```{r}
HSPS <- impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_unique, by = join_by(Gene==query)) %>% filter(grepl("HSP",gene_id)) %>% pull(Gene)

cluster_assignments %>% filter(gene %in% HSPS)
```

```{r}
heat_clustered <- HeatStressGenes_unique %>% left_join(cluster_assignments, by = join_by(query==gene)) %>%
  arrange(cluster, desc(cluster))

# plot this to show which genes are in which cluster
heat_clustered %>% filter(!is.na(cluster)) %>% ggplot(aes(y=reorder(gene_id, cluster), x=factor(cluster), fill=cluster)) +
  geom_tile() +
  theme_bw() +
  labs(x="Mfuzz Cluster", y="Gene ID", title="Heat stress genes clustered by temporal expression pattern")
```


```{r}
channel_clustered <- membrane_channels  %>% unique() %>% 
  filter(gene_set != "SLC") %>%
  left_join(cluster_assignments, by = join_by(query==gene)) %>%
  arrange(cluster)  %>%
  filter(!is.na(cluster)) 

cat("Total membrane channel genes:", nrow(membrane_channels), "\n")
cat("Found in ImpulseDE/Mfuzz data:", nrow(channel_clustered), "\n")

channel_clustered %>%
    ggplot(aes(x = factor(cluster), y = reorder(short_name, cluster), fill = gene_set)) +
  geom_tile(color = "white") +
  geom_text(aes(label = cluster), size = 2) +
  theme_bw() +
  theme(axis.text.y = element_text(size = 6)) +
  labs(x = "Mfuzz Cluster", 
       y = "Gene ID", 
       fill = "Channel Type",
       title = "Membrane Channel Genes - Cluster Assignment",
       subtitle = paste0(sum(!is.na(channel_clustered$cluster)), 
                        " genes significantly DE"))

heat_eset_membrane <- heat_eset[channel_clustered$query, ]
mfuzz_clusters_membrane <- mfuzz_clusters
mfuzz_clusters_membrane$membership <- mfuzz_clusters$membership[channel_clustered$query, ]
mfuzz_clusters_membrane$cluster <- mfuzz_clusters$cluster[channel_clustered$query]

mfuzz.plot(
  heat_eset_membrane,
  cl = mfuzz_clusters_membrane,
  new.window = FALSE,
  mfrow = c(3,3),
  time.labels = c(0,1,3,12,24,72,120)
)

# Visualize clusters
pdf(paste0(outdir,"/membrane_temporal_clusters.pdf"), width = 12, height = 10)
mfuzz.plot2(heat_eset_membrane, cl = mfuzz_clusters_membrane, mfrow = c(3, 4), 
            time.labels = c("0", "1", "3", "12", "24", "72", "120"),
            xlab = "Time (hours)",x11=FALSE)
dev.off()

plot_df <- as.data.frame(t(vsd_mat)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>% 
  pivot_longer(cols = all_of(rownames(vsd_mat)), names_to="query", values_to="expression") %>%
  mutate(is_DE = query %in% rownames(DE_05)) %>% right_join(membrane_DE_genes %>% rownames_to_column("query"))

pdf(paste0(outdir,"/ImpulseDE/membrane_sig_plots_clustered.pdf"), width = 14, height = 14)

for (cl in c(0,unique(channel_clustered$cluster))){
  plot_df_filtered <- plot_df %>% left_join(channel_clustered) %>% filter(!is.na(cluster))
  
  if (cl ==0){
    plot_df_low_membership <- plot_df_filtered %>% filter(membership<0.5)
    
    plot <- plot_df_low_membership  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
    stat_summary(fun="mean", geom="line") +
    scale_color_manual(values = treat_colors) +
    stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
    facet_wrap(short_name~query,scales="free_y") +
    theme_bw() +
    theme(strip.text = element_text(size = 6)) +
    labs(y="VST expression", x="Timepoint",title=paste("Membership < 0.5"))
    print(plot)
  } else{
    plot_df_cluster <- plot_df_filtered %>% filter(cluster==cl)
      if (nrow(plot_df_cluster)>1){
    plot <- plot_df_cluster  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
    stat_summary(fun="mean", geom="line") +
    scale_color_manual(values = treat_colors) +
    stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
    facet_wrap(short_name~query,scales="free_y") +
    theme_bw() +
    theme(strip.text = element_text(size = 6)) +
    labs(y="VST expression", x="Timepoint",title=paste("Cluster ",cl))
    print(plot)
  }
  }
}
dev.off()
```



## *Montipora capitata*

### Preproccessing

#### 1. Read in raw count data

```{r}
# set standard output directory for figures
outdir <- "../output_RNA/differential_expression/MON_MCapV3"
species <- "Mcap"

# load in data
counts_raw <- read.csv("../output_RNA/count_matrices/MON_MCapV3_gene_count_matrix.csv", row.names = 1)

# make list of samples 
samples <- colnames(counts_raw)

# read in SwissProt annotation
SwissProt <- read.delim("../references/annotation/Montipora_capitata_HIv3_Swissprot_GO.tsv")
```

#### 2. Extract metadata from sample names

```{r}
# create metadata dataframe from sample names
meta <- data.frame(
  sample = samples, 
  species = str_split(samples, "_", simplify = TRUE)[,1], #extract first part of sample name to get species
  time = str_replace(str_split(samples, "_", simplify = TRUE)[,2],"R", ""), #extract "R##" part to get timepoint then remove R
  replicate = str_split(samples, "_", simplify = TRUE)[,3], #extract "R##" part to get timepoint then remove R
  treatment = str_replace(str_split(samples, "_", simplify = TRUE)[,3],"\\d", "")
)

# add rownames
rownames(meta) <- meta$sample

# make time and treatment factors
meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
meta$treatment <- factor(meta$treatment)

# save metadata
meta <- meta %>% arrange(time, treatment)
write.csv(meta, paste0(outdir,"/RNA_seq_metadata.csv"))
```

#### 3. Confirm that sample names in metadata and count matrix match and are in the same order

```{r}
counts_raw <- counts_raw[, meta$sample]
```

Remove outliers:
```{r}
counts_raw <- counts_raw[, !(colnames(counts_raw) %in% c("MON_R72_H1","MON_R72_H2"))]
meta <- meta[!(rownames(meta) %in% c("MON_R72_H1","MON_R72_H2")),]
```

Data sanity checks!
```{r}
stopifnot(all(meta$sample %in% colnames(counts_raw))) #are all of the sample names in the metadata column names in the gene count matrix?
stopifnot(all(meta$sample == colnames(counts_raw))) #are they the same in the same order?
```

#### 4. pOverA filtering to reduce dataset

##### Note to self: maybe replace this with treatment-specific filtering. To get genes expressed only at one timepoint in one treatment

```{r}
ffun<-filterfun(pOverA(0.07,10))  # Keep genes expressed at 10+ counts in at least 7% of samples - expressed in all 3 samples at one timepoint from one treatment
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter

paste0("Number of genes after filtering: ", sum(counts_filt_poa))

paste0("% of genes kept: ", round(100*(sum(counts_filt_poa)/nrow(counts_raw)),digits=2),"%")

write.csv(filtered_counts, file = file.path(outdir, "filtered_counts.csv"))
```

### Visualization of overall patterns, outlier detection

We will use [DESeq2](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) to normalize and visualize data

#### 1. Create DESeq object and run DESeq2

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds)
```

#### 2. Check size factors

```{r}
SF.dds <- estimateSizeFactors(dds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.dds)) #View size factors

all(sizeFactors(SF.dds)) < 4
```

#### 3. VST-Transforming count data for visualization 

```{r}
vsd <- vst(dds, blind=FALSE)

#save the vsd transformation
vsd_mat <- assay(vsd)
write.csv(vsd_mat, file = file.path(outdir, "vsd_expression_matrix.csv"))
```

#### 4. Heatmap of the sample-to-sample distances

```{r}
sampleDists <- dist(t(vsd_mat))

sampleDistMatrix <- as.matrix(sampleDists)
colnames(sampleDistMatrix) <- NULL

pheatmap(sampleDistMatrix,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))
```

#### 5. PCA

```{r}
pcaData <- plotPCA(vsd, intgroup=c("time", "treatment"), returnData=TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))
PCA <- ggplot() +
  geom_point(data = subset(pcaData, treatment == "C"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Blues"), name = "Time (hrs) - Control") +
  
  #start new scale
  ggnewscale::new_scale_color() +
  geom_point(data = subset(pcaData, treatment == "H"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Oranges"), name = "Time (hrs) - Heat") +

  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()

save_ggplot(PCA, "PCA_MON")
```

#### 6. Heatmap of variable genes

```{r}
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing=TRUE), 500)

pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

### Heat stress genes

```{r}
HeatStressGenes <- read_csv(paste0("/project/pi_hputnam_uri_edu/zdellaert/snRNA_analysis/multi-sp-snRNA/reference_genes/genes_of_interest/HeatStressGenes_", species ,".csv")) %>% dplyr::select(-1) %>% dplyr::rename(query = paste0(species,"_gene")) %>% dplyr::select(query,everything()) #%>% filter(ref_first_author =="Majerova")

HeatStressGenes_unique <- HeatStressGenes %>% group_by(query) %>%
  summarize(gene_id = paste(unique(gene_id), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")
            ) 

HeatStressGenes_unique <- HeatStressGenes_unique %>% filter(query %in% rownames(vsd_mat))
 
stress_genes_ids <- unique(HeatStressGenes_unique$query) 
stress_genes_vsd <- vsd_mat[stress_genes_ids, ]

plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_unique)

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("Type1", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 1 Expressed Response genes")

save_ggplot(last_plot(), "All_Type1")

plot_df %>% filter(grepl("Type2", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 2 Expressed Response genes")

save_ggplot(last_plot(), "All_Type2")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Montipora_capitata_HIv3___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

#### DESeq LRT Test

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds, test = "LRT", reduced = ~ treatment + time)

res <- results(dds)
sig_genes <- subset(res, padj < 0.05)
lrt_res <- as.data.frame(res)

DE_05 <- lrt_res[rownames(lrt_res %>% filter(padj<0.05)),]

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- levels(meta$time)

top_500_DE_genes <- DE_05 %>% arrange(padj) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(log2FoldChange) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(desc(log2FoldChange)) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

#### DE Heat stress genes

```{r}
plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_unique) %>% left_join(DE_05 %>% rownames_to_column(var="query")) %>%
  filter(!is.na(padj))

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("Type1", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "DE_Type1")

plot_df %>% filter(grepl("Type2", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 2 DE (LRT) Response genes")

save_ggplot(last_plot(), "DE_Type2")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Montipora_capitata_HIv3___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Selected Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type1")

plot_df_norm <- plot_df %>%
  group_by(query, time) %>%
  mutate(delta = expression - mean(expression[treatment == "C"]))

plot_df_norm %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=delta, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Montipora_capitata_HIv3___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Norm to control -- Selected Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type1_norm_control")

plot_df_norm <- plot_df %>%
  group_by(query, treatment) %>%
  mutate(baseline = mean(expression[time == "0"])) %>%
  mutate(delta = expression - baseline) %>%
  ungroup()

plot_df_norm %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>%
  ggplot(aes(x=time, y=delta, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Montipora_capitata_HIv3___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Norm to T0 -- Selected Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type1_norm_T0")

plot_df %>% filter(grepl("GDH",gene_id)|grepl("GS",gene_id)|grepl("AMT1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Montipora_capitata_HIv3___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Selected Type 2 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type2")
```

#### Mechanosensing and other channels

```{r}
membrane_channels <- read_csv(paste0("../references/annotation/",species,"_membrane_channels.csv"))

membrane_channels <- membrane_channels %>% filter(query %in% rownames(vsd_mat))

plot_df <- as.data.frame(t(vsd_mat)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>% 
  pivot_longer(cols = all_of(rownames(vsd_mat)), names_to="query", values_to="expression") %>%
  mutate(is_DE = query %in% rownames(DE_05)) %>% right_join(membrane_channels)

membrane_list <- c("aquaporin","TRP",  "Mechanosensory" ,"calcium",  "ER_calcium" ,"Golgi_calcium" , "SLC",  "PMCA",  "Sodium_calcium_exchanger","Bhattacharya2016")

for (genelist in membrane_list){
  plot_df_filtered <- plot_df %>% filter(grepl(genelist,gene_set)) %>% filter(is_DE == FALSE)
  if (nrow(plot_df_filtered)>1){
    plot <- plot_df_filtered  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(short_name~paste0("DE by LRT: ",is_DE),scales="free_y") +
  theme_bw() +
  theme(strip.text = element_text(size = 6)) +
  labs(y="VST expression", x="Timepoint")
  print(plot)
  }
}

for (genelist in membrane_list){
  plot_df_filtered <- plot_df %>% filter(grepl(genelist,gene_set)) %>% filter(is_DE == TRUE)
  if (nrow(plot_df_filtered)>1){
    plot <- plot_df_filtered  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
    stat_summary(fun="mean", geom="line") +
    scale_color_manual(values = treat_colors) +
    stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
    facet_wrap(short_name~paste0("DE by LRT: ",is_DE),scales="free_y") +
    theme_bw() +
    theme(strip.text = element_text(size = 6)) +
    labs(y="VST expression", x="Timepoint")
    print(plot)
  }
}
```

### ImpulseDE2

#### Background info

Based on [this paper](https://academic.oup.com/bib/article/20/1/288/4364840#130283262), this is the best package to use other than comparing each time point against each other individually.

Repo here: https://github.com/YosefLab/ImpulseDE2

Tutorial here: http://bioconductor.statistik.tu-dortmund.de/packages/3.11/bioc/vignettes/ImpulseDE2/inst/doc/ImpulseDE2_Tutorial.html , I followed closely with the section "Case-control differential expression analysis"

Read the ImpulseDE2 paper [here](https://academic.oup.com/nar/article/46/20/e119/5068248)

David S Fischer, Fabian J Theis, Nir Yosef, Impulse model-based differential expression analysis of time course sequencing data, Nucleic Acids Research, Volume 46, Issue 20, 16 November 2018, Page e119, https://doi.org/10.1093/nar/gky675

#### 1. Install and load

```{r}
#library(devtools)
#install_github("YosefLab/ImpulseDE2")

library(ImpulseDE2)
library(ComplexHeatmap)
```

#### 2. Metadata formatting

First, reformat our metadata table to match the column names used in the ImpulseDE2 vignette.

```{r}
meta_impulse <- meta %>%
  dplyr::rename(Sample = sample, Time = time, Batch = replicate) %>% 
  mutate(Time = as.numeric(as.character(Time)),
         Condition = str_replace(treatment, "C", "control"),
         Condition = str_replace(Condition, "H", "case")
         ) %>%
  select(-c(species,treatment))
```

#### 3. Then, run ImpulseDE2

This takes a ton of time and memory, so I run it once then save as an RDS.

```{r, eval=FALSE} 
objectImpulseDE2 <- runImpulseDE2(
  matCountData    = as.matrix(counts_raw), #or use filtered_counts 
  dfAnnotation    = meta_impulse,
  boolCaseCtrl    = TRUE,
  vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
  boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
  scaNProc        = 18 )

saveRDS(objectImpulseDE2, file = paste0(outdir, "/objectImpulseDE2.rds"))
```

#### 4. View and save results

```{r}
objectImpulseDE2 <- readRDS(paste0(outdir, "/objectImpulseDE2.rds"))

impulse_results <- objectImpulseDE2$dfImpulseDE2Results
impulse_results_annot <- impulse_results %>% left_join(SwissProt, by = join_by("Gene"=="query")) %>% filter(!is.na(Gene))
write.table(impulse_results,file.path(outdir, "ImpulseDE2_Results.txt"),row.names=F,quote=F,sep="\t")

# Genes with significant treatment effect on temporal trajectory
impulse_sig_genes <- impulse_results %>% filter(padj < 0.05) 

#preview top DE genes 
impulse_sig_genes %>% arrange(padj) %>% head(5) %>% dplyr::select(!contains("converge"))

cat("\nTotal significant genes:", nrow(impulse_sig_genes), "\n")
cat("\nResponse patterns:\n")
cat("Transient:", sum(impulse_sig_genes$isTransient), "\n")
cat("Monotonous:", sum(impulse_sig_genes$isMonotonous), "\n")
cat("Complex:", sum(!impulse_sig_genes$isTransient & !impulse_sig_genes$isMonotonous), "\n")
```

#### 5. Heatmap of transient and non-transiently heat-affected genes

```{r}
lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "control",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.05)
draw(lsHeatmaps$complexHeatmapRaw) 
draw(lsHeatmaps$complexHeatmapFit) 

lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "combined",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.05)
draw(lsHeatmaps$complexHeatmapRaw) 
draw(lsHeatmaps$complexHeatmapFit) 

lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "case",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.05)
draw(lsHeatmaps$complexHeatmapRaw) 
draw(lsHeatmaps$complexHeatmapFit) 

png(paste0(outdir,"/ImpulseDE/ImpulseDE2_heatmap.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapRaw)
dev.off()

png(paste0(outdir,"/ImpulseDE/ImpulseDE2_heatmap_fit.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapFit)
dev.off()

str(lsHeatmaps$lsvecGeneGroups)
```

#### 6. Plot trajectories of top impulseDE genes and specific genes of interest

```{r}
# Plot top 10 differentially expressed (by q-value) genes
lsgplotsGenes <- plotGenes(
  vecGeneIDs       = NULL,
  scaNTopIDs       = 10,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

lsgplotsGenes
```

```{r}
# Majerova 2021 key qPCR genes
majerova_genes <- HeatStressGenes %>% filter(ref_first_author =="Majerova")
stress_genes_ids <- unique(majerova_genes$query)
plot_stress_genes <- stress_genes_ids[stress_genes_ids %in% rownames(objectImpulseDE2@matCountDataProc)] 

impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_unique, by = join_by(Gene==query))

heatgenes <- plotGenes(
  vecGeneIDs       = plot_stress_genes,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  strFileName = "stress_genes_Majerova.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

heatgenes

# HSP genes
HSPS <- impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_unique, by = join_by(Gene==query)) %>% filter(grepl("HSP",gene_id)) %>% pull(Gene)

HSPs <- plotGenes(
  vecGeneIDs       = HSPS,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  strFileName = "HSPs.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

HSPs
```

#### 7. Heatmap of top 500 impulseDE genes

```{r}
top_500_DE_genes <- impulse_results %>% arrange(padj) %>% head(500) %>% rownames()

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

#### 7.5. Heatmap of membrane channel genes

```{r}
membrane_DE_genes <- impulse_results %>%
  right_join(membrane_channels, join_by("Gene"=="query")) %>% 
  filter(gene_set != "SLC") %>%
  filter(padj < 0.05) %>% 
  column_to_rownames("Gene")

plot_df <- as.data.frame(t(vsd_mat)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>% 
  pivot_longer(cols = all_of(rownames(vsd_mat)), names_to="query", values_to="expression") %>%
  mutate(is_DE = query %in% rownames(DE_05)) %>% right_join(membrane_DE_genes %>% rownames_to_column("query"))

membrane_list <- c("aquaporin","TRP",  "Mechanosensory" ,"calcium",  "ER_calcium" ,"Golgi_calcium" , "SLC",  "PMCA",  "Sodium_calcium_exchanger","Bhattacharya2016")

pdf(paste0(outdir,"/ImpulseDE/membrane_sig_plots.pdf"), width = 14, height = 14)

pheatmap(assay(vsd)[rownames(membrane_DE_genes), ], cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=FALSE,
         labels_row=membrane_DE_genes$short_name,
         annotation_row = membrane_DE_genes %>% dplyr::select(gene_set),
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[rownames(membrane_DE_genes), ], cluster_rows=TRUE, show_rownames=TRUE,
          cluster_cols=TRUE, cutree_cols = 2,
         fontsize = 5,
         labels_row=membrane_DE_genes$short_name,
         annotation_row = membrane_DE_genes %>% dplyr::select(gene_set),
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))


for (genelist in unique(plot_df$gene_set)){
  plot_df_filtered <- plot_df %>% filter(grepl(genelist,gene_set))# %>% filter(is_DE == TRUE)
  if (nrow(plot_df_filtered)>1){
    plot <- plot_df_filtered  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
    stat_summary(fun="mean", geom="line") +
    scale_color_manual(values = treat_colors) +
    stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
    facet_wrap(short_name~query,scales="free_y") +
    theme_bw() +
    theme(strip.text = element_text(size = 6)) +
    labs(y="VST expression", x="Timepoint",title=paste(genelist))
    print(plot)
  }
}

dev.off()
```


## *Porites compressa*

### Preproccessing

#### 1. Read in raw count data

```{r}
# set standard output directory for figures
outdir <- "../output_RNA/differential_expression/POR_Pcomp"
species <- "Pcomp"

# load in data
counts_raw <- read.csv("../output_RNA/count_matrices/POR_Pcomp_gene_count_matrix.csv", row.names = 1)

# make list of samples 
samples <- colnames(counts_raw)

# read in SwissProt annotation
SwissProt <- read.delim("../references/annotation/Porites_compressa_HIv1_Swissprot_GO.tsv")
```

#### 2. Extract metadata from sample names

```{r}
# create metadata dataframe from sample names
meta <- data.frame(
  sample = samples, 
  species = str_split(samples, "_", simplify = TRUE)[,1], #extract first part of sample name to get species
  time = str_replace(str_split(samples, "_", simplify = TRUE)[,2],"R", ""), #extract "R##" part to get timepoint then remove R
  replicate = str_split(samples, "_", simplify = TRUE)[,3], #extract "R##" part to get timepoint then remove R
  treatment = str_replace(str_split(samples, "_", simplify = TRUE)[,3],"\\d", "")
)

# add rownames
rownames(meta) <- meta$sample

# make time and treatment factors
meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
meta$treatment <- factor(meta$treatment)

# save metadata
meta <- meta %>% arrange(time, treatment)
write.csv(meta, paste0(outdir,"/RNA_seq_metadata.csv"))
```

#### 3. Confirm that sample names in metadata and count matrix match and are in the same order

```{r}
counts_raw <- counts_raw[, meta$sample]
```

Remove outliers:
```{r}
#counts_raw <- counts_raw[, !(colnames(counts_raw) %in% c("MON_R72_H1","MON_R72_H2"))]
#meta <- meta[!(rownames(meta) %in% c("MON_R72_H1","MON_R72_H2")),]
```

Data sanity checks!
```{r}
stopifnot(all(meta$sample %in% colnames(counts_raw))) #are all of the sample names in the metadata column names in the gene count matrix?
stopifnot(all(meta$sample == colnames(counts_raw))) #are they the same in the same order?
```

#### 4. pOverA filtering to reduce dataset

##### Note to self: maybe replace this with treatment-specific filtering. To get genes expressed only at one timepoint in one treatment

```{r}
ffun<-filterfun(pOverA(0.07,10))  # Keep genes expressed at 10+ counts in at least 7% of samples - expressed in all 3 samples at one timepoint from one treatment
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter

paste0("Number of genes after filtering: ", sum(counts_filt_poa))

paste0("% of genes kept: ", round(100*(sum(counts_filt_poa)/nrow(counts_raw)),digits=2),"%")

write.csv(filtered_counts, file = file.path(outdir, "filtered_counts.csv"))
```

### Visualization of overall patterns, outlier detection

We will use [DESeq2](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html) to normalize and visualize data

#### 1. Create DESeq object and run DESeq2

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds)
```

#### 2. Check size factors

```{r}
SF.dds <- estimateSizeFactors(dds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.dds)) #View size factors

all(sizeFactors(SF.dds)) < 4
```

#### 3. VST-Transforming count data for visualization 

```{r}
vsd <- vst(dds, blind=FALSE)

#save the vsd transformation
vsd_mat <- assay(vsd)
write.csv(vsd_mat, file = file.path(outdir, "vsd_expression_matrix.csv"))
```

#### 4. Heatmap of the sample-to-sample distances

```{r}
sampleDists <- dist(t(vsd_mat))

sampleDistMatrix <- as.matrix(sampleDists)
colnames(sampleDistMatrix) <- NULL

pheatmap(sampleDistMatrix,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))
```

#### 5. PCA

```{r}
pcaData <- plotPCA(vsd, intgroup=c("time", "treatment"), returnData=TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))
PCA <- ggplot() +
  geom_point(data = subset(pcaData, treatment == "C"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Blues"), name = "Time (hrs) - Control") +
  
  #start new scale
  ggnewscale::new_scale_color() +
  geom_point(data = subset(pcaData, treatment == "H"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Oranges"), name = "Time (hrs) - Heat") +

  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()

save_ggplot(PCA, "PCA_POR")
```

The following samples had mapping rates < 20%:

| sample      | uniquely_mapped_percent | rRNA_matched_reads_Million | total_reads_Million |
|-------------|-------------------------|----------------------------|---------------------|
| POR_R72_H1  | 6.4                     | 12.92                      | 15.47               |
| POR_R24_H1  | 7.67                    | 14.23                      | 17.26               |
| POR_R120_C3 | 7.77                    | 18.64                      | 21.02               |
| POR_R72_H2  | 8.77                    | 12.24                      | 15.16               |
| POR_R1_H2   | 10.37                   | 14.78                      | 17.47               |
| POR_R120_C1 | 10.78                   | 11.68                      | 13.99               |
| POR_R1_H1   | 11.72                   | 15.47                      | 18.69               |
| POR_R1_C1   | 12.04                   | 13.14                      | 15.99               |
| POR_R12_C1  | 12.2                    | 12.17                      | 14.88               |
| POR_R12_C3  | 15.78                   | 11.7                       | 15.09               |
| POR_R0_H1   | 16.57                   | 12.57                      | 16.46               |
| POR_R120_H3 | 18.23                   | 12.18                      | 17.16               |
| POR_R3_C3   | 18.27                   | 11                         | 14.95               |
| POR_R0_H2   | 18.35                   | 14.14                      | 19.31               |
| POR_R24_C2  | 19.72                   | 11.55                      | 16                  |

#### 6. Heatmap of variable genes

```{r}
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing=TRUE), 500)

pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

### Heat stress genes

```{r}
HeatStressGenes <- read_csv(paste0("/project/pi_hputnam_uri_edu/zdellaert/snRNA_analysis/multi-sp-snRNA/reference_genes/genes_of_interest/HeatStressGenes_", species ,".csv")) %>% dplyr::select(-1) %>% dplyr::rename(query = paste0(species,"_gene")) %>% dplyr::select(query,everything()) #%>% filter(ref_first_author =="Majerova")

HeatStressGenes_unique <- HeatStressGenes %>% group_by(query) %>%
  summarize(gene_id = paste(unique(gene_id), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")
            ) 

HeatStressGenes_unique <- HeatStressGenes_unique %>% filter(query %in% rownames(vsd_mat))
 
stress_genes_ids <- unique(HeatStressGenes_unique$query) 
stress_genes_vsd <- vsd_mat[stress_genes_ids, ]

plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_unique)

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("Type1", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 1 Expressed Response genes")

save_ggplot(last_plot(), "All_Type1")

plot_df %>% filter(grepl("Type2", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 2 Expressed Response genes")

save_ggplot(last_plot(), "All_Type2")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Porites_compressa_HIv1___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

#### DESeq LRT Test

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds, test = "LRT", reduced = ~ treatment + time)

res <- results(dds)
sig_genes <- subset(res, padj < 0.05)
lrt_res <- as.data.frame(res)

DE_05 <- lrt_res[rownames(lrt_res %>% filter(padj<0.05)),]

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- levels(meta$time)

top_500_DE_genes <- DE_05 %>% arrange(padj) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(log2FoldChange) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(desc(log2FoldChange)) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

#### DE Heat stress genes

```{r}
plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_unique) %>% left_join(DE_05 %>% rownames_to_column(var="query")) %>%
  filter(!is.na(padj))

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("Type1", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "DE_Type1")

plot_df %>% filter(grepl("Type2", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,query))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_id, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 2 DE (LRT) Response genes")

save_ggplot(last_plot(), "DE_Type2")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Porites_compressa_HIv1___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Selected Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type1")

plot_df_norm <- plot_df %>%
  group_by(query, time) %>%
  mutate(delta = expression - mean(expression[treatment == "C"]))

plot_df_norm %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=delta, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Porites_compressa_HIv1___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Norm to control -- Selected Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type1_norm_control")

plot_df_norm <- plot_df %>%
  group_by(query, treatment) %>%
  mutate(baseline = mean(expression[time == "0"])) %>%
  mutate(delta = expression - baseline) %>%
  ungroup()

plot_df_norm %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>%
  ggplot(aes(x=time, y=delta, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Porites_compressa_HIv1___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Norm to T0 -- Selected Type 1 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type1_norm_T0")

plot_df %>% filter(grepl("GDH",gene_id)|grepl("GS",gene_id)|grepl("AMT1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(gene_id, ": ",str_replace(query,"Porites_compressa_HIv1___","")),scales="free_y") +
  theme_bw() +
  labs(y="VST expression", x="Timepoint", title = "Selected Type 2 DE (LRT) Response genes")

save_ggplot(last_plot(), "highlighted_DE_Type2")
```

#### Mechanosensing and other channels

```{r}
membrane_channels <- read_csv(paste0("../references/annotation/",species,"_membrane_channels.csv"))

membrane_channels <- membrane_channels %>% filter(query %in% rownames(vsd_mat))

plot_df <- as.data.frame(t(vsd_mat)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>% 
  pivot_longer(cols = all_of(rownames(vsd_mat)), names_to="query", values_to="expression") %>%
  mutate(is_DE = query %in% rownames(DE_05)) %>% right_join(membrane_channels)

membrane_list <- c("aquaporin","TRP",  "Mechanosensory" ,"calcium",  "ER_calcium" ,"Golgi_calcium" , "SLC",  "PMCA",  "Sodium_calcium_exchanger","Bhattacharya2016")

for (genelist in membrane_list){
  plot_df_filtered <- plot_df %>% filter(grepl(genelist,gene_set)) %>% filter(is_DE == FALSE)
  if (nrow(plot_df_filtered)>1){
    plot <- plot_df_filtered  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(short_name~paste0("DE by LRT: ",is_DE),scales="free_y") +
  theme_bw() +
  theme(strip.text = element_text(size = 6)) +
  labs(y="VST expression", x="Timepoint")
  print(plot)
  }
}

for (genelist in membrane_list){
  plot_df_filtered <- plot_df %>% filter(grepl(genelist,gene_set)) %>% filter(is_DE == TRUE)
  if (nrow(plot_df_filtered)>1){
    plot <- plot_df_filtered  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
    stat_summary(fun="mean", geom="line") +
    scale_color_manual(values = treat_colors) +
    stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
    facet_wrap(short_name~paste0("DE by LRT: ",is_DE),scales="free_y") +
    theme_bw() +
    theme(strip.text = element_text(size = 6)) +
    labs(y="VST expression", x="Timepoint")
    print(plot)
  }
}
```

### ImpulseDE2

#### Background info

Based on [this paper](https://academic.oup.com/bib/article/20/1/288/4364840#130283262), this is the best package to use other than comparing each time point against each other individually.

Repo here: https://github.com/YosefLab/ImpulseDE2

Tutorial here: http://bioconductor.statistik.tu-dortmund.de/packages/3.11/bioc/vignettes/ImpulseDE2/inst/doc/ImpulseDE2_Tutorial.html , I followed closely with the section "Case-control differential expression analysis"

Read the ImpulseDE2 paper [here](https://academic.oup.com/nar/article/46/20/e119/5068248)

David S Fischer, Fabian J Theis, Nir Yosef, Impulse model-based differential expression analysis of time course sequencing data, Nucleic Acids Research, Volume 46, Issue 20, 16 November 2018, Page e119, https://doi.org/10.1093/nar/gky675

#### 1. Install and load

```{r}
#library(devtools)
#install_github("YosefLab/ImpulseDE2")

library(ImpulseDE2)
library(ComplexHeatmap)
```

#### 2. Metadata formatting

First, reformat our metadata table to match the column names used in the ImpulseDE2 vignette.

```{r}
meta_impulse <- meta %>%
  dplyr::rename(Sample = sample, Time = time, Batch = replicate) %>% 
  mutate(Time = as.numeric(as.character(Time)),
         Condition = str_replace(treatment, "C", "control"),
         Condition = str_replace(Condition, "H", "case")
         ) %>%
  select(-c(species,treatment))
```

#### 3. Then, run ImpulseDE2

This takes a ton of time and memory, so I run it once then save as an RDS.

```{r, eval=FALSE} 
objectImpulseDE2 <- runImpulseDE2(
  matCountData    = as.matrix(counts_raw), #or use filtered_counts 
  dfAnnotation    = meta_impulse,
  boolCaseCtrl    = TRUE,
  vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
  boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
  scaNProc        = 18 )

saveRDS(objectImpulseDE2, file = paste0(outdir, "/objectImpulseDE2.rds"))
```

#### 4. View and save results

```{r}
objectImpulseDE2 <- readRDS(paste0(outdir, "/objectImpulseDE2.rds"))

impulse_results <- objectImpulseDE2$dfImpulseDE2Results
impulse_results_annot <- impulse_results %>% left_join(SwissProt, by = join_by("Gene"=="query")) %>% filter(!is.na(Gene))
write.table(impulse_results,file.path(outdir, "ImpulseDE2_Results.txt"),row.names=F,quote=F,sep="\t")

# Genes with significant treatment effect on temporal trajectory
impulse_sig_genes <- impulse_results %>% filter(padj < 0.05) 

#preview top DE genes 
impulse_sig_genes %>% arrange(padj) %>% head(5) %>% dplyr::select(!contains("converge"))

cat("\nTotal significant genes:", nrow(impulse_sig_genes), "\n")
cat("\nResponse patterns:\n")
cat("Transient:", sum(impulse_sig_genes$isTransient), "\n")
cat("Monotonous:", sum(impulse_sig_genes$isMonotonous), "\n")
cat("Complex:", sum(!impulse_sig_genes$isTransient & !impulse_sig_genes$isMonotonous), "\n")
```

#### 5. Heatmap of transient and non-transiently heat-affected genes

```{r}
lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "control",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.05)
draw(lsHeatmaps$complexHeatmapRaw) 
draw(lsHeatmaps$complexHeatmapFit) 

lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "combined",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.05)
draw(lsHeatmaps$complexHeatmapRaw) 
draw(lsHeatmaps$complexHeatmapFit) 

lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "case",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.05)
draw(lsHeatmaps$complexHeatmapRaw) 
draw(lsHeatmaps$complexHeatmapFit) 

png(paste0(outdir,"/ImpulseDE/ImpulseDE2_heatmap.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapRaw)
dev.off()

png(paste0(outdir,"/ImpulseDE/ImpulseDE2_heatmap_fit.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapFit)
dev.off()

str(lsHeatmaps$lsvecGeneGroups)
```

#### 6. Plot trajectories of top impulseDE genes and specific genes of interest

```{r}
# Plot top 10 differentially expressed (by q-value) genes
lsgplotsGenes <- plotGenes(
  vecGeneIDs       = NULL,
  scaNTopIDs       = 10,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

lsgplotsGenes
```

```{r}
# Majerova 2021 key qPCR genes
majerova_genes <- HeatStressGenes %>% filter(ref_first_author =="Majerova")
stress_genes_ids <- unique(majerova_genes$query)
plot_stress_genes <- stress_genes_ids[stress_genes_ids %in% rownames(objectImpulseDE2@matCountDataProc)] 

impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_unique, by = join_by(Gene==query))

heatgenes <- plotGenes(
  vecGeneIDs       = plot_stress_genes,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  strFileName = "stress_genes_Majerova.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

heatgenes

# HSP genes
HSPS <- impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_unique, by = join_by(Gene==query)) %>% filter(grepl("HSP",gene_id)) %>% pull(Gene)

HSPs <- plotGenes(
  vecGeneIDs       = HSPS,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut           = paste0(outdir,"/ImpulseDE/"),
  strFileName = "HSPs.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

HSPs
```

#### 7. Heatmap of top 500 impulseDE genes

```{r}
top_500_DE_genes <- impulse_results %>% arrange(padj) %>% head(500) %>% rownames()

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

#### 7.5. Heatmap of membrane channel genes

```{r}
membrane_DE_genes <- impulse_results %>%
  right_join(membrane_channels, join_by("Gene"=="query")) %>% 
  filter(gene_set != "SLC") %>%
  filter(padj < 0.05) %>% 
  column_to_rownames("Gene")

plot_df <- as.data.frame(t(vsd_mat)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>% 
  pivot_longer(cols = all_of(rownames(vsd_mat)), names_to="query", values_to="expression") %>%
  mutate(is_DE = query %in% rownames(DE_05)) %>% right_join(membrane_DE_genes %>% rownames_to_column("query"))

membrane_list <- c("aquaporin","TRP",  "Mechanosensory" ,"calcium",  "ER_calcium" ,"Golgi_calcium" , "SLC",  "PMCA",  "Sodium_calcium_exchanger","Bhattacharya2016")

pdf(paste0(outdir,"/ImpulseDE/membrane_sig_plots.pdf"), width = 14, height = 14)

pheatmap(assay(vsd)[rownames(membrane_DE_genes), ], cluster_rows=TRUE, show_rownames=TRUE,
         cluster_cols=FALSE,
         labels_row=membrane_DE_genes$short_name,
         annotation_row = membrane_DE_genes %>% dplyr::select(gene_set),
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[rownames(membrane_DE_genes), ], cluster_rows=TRUE, show_rownames=TRUE,
          cluster_cols=TRUE, cutree_cols = 2,
         fontsize = 5,
         labels_row=membrane_DE_genes$short_name,
         annotation_row = membrane_DE_genes %>% dplyr::select(gene_set),
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))


for (genelist in unique(plot_df$gene_set)){
  plot_df_filtered <- plot_df %>% filter(grepl(genelist,gene_set))# %>% filter(is_DE == TRUE)
  if (nrow(plot_df_filtered)>1){
    plot <- plot_df_filtered  %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
    stat_summary(fun="mean", geom="line") +
    scale_color_manual(values = treat_colors) +
    stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
    facet_wrap(short_name~query,scales="free_y") +
    theme_bw() +
    theme(strip.text = element_text(size = 6)) +
    labs(y="VST expression", x="Timepoint",title=paste(genelist))
    print(plot)
  }
}

dev.off()
```

