---
title: "DE"
author: "Zoe Dellaert"
date: "2025-11-20"
output:
  github_document: default
---

## Differential expression analysis of Time Series Bulk RNA Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Thoughts and Notes

- [This paper](https://academic.oup.com/bib/article/20/1/288/4364840) 

Daniel Spies, Peter F Renz, Tobias A Beyer, Constance Ciaudo, Comparative analysis of differential gene expression tools for RNA sequencing time course data, Briefings in Bioinformatics, Volume 20, Issue 1, January 2019, Pages 288–298, https://doi.org/10.1093/bib/bbx115


## Load packages

```{r packages}
library("genefilter")
library("DESeq2")
library("apeglm")
library("ggplot2")
library("pheatmap")
library("ggnewscale")
library("RColorBrewer")
library("tidyverse")

sessionInfo() #provides list of loaded packages and version of R.
```

```{r}
#set standard output directory for figures
outdir <- "../output_RNA/differential_expression"

save_ggplot <- function(plot, filename, width = 10, height = 7, units = "in", dpi = 300,bg=NULL) {
  print(plot)

  png_path <- file.path(outdir, paste0(filename, ".png"))
  pdf_dir <- file.path(outdir, "pdf_figs")
  pdf_path <- file.path(pdf_dir, paste0(filename, ".pdf"))
  
  # Ensure the pdf_figs directory exists
  if (!dir.exists(pdf_dir)) dir.create(pdf_dir, recursive = TRUE)
  
  # Save plots
  ggsave(filename = png_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
  ggsave(filename = pdf_path, plot = plot, width = width, height = height, units = units, dpi = dpi,bg = bg)
}

treat_colors <- c("C" = "lightblue4", "H" = "#D55E00")
```


## POC: pre-processing and visualization

Read in raw count data
```{r}
#set standard output directory for figures
outdir <- "../output_RNA/differential_expression/POC_PacutaV2"

counts_raw <- read.csv("../output_RNA/count_matrices/POC_PacutaV2_gene_count_matrix.csv", row.names = 1) #load in data

samples <- colnames(counts_raw)
```

Read in metadata 
```{r}
meta <- data.frame(
  sample = samples, 
  species = str_split(samples, "_", simplify = TRUE)[,1], #extract first part of sample name to get species
  time = str_replace(str_split(samples, "_", simplify = TRUE)[,2],"R", ""), #extract "R##" part to get timepoint then remove R
  replicate = str_split(samples, "_", simplify = TRUE)[,3], #extract "R##" part to get timepoint then remove R
  treatment = str_replace(str_split(samples, "_", simplify = TRUE)[,3],"\\d", "")
)

rownames(meta) <- meta$sample

meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
meta$treatment <- factor(meta$treatment)

meta <- meta %>% arrange(time, treatment)
write.csv(meta, paste0(outdir,"/RNA_seq_metadata.csv"))
```

Reorder sample columns based on factor order
```{r}
counts_raw <- counts_raw[, meta$sample]
```


Data sanity checks!
```{r}
stopifnot(all(meta$sample %in% colnames(counts_raw))) #are all of the sample names in the metadata column names in the gene count matrix?
stopifnot(all(meta$sample == colnames(counts_raw))) #are they the same in the same order?
```

pOverA filtering to reduce dataset

```{r}
ffun<-filterfun(pOverA(0.07,10))  # Keep genes expressed at 10+ counts in at least 7% of samples - expressed in all 3 samples at one timepoint from one treatment
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter

paste0("Number of genes after filtering: ", sum(counts_filt_poa))

write.csv(filtered_counts, file = file.path(outdir, "filtered_counts.csv"))
```

### [DESeq2](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

Create DESeq object and run DESeq2
```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds)
```

Check size factors. 
```{r}
SF.dds <- estimateSizeFactors(dds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.dds)) #View size factors

all(sizeFactors(SF.dds)) < 4
```

Transforming count data for visualization 

```{r}
vsd <- vst(dds, blind=FALSE)

#save the vsd transformation
vsd_mat <- assay(vsd)
write.csv(vsd_mat, file = file.path(outdir, "vsd_expression_matrix.csv"))
```

### Heatmap of the sample-to-sample distances
```{r}
sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix(sampleDists)
colnames(sampleDistMatrix) <- NULL

pheatmap(sampleDistMatrix,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))
```

### Principal component plot of the samples

```{r}
pcaData <- plotPCA(vsd, intgroup=c("time", "treatment"), returnData=TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))
PCA <- ggplot() +
  geom_point(data = subset(pcaData, treatment == "C"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Blues"), name = "Time (hrs) - Control") +
  
  #start new scale
  ggnewscale::new_scale_color() +
  geom_point(data = subset(pcaData, treatment == "H"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Oranges"), name = "Time (hrs) - Heat") +

  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()
PCA
save_ggplot(PCA, "PCA_POC")
```

### Heatmap of count matrix

```{r}
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing=TRUE), 500)

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- levels(meta$time)

#view top 500 most vairable genes
pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

### Heat stress genes

```{r}
HeatStressGenes_Pacuta <- read_csv("/project/pi_hputnam_uri_edu/zdellaert/snRNA_analysis/multi-sp-snRNA/reference_genes/genes_of_interest/HeatStressGenes_Pacuta.csv") %>% dplyr::select(-1) %>% dplyr::rename(query = Pacuta_gene) %>% dplyr::select(query,everything()) #%>% filter(ref_first_author =="Majerova")

HeatStressGenes_Pacuta_unique <- HeatStressGenes_Pacuta %>% group_by(query) %>%
  summarize(gene_id = paste(unique(gene_id), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")
            ) 

HeatStressGenes_Pacuta_unique <- HeatStressGenes_Pacuta_unique %>% filter(query %in% rownames(vsd_mat))
 
stress_genes_ids <- unique(HeatStressGenes_Pacuta_unique$query) 
stress_genes_vsd <- vsd_mat[stress_genes_ids, ]

plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_Pacuta_unique)

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(gene_id~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Pocillopora_acuta_HIv2___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

### DESeq LRT Test

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds, test = "LRT", reduced = ~ treatment + time)

res <- results(dds)
sig_genes <- subset(res, padj < 0.05)
lrt_res <- as.data.frame(res)

DE_05 <- lrt_res[rownames(lrt_res %>% filter(padj<0.05)),]

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- levels(meta$time)

top_500_DE_genes <- DE_05 %>% arrange(padj) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(log2FoldChange) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(desc(log2FoldChange)) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```


### DE Heat stress genes

```{r}
plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_Pacuta_unique) %>% left_join(DE_05 %>% rownames_to_column(var="query")) %>%
  filter(!is.na(padj))

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(gene_id~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Pocillopora_acuta_HIv2___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

### ImpulseDE2

Based on [this paper](https://academic.oup.com/bib/article/20/1/288/4364840#130283262), this is the best package to use other than comparing each time point against each other individually. I am also planning to ID gene modules via WGCNA. 

Repo here: https://github.com/YosefLab/ImpulseDE2

Tutorial here: http://bioconductor.statistik.tu-dortmund.de/packages/3.11/bioc/vignettes/ImpulseDE2/inst/doc/ImpulseDE2_Tutorial.html , I followed closely with the section "Case-control differential expression analysis"

Read the ImpulseDE2 paper [here](https://academic.oup.com/nar/article/46/20/e119/5068248)

David S Fischer, Fabian J Theis, Nir Yosef, Impulse model-based differential expression analysis of time course sequencing data, Nucleic Acids Research, Volume 46, Issue 20, 16 November 2018, Page e119, https://doi.org/10.1093/nar/gky675

```{r}
#library(devtools)
#install_github("YosefLab/ImpulseDE2")

library(ImpulseDE2)
```

First, reformat our metadata table to match the column names used in the ImpulseDE2 vignette.

```{r}
meta_impulse <- meta %>%
  dplyr::rename(Sample = sample, Time = time, Batch = replicate) %>% 
  mutate(Time = as.numeric(as.character(Time)),
         #Time = as.numeric(Time),
         Condition = str_replace(treatment, "C", "control"),
         Condition = str_replace(Condition, "H", "case")
         ) %>%
  select(-c(species,treatment))
```

Then, generate the ImpulseDE2 object

```{r, eval=FALSE} 
#test with just 500 genes that I determined to be DE by treatment/timepoint with DESeq2 
objectImpulseDE2 <- runImpulseDE2(
  matCountData    = as.matrix(filtered_counts)[top_500_DE_genes,], #or use filtered_counts 
  dfAnnotation    = meta_impulse,
  boolCaseCtrl    = TRUE,
  vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
  boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
  scaNProc        = 8 )

#run with all genes
objectImpulseDE2 <- runImpulseDE2(
  matCountData    = as.matrix(counts_raw), #or use filtered_counts 
  dfAnnotation    = meta_impulse,
  boolCaseCtrl    = TRUE,
  vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
  boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
  scaNProc        = 18 )

saveRDS(objectImpulseDE2, file = paste0(outdir, "/objectImpulseDE2.rds"))
```

```{r}
objectImpulseDE2 <- readRDS(paste0(outdir, "/objectImpulseDE2.rds"))

impulse_results <- objectImpulseDE2$dfImpulseDE2Results
head(impulse_results)
write.table(impulse_results,file.path(outdir, "ImpulseDE2_Results.txt"),row.names=F,quote=F,sep="\t")

# Genes with significant treatment effect on temporal trajectory
sig_genes <- impulse_results[impulse_results$padj < 0.05 & 
                               impulse_results$loglik_full > impulse_results$loglik_red, ]

nrow(sig_genes)
head(sig_genes[order(sig_genes$padj), ])
```


```{r}
library(ComplexHeatmap)

lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "case",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.01)
draw(lsHeatmaps$complexHeatmapRaw) 

png(paste0(outdir,"/ImpulseDE/ImpulseDE2_heatmap.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapRaw)
dev.off()
```


```{r}
majerova_genes <- HeatStressGenes_Pacuta %>% filter(ref_first_author =="Majerova")
stress_genes_ids <- unique(majerova_genes$query)
plot_stress_genes <- stress_genes_ids[stress_genes_ids %in% rownames(objectImpulseDE2@matCountDataProc)] 

impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_Pacuta_unique, by = join_by(Gene==query))

heatgenes <- plotGenes(
  vecGeneIDs       = plot_stress_genes,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = "../output_RNA/differential_expression/POC_PacutaV2/ImpulseDE/",
  strFileName = "stress_genes_Majerova.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

heatgenes

HSPS <- impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_Pacuta_unique, by = join_by(Gene==query)) %>% filter(grepl("HSP",gene_id)) %>% pull(Gene)

HSPs <- plotGenes(
  vecGeneIDs       = HSPS,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut           = "../output_RNA/differential_expression/POC_PacutaV2/ImpulseDE/",
  strFileName = "HSPs.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

HSPs

lsgplotsGenes <- plotGenes(
  vecGeneIDs       = NULL,
  scaNTopIDs       = 10,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = "../output_RNA/differential_expression/POC_PacutaV2/ImpulseDE/",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

lsgplotsGenes
```

This is so freaking cool!!!! TNFRs and HSP are going crazy :) 


```{r}
top_500_DE_genes <- impulse_results %>% arrange(padj) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, #cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

```


## MON: pre-processing and visualization

Read in raw count data
```{r}
#set standard output directory for figures
outdir <- "../output_RNA/differential_expression/MON_MCapV3"

counts_raw <- read.csv("../output_RNA/count_matrices/MON_MCapV3_gene_count_matrix.csv", row.names = 1) #load in data

samples <- colnames(counts_raw)
```

Read in metadata 
```{r}
meta <- data.frame(
  sample = samples, 
  species = str_split(samples, "_", simplify = TRUE)[,1], #extract first part of sample name to get species
  time = str_replace(str_split(samples, "_", simplify = TRUE)[,2],"R", ""), #extract "R##" part to get timepoint then remove R
  replicate = str_split(samples, "_", simplify = TRUE)[,3], #extract "R##" part to get timepoint then remove R
  treatment = str_replace(str_split(samples, "_", simplify = TRUE)[,3],"\\d", "")
)

rownames(meta) <- meta$sample

meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
meta$treatment <- factor(meta$treatment)

meta <- meta %>% arrange(time, treatment)
write.csv(meta, paste0(outdir,"/RNA_seq_metadata.csv"))
```

Reorder sample columns based on factor order
```{r}
counts_raw <- counts_raw[, meta$sample]
```

Remove outliers:
```{r}
counts_raw <- counts_raw[, !(colnames(counts_raw) %in% c("MON_R72_H1","MON_R72_H2"))]
meta <- meta[!(rownames(meta) %in% c("MON_R72_H1","MON_R72_H2")),]
```

Data sanity checks!
```{r}
stopifnot(all(meta$sample %in% colnames(counts_raw))) #are all of the sample names in the metadata column names in the gene count matrix?
stopifnot(all(meta$sample == colnames(counts_raw))) #are they the same in the same order?
```

pOverA filtering to reduce dataset

```{r}
ffun<-filterfun(pOverA(0.07,10))  # Keep genes expressed at 10+ counts in at least 7% of samples - expressed in all 3 samples at one timepoint from one treatment
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter

paste0("Number of genes after filtering: ", sum(counts_filt_poa))

write.csv(filtered_counts, file = file.path(outdir, "filtered_counts.csv"))
```

### [DESeq2](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

Create DESeq object and run DESeq2
```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds)
```

Check size factors. 
```{r}
SF.dds <- estimateSizeFactors(dds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.dds)) #View size factors

all(sizeFactors(SF.dds)) < 4
```

Transforming count data for visualization 

```{r}
vsd <- vst(dds, blind=FALSE)

#save the vsd transformation
vsd_mat <- assay(vsd)
write.csv(vsd_mat, file = file.path(outdir, "vsd_expression_matrix.csv"))
```

### Heatmap of the sample-to-sample distances
```{r}
sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix(sampleDists)
colnames(sampleDistMatrix) <- NULL

pheatmap(sampleDistMatrix,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))
```

### Principal component plot of the samples

```{r}
pcaData <- plotPCA(vsd, intgroup=c("time", "treatment"), returnData=TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))
PCA <- ggplot() +
  geom_point(data = subset(pcaData, treatment == "C"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Blues"), name = "Time (hrs) - Control") +
  
  #start new scale
  ggnewscale::new_scale_color() +
  geom_point(data = subset(pcaData, treatment == "H"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Oranges"), name = "Time (hrs) - Heat") +

  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()
PCA
save_ggplot(PCA, "PCA_MON")
```

### Heatmap of count matrix

```{r}
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing=TRUE), 500)

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- levels(meta$time)

#view top 500 most vairable genes
pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```


### Heat stress genes

```{r}
HeatStressGenes_Mcap <- read_csv("/project/pi_hputnam_uri_edu/zdellaert/snRNA_analysis/multi-sp-snRNA/reference_genes/genes_of_interest/HeatStressGenes_Mcap.csv") %>% dplyr::select(-1) %>% dplyr::rename(query = Mcap_gene) %>% dplyr::select(query,everything()) #%>% filter(ref_first_author =="Majerova")

HeatStressGenes_Mcap_unique <- HeatStressGenes_Mcap %>% group_by(query) %>%
  summarize(gene_id = paste(unique(gene_id), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")
            ) 

HeatStressGenes_Mcap_unique <- HeatStressGenes_Mcap_unique %>% filter(query %in% rownames(vsd_mat))
 
stress_genes_ids <- unique(HeatStressGenes_Mcap_unique$query) 
stress_genes_vsd <- vsd_mat[stress_genes_ids, ]

plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_Mcap_unique)

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(gene_id~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Montipora_capitata_HIv3___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

### DESeq LRT Test

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds, test = "LRT", reduced = ~ treatment + time)

res <- results(dds)
sig_genes <- subset(res, padj < 0.05)
lrt_res <- as.data.frame(res)

DE_05 <- lrt_res[rownames(lrt_res %>% filter(padj<0.05)),]

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- levels(meta$time)

top_500_DE_genes <- DE_05 %>% arrange(padj) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(log2FoldChange) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(desc(log2FoldChange)) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```


### DE Heat stress genes

```{r}
plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_Mcap_unique) %>% left_join(DE_05 %>% rownames_to_column(var="query")) %>%
  filter(!is.na(padj))

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(gene_id~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Montipora_capitata_HIv3___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

### ImpulseDE2

Based on [this paper](https://academic.oup.com/bib/article/20/1/288/4364840#130283262), this is the best package to use other than comparing each time point against each other individually. I am also planning to ID gene modules via WGCNA. 

Repo here: https://github.com/YosefLab/ImpulseDE2

Tutorial here: http://bioconductor.statistik.tu-dortmund.de/packages/3.11/bioc/vignettes/ImpulseDE2/inst/doc/ImpulseDE2_Tutorial.html , I followed closely with the section "Case-control differential expression analysis"

Read the ImpulseDE2 paper [here](https://academic.oup.com/nar/article/46/20/e119/5068248)

David S Fischer, Fabian J Theis, Nir Yosef, Impulse model-based differential expression analysis of time course sequencing data, Nucleic Acids Research, Volume 46, Issue 20, 16 November 2018, Page e119, https://doi.org/10.1093/nar/gky675

```{r}
#library(devtools)
#install_github("YosefLab/ImpulseDE2")

library(ImpulseDE2)
```

First, reformat our metadata table to match the column names used in the ImpulseDE2 vignette.

```{r}
meta_impulse <- meta %>%
  dplyr::rename(Sample = sample, Time = time, Batch = replicate) %>% 
  mutate(Time = as.numeric(as.character(Time)),
         #Time = as.numeric(Time),
         Condition = str_replace(treatment, "C", "control"),
         Condition = str_replace(Condition, "H", "case")
         ) %>%
  select(-c(species,treatment))
```

Then, generate the ImpulseDE2 object

```{r, eval=FALSE} 
#test with just 500 genes that I determined to be DE by treatment/timepoint with DESeq2 
objectImpulseDE2 <- runImpulseDE2(
  matCountData    = as.matrix(filtered_counts)[top_500_DE_genes,], #or use filtered_counts 
  dfAnnotation    = meta_impulse,
  boolCaseCtrl    = TRUE,
  vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
  boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
  scaNProc        = 8 )

#run with all genes
objectImpulseDE2 <- runImpulseDE2(
  matCountData    = as.matrix(counts_raw), #or use filtered_counts 
  dfAnnotation    = meta_impulse,
  boolCaseCtrl    = TRUE,
  vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
  boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
  scaNProc        = 18 )

saveRDS(objectImpulseDE2, file = paste0(outdir, "/objectImpulseDE2.rds"))
```

```{r}
objectImpulseDE2 <- readRDS(paste0(outdir, "/objectImpulseDE2.rds"))

impulse_results <- objectImpulseDE2$dfImpulseDE2Results
head(impulse_results)
write.table(impulse_results,file.path(outdir, "ImpulseDE2_Results.txt"),row.names=F,quote=F,sep="\t")

# Genes with significant treatment effect on temporal trajectory
sig_genes <- impulse_results[impulse_results$padj < 0.05 & 
                               impulse_results$loglik_full > impulse_results$loglik_red, ]

nrow(sig_genes)
head(sig_genes[order(sig_genes$padj), ])
```


```{r}
library(ComplexHeatmap)

lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "case",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.01)
draw(lsHeatmaps$complexHeatmapRaw) 

png(paste0(outdir,"/ImpulseDE/ImpulseDE2_heatmap.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapRaw)
dev.off()
```


```{r}
majerova_genes <- HeatStressGenes_Mcap %>% filter(ref_first_author =="Majerova")
stress_genes_ids <- unique(majerova_genes$query)
plot_stress_genes <- stress_genes_ids[stress_genes_ids %in% rownames(objectImpulseDE2@matCountDataProc)]

impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_Mcap_unique, by = join_by(Gene==query))

heatgenes <- plotGenes(
  vecGeneIDs       = plot_stress_genes,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = "../output_RNA/differential_expression/MON_MCapV3/ImpulseDE/",
  strFileName = "stress_genes_Majerova.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

heatgenes

HSPS <- impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_Mcap_unique, by = join_by(Gene==query)) %>% filter(grepl("HSP",gene_id)) %>% pull(Gene)

HSPs <- plotGenes(
  vecGeneIDs       = HSPS,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut           = "../output_RNA/differential_expression/MON_MCapV3/ImpulseDE/",
  strFileName = "HSPs.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

HSPs

lsgplotsGenes <- plotGenes(
  vecGeneIDs       = NULL,
  scaNTopIDs       = 10,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = "../output_RNA/differential_expression/MON_MCapV3/ImpulseDE/",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

lsgplotsGenes
```



```{r}
top_500_DE_genes <- impulse_results %>% arrange(padj) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(vsd_mat[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, #cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(vsd_mat[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

```

## POR: pre-processing and visualization

Read in raw count data
```{r}
#set standard output directory for figures
outdir <- "../output_RNA/differential_expression/POR_Pcomp"

counts_raw <- read.csv("../output_RNA/count_matrices/POR_Pcomp_gene_count_matrix.csv", row.names = 1) #load in data

samples <- colnames(counts_raw)
```

### Read in metadata 
```{r}
meta <- data.frame(
  sample = samples, 
  species = str_split(samples, "_", simplify = TRUE)[,1], #extract first part of sample name to get species
  time = str_replace(str_split(samples, "_", simplify = TRUE)[,2],"R", ""), #extract "R##" part to get timepoint then remove R
  replicate = str_split(samples, "_", simplify = TRUE)[,3], #extract "R##" part to get timepoint then remove R
  treatment = str_replace(str_split(samples, "_", simplify = TRUE)[,3],"\\d", "")
)

rownames(meta) <- meta$sample

meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
meta$treatment <- factor(meta$treatment)

meta <- meta %>% arrange(time, treatment)
write.csv(meta, paste0(outdir,"/RNA_seq_metadata.csv"))
```

Reorder sample columns based on factor order
```{r}
counts_raw <- counts_raw[, meta$sample]
```

Remove outliers:
```{r}
#counts_raw <- counts_raw[, !(colnames(counts_raw) %in% c("MON_R72_H1","MON_R72_H2"))]
#meta <- meta[!(rownames(meta) %in% c("MON_R72_H1","MON_R72_H2")),]
```

Data sanity checks!
```{r}
stopifnot(all(meta$sample %in% colnames(counts_raw))) #are all of the sample names in the metadata column names in the gene count matrix?
stopifnot(all(meta$sample == colnames(counts_raw))) #are they the same in the same order?
```

### pOverA filtering to reduce dataset

```{r}
ffun<-filterfun(pOverA(0.07,10))  # Keep genes expressed at 10+ counts in at least 7% of samples - expressed in all 3 samples at one timepoint from one treatment
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter

paste0("Number of genes after filtering: ", sum(counts_filt_poa))

write.csv(filtered_counts, file = file.path(outdir, "filtered_counts.csv"))
```

### [DESeq2](https://www.bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

Create DESeq object and run DESeq2
```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds)
```

Check size factors. 
```{r}
SF.dds <- estimateSizeFactors(dds) #estimate size factors to determine if we can use vst  to transform our data. Size factors should be less than 4 for us to use vst
print(sizeFactors(SF.dds)) #View size factors

all(sizeFactors(SF.dds)) < 4
```

Transforming count data for visualization 

```{r}
vsd <- vst(dds, blind=FALSE)

#save the vsd transformation
vsd_mat <- assay(vsd)
write.csv(vsd_mat, file = file.path(outdir, "vsd_expression_matrix.csv"))
```

### Heatmap of the sample-to-sample distances
```{r}
sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix(sampleDists)
colnames(sampleDistMatrix) <- NULL

pheatmap(sampleDistMatrix,
         col=colorRampPalette( rev(brewer.pal(9, "Blues")) )(255))
```

### Principal component plot of the samples

```{r}
pcaData <- plotPCA(vsd, intgroup=c("time", "treatment"), returnData=TRUE)

percentVar <- round(100 * attr(pcaData, "percentVar"))
PCA <- ggplot() +
  geom_point(data = subset(pcaData, treatment == "C"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Blues"), name = "Time (hrs) - Control") +
  
  #start new scale
  ggnewscale::new_scale_color() +
  geom_point(data = subset(pcaData, treatment == "H"),
             aes(x=PC1, y=PC2, color=time),
                 size=2) +
             scale_color_manual(values=brewer.pal(7, "Oranges"), name = "Time (hrs) - Heat") +

  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw()
PCA
save_ggplot(PCA, "PCA_POR")
```
The following samples had mapping rates < 20%:

| sample      | uniquely_mapped_percent | rRNA_matched_reads_Million | total_reads_Million |
|-------------|-------------------------|----------------------------|---------------------|
| POR_R72_H1  | 6.4                     | 12.92                      | 15.47               |
| POR_R24_H1  | 7.67                    | 14.23                      | 17.26               |
| POR_R120_C3 | 7.77                    | 18.64                      | 21.02               |
| POR_R72_H2  | 8.77                    | 12.24                      | 15.16               |
| POR_R1_H2   | 10.37                   | 14.78                      | 17.47               |
| POR_R120_C1 | 10.78                   | 11.68                      | 13.99               |
| POR_R1_H1   | 11.72                   | 15.47                      | 18.69               |
| POR_R1_C1   | 12.04                   | 13.14                      | 15.99               |
| POR_R12_C1  | 12.2                    | 12.17                      | 14.88               |
| POR_R12_C3  | 15.78                   | 11.7                       | 15.09               |
| POR_R0_H1   | 16.57                   | 12.57                      | 16.46               |
| POR_R120_H3 | 18.23                   | 12.18                      | 17.16               |
| POR_R3_C3   | 18.27                   | 11                         | 14.95               |
| POR_R0_H2   | 18.35                   | 14.14                      | 19.31               |
| POR_R24_C2  | 19.72                   | 11.55                      | 16                  |

### Heatmap of count matrix

```{r}
topVarGenes <- head(order(rowVars(assay(vsd)), decreasing=TRUE), 500)

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- levels(meta$time)

#view top 500 most vairable genes
pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```


### Heat stress genes

```{r}
HeatStressGenes_Pcomp <- read_csv("/project/pi_hputnam_uri_edu/zdellaert/snRNA_analysis/multi-sp-snRNA/reference_genes/genes_of_interest/HeatStressGenes_Pcomp.csv") %>% dplyr::select(-1) %>% dplyr::rename(query = Pcomp_gene) %>% dplyr::select(query,everything()) #%>% filter(ref_first_author =="Majerova")

HeatStressGenes_Pcomp_unique <- HeatStressGenes_Pcomp %>% group_by(query) %>%
  summarize(gene_id = paste(unique(gene_id), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")
            ) 

HeatStressGenes_Pcomp_unique <- HeatStressGenes_Pcomp_unique %>% filter(query %in% rownames(vsd_mat))
 
stress_genes_ids <- unique(HeatStressGenes_Pcomp_unique$query) 
stress_genes_vsd <- vsd_mat[stress_genes_ids, ]

plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_Pcomp_unique)

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(gene_id~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Porites_compressa_HIv1___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

### DESeq LRT Test

```{r}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds, test = "LRT", reduced = ~ treatment + time)

res <- results(dds)
sig_genes <- subset(res, padj < 0.05)
lrt_res <- as.data.frame(res)

DE_05 <- lrt_res[rownames(lrt_res %>% filter(padj<0.05)),]

time_colors <- colorRampPalette(c("#ffffcc","#0c2c84"))(7)
names(time_colors) <- levels(meta$time)

top_500_DE_genes <- DE_05 %>% arrange(padj) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(log2FoldChange) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

top_500_DE_genes <- DE_05 %>% arrange(desc(log2FoldChange)) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(assay(vsd)[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```


### DE Heat stress genes

```{r}
plot_df <- as.data.frame(t(stress_genes_vsd)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="query", values_to="expression") %>%
  left_join(HeatStressGenes_Pcomp_unique) %>% left_join(DE_05 %>% rownames_to_column(var="query")) %>%
  filter(!is.na(padj))

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_id, group=gene_id)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(gene_id~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("HSP",gene_id)|grepl("Nrf2",gene_id)|grepl("HSF1",gene_id)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(query,"Porites_compressa_HIv1___",""), ": ", gene_id)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

### ImpulseDE2

Based on [this paper](https://academic.oup.com/bib/article/20/1/288/4364840#130283262), this is the best package to use other than comparing each time point against each other individually. I am also planning to ID gene modules via WGCNA. 

Repo here: https://github.com/YosefLab/ImpulseDE2

Tutorial here: http://bioconductor.statistik.tu-dortmund.de/packages/3.11/bioc/vignettes/ImpulseDE2/inst/doc/ImpulseDE2_Tutorial.html , I followed closely with the section "Case-control differential expression analysis"

Read the ImpulseDE2 paper [here](https://academic.oup.com/nar/article/46/20/e119/5068248)

David S Fischer, Fabian J Theis, Nir Yosef, Impulse model-based differential expression analysis of time course sequencing data, Nucleic Acids Research, Volume 46, Issue 20, 16 November 2018, Page e119, https://doi.org/10.1093/nar/gky675

```{r}
#library(devtools)
#install_github("YosefLab/ImpulseDE2")

library(ImpulseDE2)
```

First, reformat our metadata table to match the column names used in the ImpulseDE2 vignette.

```{r}
meta_impulse <- meta %>%
  dplyr::rename(Sample = sample, Time = time, Batch = replicate) %>% 
  mutate(Time = as.numeric(as.character(Time)),
         #Time = as.numeric(Time),
         Condition = str_replace(treatment, "C", "control"),
         Condition = str_replace(Condition, "H", "case")
         ) %>%
  select(-c(species,treatment))
```

Then, generate the ImpulseDE2 object

```{r, eval=FALSE} 
#test with just 500 genes that I determined to be DE by treatment/timepoint with DESeq2 
objectImpulseDE2 <- runImpulseDE2(
  matCountData    = as.matrix(filtered_counts)[top_500_DE_genes,], #or use filtered_counts 
  dfAnnotation    = meta_impulse,
  boolCaseCtrl    = TRUE,
  vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
  boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
  scaNProc        = 8 )

#run with all genes
objectImpulseDE2 <- runImpulseDE2(
  matCountData    = as.matrix(counts_raw), #or use filtered_counts 
  dfAnnotation    = meta_impulse,
  boolCaseCtrl    = TRUE,
  vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
  boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
  scaNProc        = 18 )

saveRDS(objectImpulseDE2, file = paste0(outdir, "/objectImpulseDE2.rds"))
```

```{r}
objectImpulseDE2 <- readRDS(paste0(outdir, "/objectImpulseDE2.rds"))

impulse_results <- objectImpulseDE2$dfImpulseDE2Results
head(impulse_results)
write.table(impulse_results,file.path(outdir, "ImpulseDE2_Results.txt"),row.names=F,quote=F,sep="\t")

# Genes with significant treatment effect on temporal trajectory
sig_genes <- impulse_results[impulse_results$padj < 0.05 & 
                               impulse_results$loglik_full > impulse_results$loglik_red, ]

nrow(sig_genes)
head(sig_genes[order(sig_genes$padj), ])
```


```{r}
library(ComplexHeatmap)

lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "case",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = 0.01)
draw(lsHeatmaps$complexHeatmapRaw) 

png(paste0(outdir,"/ImpulseDE/ImpulseDE2_heatmap.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapRaw)
dev.off()
```


```{r}
majerova_genes <- HeatStressGenes_Pcomp %>% filter(ref_first_author =="Majerova")
stress_genes_ids <- unique(majerova_genes$query)
plot_stress_genes <- stress_genes_ids[stress_genes_ids %in% rownames(objectImpulseDE2@matCountDataProc)]

impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_Pcomp_unique, by = join_by(Gene==query))

heatgenes <- plotGenes(
  vecGeneIDs       = plot_stress_genes,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,   boolCaseCtrl     = TRUE,
  dirOut           = "../output_RNA/differential_expression/POR_Pcomp/ImpulseDE/",
  strFileName = "stress_genes_Majerova.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

heatgenes

HSPS <- impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_Pcomp_unique, by = join_by(Gene==query)) %>% filter(grepl("HSP",gene_id)) %>% pull(Gene)

HSPs <- plotGenes(
  vecGeneIDs       = HSPS,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut           = "../output_RNA/differential_expression/POR_Pcomp/ImpulseDE/",
  strFileName = "HSPs.pdf",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

HSPs

lsgplotsGenes <- plotGenes(
  vecGeneIDs       = NULL,
  scaNTopIDs       = 10,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut           = "../output_RNA/differential_expression/POR_Pcomp/ImpulseDE/",
  boolMultiplePlotsPerPage = FALSE,
  strNameRefMethod = NULL)

lsgplotsGenes
```



```{r}
top_500_DE_genes <- impulse_results %>% arrange(padj) %>% head(500) %>% rownames()

#view top 500 most vairable genes
pheatmap(vsd_mat[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, #cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(vsd_mat[top_500_DE_genes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

```
