---
title: "RNA-seq QC Report: Time Series Bulk RNA-seq"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    html_preview: false
---

```{r setup, include=FALSE}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath("../../output_RNA/reports/"), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 10, fig.height = 7,
                      fig.path = "00_library_QC_files/figure-gfm/")

#load packages
library(tidyverse)

sessionInfo() #provides list of loaded packages and version of R
```

## 0. Read in all data

```{r}
# set up necessary output directories if they don't exist
reportdir <- "../../output_RNA/reports/00_library_QC_files/figure-gfm/"
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)

# load in multiqc data of raw reads
QC_raw <- read.delim("../../output_RNA/raw_qc/multiqc_data_1/multiqc_fastqc.txt") %>%
  # remove run-2- and run-3_ (yes the underscore and dash were different in the original files) from the sample names to group by sample
  mutate(Sample=str_replace(Sample, "run-[23][-_]", "") %>% str_remove("_S.*"))  %>%
  # for each sample, get the total # of reads for both directions from all runs together
  group_by(Sample) %>%
  summarise(total_reads_raw = sum(Total.Sequences),n_fastqs = n(),.groups = "drop")

# load in multiqc data of trimmed reads
QC_trim <- read.delim("../../output_RNA/trimmed_qc/multiqc_data_1/multiqc_fastqc.txt") %>%
  #remove weird column name prefixes from fastqc
  rename_with(~str_remove(., "FastQC_mqc.generalstats.fastqc."), starts_with("FastQC_mqc.generalstats.fastqc.")) %>%
  # remove run-2- and run-3_ (yes the underscore and dash were different in the original files) from the sample names to group by sample
  mutate(Sample=str_replace(Sample, "run-[23][-_]", "") %>% str_remove("_S.*"))  %>%
  # for each sample, get the total # of reads for both directions from all runs together
  group_by(Sample) %>%
  summarise(total_reads_trim = sum(Total.Sequences),.groups = "drop")

# host mapping rates
host_map <- bind_rows(
  read.delim("../../output_RNA/alignment_qc/POR_Pcomp/multiqc_data/multiqc_star.txt") %>% mutate(species="POR",genome="Pcomp"),
  read.delim("../../output_RNA/alignment_qc/MON_MCapV3/multiqc_data/multiqc_star.txt") %>% mutate(species="MON",genome="MCapV3"),
  read.delim("../../output_RNA/alignment_qc/POC_PacutaV2/multiqc_data/multiqc_star.txt") %>% mutate(species="POC",genome="PacutaV2")) %>%
  rename(total_read_pairs_mapped=total_reads) %>% mutate(total_reads_Million = total_read_pairs_mapped/1000000) %>%
  mutate(uniquely_mapped_Million = uniquely_mapped/1000000)

# cross-host mapping rates (to confirm samples were not misassigned to species)
cross_map <- bind_rows(
  read.delim("../../output_RNA/alignment_qc/POR_MCapV3/multiqc_data/multiqc_star.txt") %>% mutate(species="POR",genome="MCapV3"),
  read.delim("../../output_RNA/alignment_qc/POR_PacutaV2/multiqc_data/multiqc_star.txt") %>% mutate(species="POR",genome="PacutaV2"),
  read.delim("../../output_RNA/alignment_qc/MON_Pcomp/multiqc_data/multiqc_star.txt") %>% mutate(species="MON",genome="Pcomp"),
  read.delim("../../output_RNA/alignment_qc/MON_PacutaV2/multiqc_data/multiqc_star.txt") %>% mutate(species="MON",genome="PacutaV2"),
  read.delim("../../output_RNA/alignment_qc/POC_Pcomp/multiqc_data/multiqc_star.txt") %>% mutate(species="POC",genome="Pcomp"),
  read.delim("../../output_RNA/alignment_qc/POC_MCapV3/multiqc_data/multiqc_star.txt") %>% mutate(species="POC",genome="MCapV3")
  ) %>%
  rename(total_read_pairs_mapped=total_reads) %>% mutate(total_reads_Million = total_read_pairs_mapped/1000000) %>%
  mutate(uniquely_mapped_Million = uniquely_mapped/1000000)

# symbiont mapping rates
sym_map <- bind_rows(
  read.delim("../../output_RNA/alignment_qc/POR_Cgoreaui_V2/multiqc_data/multiqc_star.txt") %>% mutate(species="POR",genome="Cgoreaui"),
  read.delim("../../output_RNA/alignment_qc/POR_Dtrenchii/multiqc_data/multiqc_star.txt") %>% mutate(species="POR",genome="Dtrenchii"),
  read.delim("../../output_RNA/alignment_qc/MON_Cgoreaui_V2/multiqc_data/multiqc_star.txt") %>% mutate(species="MON",genome="Cgoreaui"),
  read.delim("../../output_RNA/alignment_qc/MON_Dtrenchii/multiqc_data/multiqc_star.txt") %>% mutate(species="MON",genome="Dtrenchii"),
  read.delim("../../output_RNA/alignment_qc/POC_Cgoreaui_V2/multiqc_data/multiqc_star.txt") %>% mutate(species="POC",genome="Cgoreaui"),
  read.delim("../../output_RNA/alignment_qc/POC_Dtrenchii/multiqc_data/multiqc_star.txt") %>% mutate(species="POC",genome="Dtrenchii")
  ) %>%
  rename(total_read_pairs_mapped=total_reads) %>% mutate(total_reads_Million = total_read_pairs_mapped/1000000) %>%
  mutate(uniquely_mapped_Million = uniquely_mapped/1000000)

# kraken contamination screen
kraken <- read.csv("../../output_RNA/contam_screen/contamination_kraken.csv") %>% mutate(species = substr(sample,1,3))

# bbduk rRNA contamination screen from SILVA databases
rRNA <- read.csv("../../output_RNA/rRNA_screen/rRNA_contamination_bbduk_SILVA.csv")  %>%
            mutate(matched_reads_total = matched_reads_LSU + matched_reads_SSU,
                   percent_rRNA = (matched_reads_total/in_reads_LSU) * 100)

# combine all into one giant dataframe
qc_all <- host_map %>% select(Sample,species,genome, total_read_pairs_mapped, uniquely_mapped, uniquely_mapped_percent) %>%
  left_join(QC_raw, by = join_by(Sample)) %>%
  left_join(QC_trim, by = join_by(Sample)) %>%
  left_join(kraken %>% filter(classification=="Bacteria") %>% select(sample,percent_reads) %>% rename(Sample=sample, bacteria_percent=percent_reads), by = join_by(Sample)) %>%
    replace_na(list(bacteria_percent=0)) %>%
  left_join(rRNA %>% select(sample,percent_rRNA) %>% rename(Sample=sample), by = join_by(Sample))
```

## 1. Assess total number of reads per sample and whether we received the data we paid for (30M reads/sample)

### Raw sequences

```{r}
cutoff_M <- 30

qc_all %>%
  group_by(species) %>%
  summarise(
    n_samples = n(),
    min_reads_million = min(total_reads_raw/1000000),
    max_reads_million = max(total_reads_raw/1000000),
    .groups = "drop"
  ) %>%
  knitr::kable(format = "markdown", digits = 1)

low_samples <- qc_all %>% filter(total_reads_raw < cutoff_M*1e6) %>% select(Sample,species,total_reads_raw) %>% mutate(total_reads_million=total_reads_raw/1000000)

paste0(nrow(low_samples), " samples have fewer than ", cutoff_M/2, "M reads in each direction")
if(nrow(low_samples > 0)){print(low_samples)}
write.csv(QC_raw %>% mutate(total_reads_million=total_reads_raw/1000000) %>% arrange(total_reads_million), "../../output_RNA/raw_qc/multiqc_data_1/concatenated_runs_nreads.csv")
```

### After trimming and low-quality filtering

```{r}
cutoff_M <- 30

qc_all %>%
  group_by(species) %>%
  summarise(
    n_samples = n(),
    min_reads_million = min(total_reads_trim/1000000),
    max_reads_million = max(total_reads_trim/1000000),
    .groups = "drop"
  ) %>%
  knitr::kable(format = "markdown", digits = 1)

low_samples <- qc_all %>% filter(total_reads_trim < cutoff_M*1e6) %>% select(Sample,species,total_reads_trim) %>% mutate(total_reads_million=total_reads_trim/1000000)

paste0(nrow(low_samples), " samples have fewer than ", cutoff_M/2, "M reads in each direction")
if(nrow(low_samples > 0)){knitr::kable(low_samples,format = "markdown", digits = 1)}

write.csv(QC_trim %>% mutate(total_reads_million=total_reads_trim/1000000) %>% arrange(total_reads_million), "../../output_RNA/trimmed_qc/multiqc_data_1/concatenated_runs_nreads.csv")
```

## 2. Host: Assess mapping rates of each species to its respective host genome and screen for outliers/low mapping

```{r}
ggplot(qc_all, aes(x = reorder(Sample, uniquely_mapped_percent), 
                       y = uniquely_mapped_percent, fill = species)) + 
  geom_col(color="white") +
  facet_grid(~species, scales = "free_x") +
  labs(title = "Mapping Rate to Host Genome",
       y = "Unique Mapping %", x = NULL) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, size = 6)) +
  guides(fill = "none")
```

```{r}
host_mapping_outliers <- data.frame("Sample"=c(), "species"=c(), "genome"=c(), "uniquely_mapped_percent"=c(),"total_reads_Million"=c(),"uniquely_mapped_Million"=c())

for (sp in c("POC","MON","POR")){
  df <- host_map %>% filter(species ==sp)
  
  # IQR method 
  Q1 <- quantile(df$uniquely_mapped_percent, 0.25)
  Q3 <- quantile(df$uniquely_mapped_percent, 0.75)
  IQR_val <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  outliers <- df[df$uniquely_mapped_percent < lower_bound | df$uniquely_mapped_percent > upper_bound, ] %>% select(Sample, species,genome, uniquely_mapped_percent,total_reads_Million,uniquely_mapped_Million)
  print(paste0(sp,": IQR = ", Q1,"-",Q3,"; 1.5 x IQR = ",lower_bound,"-",upper_bound,"; ",nrow(outliers)," outliers"))
  
  host_mapping_outliers <- bind_rows(host_mapping_outliers,outliers)
}

if(nrow(host_mapping_outliers) > 0){knitr::kable(host_mapping_outliers,format = "markdown", digits = 1)}
```

## 3. Alt. Hosts: Assess mapping rates of each species to non-host genomes

If any of these samples show a higher mapping rate below than to their assigned genome above, there is an issue that could be due to sample mis-ID.

```{r}
cross_map %>% bind_rows(host_map) %>% filter(Sample %in% host_mapping_outliers$Sample) %>% select(Sample, genome, uniquely_mapped_percent) %>%
  pivot_wider(names_from = genome,values_from = uniquely_mapped_percent) %>% knitr::kable(format = "markdown", digits = 1)
```

## 4. Symbionts: Assess mapping rates of each species to symbiont genomes

```{r}
sym_map %>% group_by(species, genome) %>%
  summarize(range_percent=paste0(min(uniquely_mapped_percent),"-",max(uniquely_mapped_percent)),
            mean_percent=mean(uniquely_mapped_percent)) %>% knitr::kable(format = "markdown", digits = 1)

ggplot(sym_map, aes(x = reorder(Sample, desc(uniquely_mapped_percent)), y = uniquely_mapped_percent, fill = genome)) + 
  geom_col(color="white") +
  facet_grid(~species, scales = "free_x") +
  labs(title = "Mapping Rate to Cladocopium & Durusdinium Genomes (Could be overlapping reads)",
       y = "Unique Mapping %", x = NULL) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, size = 6))  + scale_fill_brewer(palette = "Set1")
```

```{r}
host_sym_compare <- host_map %>% select(Sample, species,uniquely_mapped_percent) %>% rename(host_mapping_rate=uniquely_mapped_percent) %>% left_join(sym_map %>% select(Sample, genome, uniquely_mapped_percent) %>% rename(sym_mapping_rate=uniquely_mapped_percent)) 

ggplot(host_sym_compare, aes(x=host_mapping_rate, y=sym_mapping_rate, color=genome)) + geom_point() +
  facet_grid(~species) +
  labs(title = "Mapping Rate to Cladocopium & Durusdinium Genomes vs. Host Genomes (Could be overlapping reads)",
       y = "Symbiont Mapping %", x = "Host Mapping %") +
  theme_bw() + scale_color_brewer(palette = "Set1")
```


```{r}
sym_mapping_outliers <- data.frame("Sample"=c(), "species"=c(), "genome"=c(), "uniquely_mapped_percent"=c(),"total_reads_Million"=c(),"uniquely_mapped_Million"=c())

for (sp in c("POC","MON","POR")){
  df_sp <- sym_map %>% filter(species ==sp)
  
  for (sym in c("Cgoreaui","Dtrenchii")){
  df <- df_sp %>% filter(genome ==sym)
  
  # IQR method 
  Q1 <- quantile(df$uniquely_mapped_percent, 0.25)
  Q3 <- quantile(df$uniquely_mapped_percent, 0.75)
  IQR_val <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  outliers <- df[df$uniquely_mapped_percent < lower_bound | df$uniquely_mapped_percent > upper_bound, ] %>% select(Sample, species,genome, uniquely_mapped_percent,total_reads_Million,uniquely_mapped_Million)
  print(paste0(sp,"-",sym,": IQR = ", Q1,"-",Q3,"; 1.5 x IQR = ",lower_bound,"-",upper_bound,"; ",nrow(outliers)," outliers"))
  
  sym_mapping_outliers <- bind_rows(sym_mapping_outliers,outliers)}
}

if(nrow(sym_mapping_outliers) > 0){knitr::kable(sym_mapping_outliers,format = "markdown", digits = 1)}
```

## 4. Kraken contamination and mapping comparison

```{r}
kraken %>% ggplot(aes(x=sample,y = percent_reads,fill=classification)) +geom_col() +
  facet_grid(~species,scales="free_x")  + theme_classic() + theme(axis.text.x = element_text(angle=90,size=6)) + scale_fill_brewer(palette = "Set2")

ggplot(qc_all, aes(x = uniquely_mapped_percent, y = bacteria_percent)) +
  geom_point(size = 2) +
  geom_smooth(method = "lm", se = FALSE, alpha = 0.3) +
  labs(title = "Host Mapping Rate vs. Bacterial Contamination",
       x = "Host Mapping %", y = "Bacterial Reads %") +
  theme_bw() +
  facet_grid(~species,scales="free_x")
```

```{r}
contam_outliers <- data.frame("sample"=c(), "species"=c(),"classification"=c(),"percent_reads"=c())

for (sp in c("POC","MON","POR")){
   df <- kraken %>% filter(species ==sp & classification=="Bacteria")
  
  # IQR method 
  Q1 <- quantile(df$percent_reads, 0.25)
  Q3 <- quantile(df$percent_reads, 0.75)
  IQR_val <- Q3 - Q1
  
  lower_bound <- Q1 - 1.5 * IQR_val
  upper_bound <- Q3 + 1.5 * IQR_val
  outliers <- df[df$percent_reads < lower_bound | df$percent_reads > upper_bound, ] %>% select(sample, species, classification, percent_reads)
  print(paste0(sp,": IQR = ", Q1,"-",Q3,"; 1.5 x IQR = ",lower_bound,"-",upper_bound,"; ",nrow(outliers)," outliers"))
  
  contam_outliers <- bind_rows(contam_outliers,outliers)
}

if(nrow(contam_outliers) > 0){knitr::kable(contam_outliers,format = "markdown", digits = 1)}
```

## 6. rRNA contamination and mapping comparison

```{r}
rRNA <- rRNA  %>%  left_join(host_map %>% select(Sample,total_read_pairs_mapped, uniquely_mapped, uniquely_mapped_percent,species), by = join_by(sample==Sample))

rRNA <- rRNA  %>% mutate(total_reads_Million = total_read_pairs_mapped/1000000) %>%
  mutate(matched_reads_pairs = matched_reads_total/2) %>%
  mutate(matched_reads_pairs_Million = matched_reads_pairs/1000000) %>%
  mutate(reads_non_rRNA_Million = total_reads_Million-matched_reads_pairs_Million) %>%
  mutate(uniquely_mapped_Million = uniquely_mapped/1000000)

plot <- qc_all %>% mutate(Raw_Reads_Million=total_reads_raw/1000000) %>%
  ggplot(aes(x = uniquely_mapped_percent, y = percent_rRNA)) +
  geom_smooth(method = "lm", se = FALSE, alpha = 0.3) +
  geom_point(size = 2,shape=21,aes(fill=Raw_Reads_Million)) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "Host Mapping Rate vs. rRNA Contamination",
       x = "Host Mapping %", y = "rRNA Reads %") +
  theme_bw() +
  facet_grid(~species,scales="free_x")
plot
ggsave(paste0(reportdir,"rRNA_v_mapping.png"))

qc_all %>% mutate(Raw_Reads_Million=total_reads_raw/1000000) %>%
  ggplot(aes(x = Raw_Reads_Million, y = percent_rRNA)) +
  geom_smooth(method = "lm", se = FALSE, alpha = 0.3) +
  geom_point(size = 2,shape=21,aes(fill=Raw_Reads_Million)) +
  scale_fill_distiller(palette = "Spectral") +
  labs(title = "Host Mapping Rate vs. rRNA Contamination",
       x = "# of Raw Reads (Million)", y = "rRNA Reads %") +
  theme_bw() +
  facet_grid(~species,scales="free_x")

rRNA_summary <- rRNA %>% select(sample,species,uniquely_mapped_percent,matched_reads_pairs_Million,total_reads_Million,percent_rRNA) %>% 
  dplyr::rename(rRNA_matched_reads_pairs_Million=matched_reads_pairs_Million,
                total_reads_pairs_Million = total_reads_Million
                ) %>% arrange(uniquely_mapped_percent) %>%
  mutate(across(contains("_"), ~ round(as.numeric(.x), 2)))

qc_all %>% pivot_longer(cols = contains("_"),
               names_to = "metric",
               values_to = "value") %>%
  ggplot(aes(x = value,color=species)) +
  geom_density() +
  facet_wrap(~ metric, scales = "free") +
  theme_bw() 
```


## 7. Compare to RNA Extraction QC metrics

```{r}
# Load and prepare extraction QC metadata
metadata <- read.csv("../../data_RNA/completed_bulk_RNA_extractions_3species.csv") %>%
  select(fragment_name, t_hours, treatment, Tank, Extraction_Date, 
         RNA_QBIT_AVG, RNA_Gel_quality_score, ng_RNA_total) %>%
  mutate(across(c(t_hours, treatment, Tank, Extraction_Date), factor),
         across(c(RNA_QBIT_AVG, RNA_Gel_quality_score, ng_RNA_total), as.numeric))

# Join read and extraction QC data
summary_meta <- qc_all %>% 
  left_join(metadata, by = join_by("Sample" == "fragment_name"))

# Visualize potential batch effects
library(patchwork)

for(sp in unique(qc_all$species)){
  df <- summary_meta %>% filter(species==sp)
  
# Continuous variables
cont_plots <- c("RNA_QBIT_AVG", "RNA_Gel_quality_score", "ng_RNA_total") %>%
  map(~ ggplot(df, aes(x = .data[[.x]], y = percent_rRNA)) +
        geom_point(alpha = 0.6) +
        geom_smooth(method = "lm", se = TRUE) +
        theme_bw() +
        labs(title = .x))

# Categorical variables (biological + batch)
cat_plots <- c("treatment", "t_hours", "Tank", "Extraction_Date") %>%
  map(~ ggplot(df, aes(x = .data[[.x]], y = percent_rRNA, color=treatment)) +
        geom_boxplot(outlier.shape = NA) +
        geom_jitter(alpha = 0.5, width = 0.15) +
        theme_bw() +
        labs(title = .x))

print(wrap_plots(c(cont_plots, cat_plots))+plot_annotation(title = paste0(sp," QC Metrics")))

# Model with treatment×time interaction
model <- lm(percent_rRNA ~ RNA_QBIT_AVG + RNA_Gel_quality_score + ng_RNA_total + 
              treatment * t_hours + Tank + Extraction_Date,
            data = df)
print(paste0("=======",sp," QC Metrics======="))
print(summary(model))
print(anova(model))

# Diagnostic plots
par(mfrow = c(2, 2))
plot(model)
}

# Examine batch effect (Extraction_Date)
summary_meta %>%
  group_by(Extraction_Date,species) %>%
  summarise(n = n(), 
            mean_rRNA = mean(percent_rRNA, na.rm = TRUE),
            sd_rRNA = sd(percent_rRNA, na.rm = TRUE)) %>%
  arrange(desc(mean_rRNA))
```

## 8. POR-only rRNA contamination and mapping comparison

Note: the bbduk reads are unpaired, so the bbduk total reads is 2x total reads in the mapping QC analysis

```{r}
# load in multiqc data of mapping data
POR_map <- host_map %>% filter(species == "POR")
POR_rRNA <- read.csv("../../output_RNA/rRNA_screen/rRNA_contamination_bbduk_POR.csv") %>% rename(bbduk_total_reads=total_reads)
POR_rRNA_Fede <- read.csv("../../output_RNA/rRNA_screen/rRNA_contamination_bbduk_POR_Federica.csv") %>% select(-total_reads)
```

```{r}
POR_rRNA <- POR_rRNA %>% left_join(POR_map %>% select(Sample,total_read_pairs_mapped, uniquely_mapped, uniquely_mapped_percent), by = join_by(sample==Sample))

POR_rRNA <- POR_rRNA %>% mutate(total_reads_Million = total_read_pairs_mapped/1000000) %>%
  mutate(matched_reads_pairs = matched_reads/2) %>%
  mutate(matched_reads_pairs_Million = matched_reads_pairs/1000000) %>%
  mutate(reads_non_rRNA_Million = total_reads_Million-matched_reads_pairs_Million) %>%
  mutate(uniquely_mapped_Million = uniquely_mapped/1000000)


plot <- ggplot(POR_rRNA) + 
  geom_point(aes(x = percent_rrna, y =uniquely_mapped_percent, color=total_reads_Million)) + theme_bw() + 
  labs(title = "Percent rRNA reads vs. Unique Mapping Rates:\nPOR Samples")
plot
ggsave(paste0(reportdir,"POR_rRNA_v_mapping.png"))

ggplot(POR_rRNA) + 
  geom_point(aes(x = percent_rrna, y =uniquely_mapped_percent, color=total_reads_Million)) + theme_bw() + 
  labs(title = "Percent rRNA reads vs. Unique Mapping Rates:\nPOR Samples")

ggplot(POR_rRNA) + 
  geom_point(aes(x = percent_rrna, y =total_reads_Million, color=total_reads_Million)) + theme_bw() + 
  labs(title = "Percent rRNA reads vs. Unique Mapping Rates:\nPOR Samples")


POR_rRNA_summary <- POR_rRNA %>% select(sample,uniquely_mapped_percent,matched_reads_pairs_Million,total_reads_Million,percent_rrna) %>% 
  dplyr::rename(rRNA_matched_reads_pairs_Million=matched_reads_pairs_Million,
                total_reads_pairs_Million = total_reads_Million
                ) %>% arrange(uniquely_mapped_percent) %>%
  mutate(across(contains("_"), ~ round(as.numeric(.x), 2)))


POR_rRNA_summary %>% pivot_longer(cols = contains("_"),
               names_to = "metric",
               values_to = "value") %>%
  ggplot(aes(x = value)) +
  geom_density(fill="lightgray") +
  facet_wrap(~ metric, scales = "free") +
  theme_bw() 


write_csv(POR_rRNA_summary, "../../output_RNA/rRNA_screen/POR_Pcomp_rRNA_and_mapping_metadata.csv")
```

## 9. POR mRNA-decontaminated rarefaction analysis


```{r}
# Directory containing all rarefaction CSV outputs
resdir <- "/scratch4/workspace/zdellaert_uri_edu-shared_TimeSeries/TimeSeries/rRNA_rarefaction/"

# Load all CSVs
csv_files <- list.files(resdir, pattern = "_rarefaction.csv$", recursive = TRUE, full.names = TRUE)

all_results <- map_df(csv_files, function(f) {
  # Extract sample name from parent directory
  sample_name <- basename(dirname(f))
  
  read_csv(f) %>%
    mutate(sample = sample_name)
})

head(all_results)
```

```{r}
all_results <- all_results %>%
  mutate(
    depth = as.numeric(depth),
    pairs_processed = as.numeric(pairs_processed),
    pairs_mapped = as.numeric(pairs_mapped),
    percent_mapped = as.numeric(percent_mapped),
    percent_unique = as.numeric(percent_unique),
    genes_detected = as.numeric(genes_detected),
  )
```

```{r}
# Genes detected
ggplot(all_results, aes(x = depth, y = pairs_processed, color = sample)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Reads processed at each subsampling depth",
    x = "Fraction of original reads used",
    y = "Reads processed"
  ) +
  theme_bw()

# Genes detected
ggplot(all_results, aes(x = depth, y = pairs_mapped, color = sample)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Reads pseudoaligned by subsampling depth",
    x = "Fraction of original reads used",
    y = "Reads pseudoaligned"
  ) +
  theme_bw()

# Genes detected
ggplot(all_results, aes(x = depth, y = genes_detected, color = sample)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Rarefaction of detected genes by subsampling depth",
    x = "Fraction of original reads used",
    y = "Number of genes detected (TPM > 0.5)"
  ) +
  theme_bw()

# Mapping rate
ggplot(all_results, aes(x = depth, y = percent_mapped, color = sample)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Rarefaction of pseudoalignment (mapping) rate by subsampling depth",
    x = "Fraction of original reads used",
    y = "Pseudoalignment (mapping) rate"
  ) +
  theme_bw()


# Mapping rate
ggplot(all_results, aes(x = depth, y = percent_unique, color = sample)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(labels = scales::percent_format()) +
  labs(
    title = "Rarefaction of unique pseudoalignment (mapping) rate by subsampling depth",
    x = "Fraction of original reads used",
    y = "Unique Pseudoalignment (mapping) rate"
  ) +
  theme_bw()

```
```{r}
plateau <- all_results %>%
  group_by(sample) %>%
  arrange(depth) %>%
  mutate(delta = genes_detected - lag(genes_detected))# %>%
  #filter(!is.na(delta) & delta < 50)  # threshold for minimal new genes

head(plateau)
```

```{r}
POR_combined <- all_results %>% left_join(POR_rRNA_summary)

# Genes detected
ggplot(POR_combined, aes(x = depth, y = genes_detected, group=sample, color = uniquely_mapped_percent)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_distiller(palette = "Spectral") +
  labs(
    title = "Rarefaction of detected genes by subsampling depth",
    x = "Fraction of original reads used",
    y = "Number of genes detected (TPM > 0.5)"
  ) +
  theme_bw()
```

```{r}
POR_rRNA <- read.csv("../../output_RNA/rRNA_screen/rRNA_contamination_bbduk_POR.csv")  %>% 
  mutate(total_reads_M_correct = total_reads/1000000)
POR_rRNA_Fede <- read.csv("../../output_RNA/rRNA_screen/rRNA_contamination_bbduk_POR_Federica.csv")  %>% 
  mutate(total_reads_M_correct = total_reads/1000000)

POR_rRNA$dataset <- "TimeSeries"
POR_rRNA_Fede$dataset <- "TPC"
combined <- rbind(POR_rRNA,POR_rRNA_Fede)

ggplot(combined, aes(x=percent_rrna, fill = dataset)) +
         geom_histogram(alpha=0.5, binwidth=1) + 
        coord_cartesian(xlim = c(0, 100)) + theme_bw()

library(cowplot) 
pmain <- ggplot(combined, aes(x=percent_rrna,y=total_reads_M_correct, color = dataset))+
  geom_point()

xdens <- axis_canvas(pmain, axis = "x")+
  geom_density(data = combined, aes(x = percent_rrna, fill = dataset), alpha = 0.7, size=0.2)

ydens <- axis_canvas(pmain, axis = "y", coord_flip = TRUE)+
  geom_density(data = combined, aes(x = total_reads_M_correct, fill = dataset), alpha = 0.7, size=0.2)+
  coord_flip()
p1 <- insert_xaxis_grob(pmain, xdens, grid::unit(.2, "null"), position = "top")
p2 <- insert_yaxis_grob(p1, ydens, grid::unit(.2, "null"), position = "right")
ggdraw(p2)
```



