---
title: "RNA-seq Preprocessing and Normalization"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
params:
  species: "Pacuta" # default for interactive use
output:
  github_document:
    toc: true
    html_preview: false
---

# Preproccessing of bulk RNA-seq data

```{r setup}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath(paste0("../../output_RNA/reports/", params$species, "/")), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,fig.width = 10, fig.height = 8,
                      fig.path = "01_preprocessing_files/figure-gfm/")

#load packages
library(tidyverse)
library(DESeq2)
library(pheatmap)
library(RColorBrewer)
library(genefilter)
library(ggnewscale)
library(BiocParallel)

#load in parameters and functions
source("species_parameters.R")
source("utils.R")

# set number of cores to use for parallel DESeq2 processing
register(MulticoreParam(workers = global_params$n_cores))

sessionInfo() #provides list of loaded packages and version of R
```

## 0. Setup species-specific parameters

```{r get-species}
# get species
species <- params$species

# get parameters for this species
config <- get_params(species)
print_config(species)

# set up necessary output directories if they don't exist
outdir <- file.path("../../output_RNA/counts_filt_norm", species)
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)

reportdir <- file.path("../../output_RNA/reports", params$species, "01_preprocessing_files/figure-gfm/")
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)
```

## 1. Read in raw count data

```{r load-counts}
# load in data
counts_raw <- read.csv(file.path("../../output_RNA/count_matrices", config$count_matrix), row.names = 1)

# make list of samples 
samples <- colnames(counts_raw)
cat("Raw counts:", nrow(counts_raw), "genes x", ncol(counts_raw), "samples")

# read in SwissProt annotation
SwissProt <- read.delim(file.path(annot_dir,config$SwissProt))
cat("Annotations:", nrow(SwissProt), "Swissprot-annotated genes")
```

## 2. Extract metadata from sample names

```{r metadata}
# create metadata dataframe from sample names
meta <- data.frame(
  sample = samples, 
  species = str_split(samples, "_", simplify = TRUE)[,1], #extract first part of sample name to get species
  time = str_replace(str_split(samples, "_", simplify = TRUE)[,2],"R", ""), #extract "R##" part to get timepoint then remove R
  replicate = str_split(samples, "_", simplify = TRUE)[,3], #extract "R##" part to get timepoint then remove R
  treatment = str_replace(str_split(samples, "_", simplify = TRUE)[,3],"\\d", "")
)

# add rownames
rownames(meta) <- meta$sample

# make time and treatment factors
meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
meta$treatment <- factor(meta$treatment)

# save metadata
meta <- meta %>% arrange(time, treatment)
write.csv(meta, paste0("../../output_RNA/",species,"_RNA_seq_metadata.csv"))
cat("Metadata file saved to:", paste0("../../output_RNA/",species,"_RNA_seq_metadata.csv"))

# reorder count matrix to be in order of metadata table (should be already but just in case)
counts_raw <- counts_raw[, meta$sample]
```

## 3. Remove outliers, if identified

```{r outliers}
outlier_samples <- config$outlier_samples

if(length(outlier_samples) > 0) {
    counts_raw <- counts_raw[, !colnames(counts_raw) %in% outlier_samples]
    meta <- meta[!rownames(meta) %in% outlier_samples, ]
}

#Confirm that sample names in metadata and count matrix match and are in the same order
stopifnot(all(meta$sample %in% colnames(counts_raw))) #are all of the sample names in the metadata column names in the gene count matrix?
stopifnot(all(meta$sample == colnames(counts_raw))) #are they the same in the same order?
```

## 4. pOverA filtering to reduce dataset

### Note to self: maybe replace this with treatment-specific filtering. To get genes expressed only at one timepoint in one treatment

```{r filtering}
# Keep genes expressed at 10+ counts in at least 7% of samples - expressed in all 3 samples at one timepoint from one treatment, can change parameters in species_parameters.R script

ffun<-filterfun(pOverA(global_params$pOverA_proportion,global_params$pOverA_counts))
counts_filt_poa <- genefilter((counts_raw), ffun) #apply filter

filtered_counts <- counts_raw[counts_filt_poa,] #keep only rows that passed filter

paste0("Number of genes after filtering: ", sum(counts_filt_poa))
paste0("% of genes kept: ", round(100*(sum(counts_filt_poa)/nrow(counts_raw)),digits=2),"%")

write.csv(filtered_counts, file = file.path(outdir, "filtered_counts.csv"))
cat("Filtered counts saved to:", file.path(outdir, "filtered_counts.csv"))
```

## 5. Create DESeq object and run DESeq2

```{r deseq}
dds <- DESeqDataSetFromMatrix(countData = filtered_counts,
                              colData = meta,
                              design= ~ treatment + time + treatment:time)

dds <- DESeq(dds, parallel = TRUE)

# Estimate size factors to determine if we can use VST
SF.dds <- estimateSizeFactors(dds) 
print(sort(sizeFactors(SF.dds))) #View size factors

# if all are less than 4 we can use the VST transformation
all(sizeFactors(SF.dds)) < 4
```

## 6. VST-Transforming count data for visualization 

```{r vst-transform}
vsd <- vst(dds, blind=FALSE)

#save the vsd transformation
vsd_mat <- assay(vsd)
write.csv(vsd_mat, file = file.path(outdir, "vsd_expression_matrix.csv"))
cat("VST matrix saved to:", file.path(outdir, "vsd_expression_matrix.csv"))
```

## 7. Two tools to identiy potential outliers:

### PCA

```{r pca}
pcaData <- plotPCA(vsd, intgroup=c("time", "treatment"), returnData=TRUE)
percentVar <- round(100 * attr(pcaData, "percentVar"))

PCA <- ggplot() +
  geom_point(data = subset(pcaData, treatment == "C"),
             aes(x=PC1, y=PC2, color=time),
                 size=3) +
             scale_color_manual(values=brewer.pal(7, "Blues"), name = "Time (hrs) - Control") +
  
  #start new scale
  ggnewscale::new_scale_color() +
  geom_point(data = subset(pcaData, treatment == "H"),
             aes(x=PC1, y=PC2, color=time),
                 size=3) +
             scale_color_manual(values=brewer.pal(7, "Oranges"), name = "Time (hrs) - Heat") +

  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() + theme_bw() + ggtitle(paste(species, "- PCA of VST-transformed counts"))

print(PCA)
save_ggplot(PCA, "PCA")

PCA_simple <- ggplot(data = pcaData, aes(x=PC1, y=PC2, color=treatment, shape=time)) +
  geom_point(size=4) +
  scale_color_manual(values= c("C"= "#4292C6", "H" = "#D94801"), labels = c("Control", "Heat")) +
  scale_shape_manual(values = c(16, 17, 15, 18, 0, 1, 2)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  labs(color = "Treatment", shape = "Time (h)") +
  coord_fixed() + theme_bw() + ggtitle(paste(species, "- PCA of VST-transformed counts"))

print(PCA_simple)
save_ggplot(PCA_simple, "PCA_simple", width = 8, height = 6)
```

### Hierarchical Clustering

```{r cluster}
sampleTree <- hclust(dist(t(vsd_mat)), method = "average")

par(mar = c(8, 4, 2, 2))
plot(sampleTree, 
     xlab = "", sub = "", cex = 0.7)
abline(h = 100, col = "red", lty = 2)
```

### Note: If outliers are identified, add them to species_parameters.R for this species.

## Final summary

### Heatmap of variable genes

```{r}
topVarGenes <- head(order(rowVars(vsd_mat), decreasing=TRUE), 500)

pheatmap(vsd_mat[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=FALSE, 
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))

pheatmap(vsd_mat[topVarGenes, ], cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, cutree_cols = 2,
         annotation_col= meta[,c("treatment","time")],
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors))
```

### Text summary

```{r final, echo=FALSE}
cat("Preprocessing Summary:", species, "\n")

cat("Input\n")
cat(strrep("-", 40), "\n")
cat("  Count matrix:", config$count_matrix, "\n")
cat("  Initial genes:", nrow(counts_raw), "\n")
cat("  Initial samples:", ncol(counts_raw), "\n\n")

cat("Filtering\n")
cat(strrep("-", 40), "\n")
cat("  Outliers removed:", length(config$outlier_samples), "\n")
if (length(config$outlier_samples) > 0) {
  cat("    ", paste(config$outlier_samples, collapse = ", "), "\n")
}
cat("  Low-expression genes removed:", sum(!counts_filt_poa), "\n")
cat("  pOverA filter: >=", global_params$pOverA_counts, "counts in >=", 
    global_params$pOverA_proportion * 100, "% of samples\n\n")

cat("Output\n")
cat(strrep("-", 40), "\n")
cat("  Final genes:", nrow(vsd_mat), "\n")
cat("  Final samples:", ncol(vsd_mat), "\n")
cat("  Output directory:", outdir, "\n\n")

cat("QC Notes\n")
cat(strrep("-", 40), "\n")
cat("  Size factors range:", round(min(sizeFactors(SF.dds)), 2), "-", round(max(sizeFactors(SF.dds)), 2), "\n")
cat("  VST appropriate:", ifelse(all(sizeFactors(SF.dds) < 4), "Yes", "No - consider rlog"), "\n")
cat("  PC1 variance:", percentVar[1], "%\n")
cat("  PC2 variance:", percentVar[2], "%\n")
```

```{r}
sessionInfo()
```

