---
title: "ImpulseDE2 Temporal Analysis"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
params:
  species: "Pacuta" # default for interactive use
  run_ImpulseDE2: FALSE # default for interactive use
output:
  github_document:
    toc: true
    html_preview: false
---

# Bulk RNA-seq Time Course Trajectory Analysis and Clustering

## Introduction to packages

### ImpulseDE2

 - Based on [this paper](https://academic.oup.com/bib/article/20/1/288/4364840#130283262), this is the best package to use other than comparing each time point against each other individually.
 - Repo here: https://github.com/YosefLab/ImpulseDE2
 - Tutorial here: http://bioconductor.statistik.tu-dortmund.de/packages/3.11/bioc/vignettes/ImpulseDE2/inst/doc/ImpulseDE2_Tutorial.html , I followed closely with the section "Case-control differential expression analysis"
- Read the ImpulseDE2 paper [here](https://academic.oup.com/nar/article/46/20/e119/5068248)

*David S Fischer, Fabian J Theis, Nir Yosef, Impulse model-based differential expression analysis of time course sequencing data, Nucleic Acids Research, Volume 46, Issue 20, 16 November 2018, Page e119, https://doi.org/10.1093/nar/gky675*

To install the package

```{r, eval=FALSE}
library(devtools)
install_github("YosefLab/ImpulseDE2")
```

### Mfuzz: clustering of temporal trajectories

For this we will use the package [Mfuzz](https://bioconductor.org/packages/release/bioc/html/Mfuzz.html), [vignette here](https://bioconductor.org/packages/release/bioc/vignettes/Mfuzz/inst/doc/Mfuzz.pdf) and more documentation [here](http://mfuzz.sysbiolab.eu/).

To install the package

```{r, eval=FALSE}
BiocManager::install("Mfuzz")
```

---

# Setup

## 1. Load packages and functions

```{r setup}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath(paste0("../../output_RNA/reports/", params$species, "/")), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,fig.width = 10, fig.height = 8,
                      fig.path = "02_ImpulseDE_files/figure-gfm/")

#load packages
library(ImpulseDE2)
library(tidyverse)
library(ComplexHeatmap)
library(pheatmap)
library(Mfuzz)

#load in parameters and functions
source("species_parameters.R")
source("utils.R")

sessionInfo() #provides list of loaded packages and version of R
```

## 2. Setup species-specific parameters

```{r get-species}
# get species
species <- params$species

# get parameters for this species
config <- get_params(species)
print_config(species)

# define preprocessing output directory (from 01_preprocessing.Rmd)
input_dir <- file.path("../../output_RNA/counts_filt_norm", species)

# set up necessary output directories if they don't exist
outdir <- file.path("../../output_RNA/ImpulseDE2", species)
outdir_mfuzz <- file.path(outdir,"Mfuzz")
outdir_plots <- file.path(outdir,"plots")
outdir_plots_pdf <- file.path(outdir_plots,"pdf_figs")
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
if (!dir.exists(outdir_mfuzz)) dir.create(outdir_mfuzz, recursive = TRUE)
if (!dir.exists(outdir_plots)) dir.create(outdir_plots, recursive = TRUE)
if (!dir.exists(outdir_plots_pdf)) dir.create(outdir_plots_pdf, recursive = TRUE)

reportdir <- file.path("../../output_RNA/reports", params$species, "02_ImpulseDE_files/figure-gfm/")
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)
```

## 3. Load in filtered raw counts, transformed counts, metadata, and annotations

```{r load-counts}
# load in filtered counts data
filtered_counts <- read.csv(file.path(input_dir, "filtered_counts.csv"), row.names = 1)

# load in vst-transformed counts
vst <- read.csv(file.path(input_dir, "vst_expression_matrix.csv"))
vst <- vst %>% column_to_rownames(var = "X")

# load in metadata
meta <- read.csv(paste0("../../output_RNA/",species,"_RNA_seq_metadata.csv"))
meta <- meta %>% column_to_rownames(var = "X") #%>% select(-c(species, replicate))

# make time and treatment factors
meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
meta$treatment <- factor(meta$treatment)

# remove outliers that are still in metadata and raw_counts files but were removed prior to the vst transformation
outlier_samples <- config$outlier_samples

if(length(outlier_samples) > 0) {
    filtered_counts <- filtered_counts[, !colnames(filtered_counts) %in% outlier_samples]
    meta <- meta[!rownames(meta) %in% outlier_samples, ]
}

all(rownames(meta) %in% colnames(filtered_counts))
all(rownames(meta) == colnames(filtered_counts))

all(rownames(meta) %in% colnames(vst))
all(rownames(meta) == colnames(vst))

# once outliers are removed, define extra sample lists
heat_samples <- meta %>% filter(treatment == "H") %>% pull(sample)
```

```{r}
# read in SwissProt annotation
SwissProt <- read.delim(file.path(annot_dir,config$SwissProt))
cat("Annotations:", nrow(SwissProt), "Swissprot-annotated genes")
```

---

# ImpulseDE2 Analysis

## 1. Metadata formatting

First, reformat our metadata table to match the column names used in the ImpulseDE2 vignette.

```{r}
meta_impulse <- meta %>%
  dplyr::rename(Sample = sample, Time = time, Batch = replicate) %>% 
  mutate(Time = as.numeric(as.character(Time)),
         Condition = str_replace(treatment, "C", "control"),
         Condition = str_replace(Condition, "H", "case")
         ) %>%
  select(-c(species,treatment))
```

## 2. Run ImpulseDE2

This takes a ton of time and memory, so I run it once then save as an RDS.

```{r} 
if(params$run_ImpulseDE2 == TRUE) {
  objectImpulseDE2 <- runImpulseDE2(
    matCountData    = as.matrix(filtered_counts), #or use filtered_counts  
    dfAnnotation    = meta_impulse,
    boolCaseCtrl    = TRUE,
    vecConfounders  = c("Batch"), #only use if you want to try to control for batch effects
    boolIdentifyTransients = TRUE, #use if you want to ID transiently- vs permanently-regulated genes
    scaNProc        = 18 )
  
  saveRDS(objectImpulseDE2, file.path(outdir, "objectImpulseDE2.rds"))
} else {
  objectImpulseDE2 <- readRDS(file.path(outdir, "objectImpulseDE2.rds"))
}
```

```{r}
# Print processing report
cat(objectImpulseDE2@strReport)
```

## 3. Extract ImpulseDE2 results

### All genes

Extract and save results for all non-zero genes

```{r}
impulse_results <- objectImpulseDE2$dfImpulseDE2Results
impulse_results <- impulse_results %>% filter(allZero==FALSE) #remove genes with zero counts

# classify genes as transiently or monotonously regulated if impulseDE gave them one of those labels
impulse_results <- impulse_results %>%
  mutate(response_type = case_when(
    isTransient & !is.na(isTransient) ~ "Transient",
    isMonotonous & !is.na(isMonotonous) ~ "Monotonous",
    .default = "Other"
  ))

impulse_results_annot <- impulse_results %>%
  left_join(SwissProt %>% select(query,ProteinNames,BiologicalProcess), by = join_by("Gene"=="query"))
```

### Significant genes 

Extract genes with significant treatment effect on temporal trajectory and save results

```{r}
impulse_sig <- impulse_results %>%
  filter(padj < global_params$padj_threshold)

#preview top DE genes and annotations
impulse_sig %>% arrange(padj) %>% head(20) %>% dplyr::select(Gene,padj,loglik_red,response_type) %>%
  left_join(SwissProt %>% select(query,ProteinNames,BiologicalProcess), by = join_by("Gene"=="query"))
```

### Quick summary

```{r echo=FALSE}
cat("Total significant genes:", nrow(impulse_sig), "\n")
cat("Response patterns:\n")
cat("Transient:", sum(impulse_sig$response_type == "Transient"), "\n")
cat("Monotonous:", sum(impulse_sig$response_type == "Monotonous"), "\n")
cat("Other:", sum(impulse_sig$response_type == "Other"), "\n")
```

## 4. Visualize ImpulseDE2 Results

### Classify significant genes by up or downregulation

```{r}
lsHeatmaps <- plotHeatmap(
  objectImpulseDE2       = objectImpulseDE2,
  strCondition           = "case",
  boolIdentifyTransients = TRUE, #set to true if true above
  scaQThres              = global_params$padj_threshold)

# this code is also handy because the object it creates sorts signficiant genes into the following four categories:
transition_up <- lsHeatmaps$lsvecGeneGroups$transition_up
transition_down <- lsHeatmaps$lsvecGeneGroups$transition_down
transient_up <- lsHeatmaps$lsvecGeneGroups$transient_up
transient_down <- lsHeatmaps$lsvecGeneGroups$transient_down
classified <- c(transition_up,transition_down,transient_up,transient_down)
# the unclassified ones below are the same as response_type=="Other"
unclassified <- impulse_sig %>% filter(!(Gene %in% classified)) %>% pull(Gene)

gene_classification <- data.frame(
  Gene = c(transition_up, transition_down, transient_up, transient_down, unclassified),
  classification = c(
    rep("transition_up", length(transition_up)),
    rep("transition_down", length(transition_down)),
    rep("transient_up", length(transient_up)),
    rep("transient_down", length(transient_down)),
    rep("unclassified", length(unclassified))
  )
)


impulse_classifications <- impulse_results %>% left_join(gene_classification)
write.csv(impulse_classifications, file.path(outdir, "ImpulseDE2_results.csv"), row.names = FALSE)

impulse_sig_classifications <- impulse_sig %>% left_join(gene_classification)
write.csv(impulse_sig_classifications, file.path(outdir, "ImpulseDE2_significant.csv"), row.names = FALSE)
```

### Heatmap of significant genes by trajectory

```{r}
# complexHeatmapRaw = Heatmap of raw data by time point: Average of the size factor (and batch factor) normalised counts per time point and gene.
draw(lsHeatmaps$complexHeatmapRaw)

# complexHeatmapFit = Heatmap of impulse-fitted data by time point.
draw(lsHeatmaps$complexHeatmapFit)

png(file.path(outdir_plots,"ImpulseDE2_heatmap_case_fit.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapFit)
dev.off()

pdf(file.path(outdir_plots,"pdf_figs/ImpulseDE2_heatmap_case_fit.pdf"), width = 10, height = 12)
draw(lsHeatmaps$complexHeatmapFit)
dev.off()

png(file.path(outdir_plots,"ImpulseDE2_heatmap_case.png"), width = 2000, height = 2400, res = 300)
draw(lsHeatmaps$complexHeatmapRaw)
dev.off()

pdf(file.path(outdir_plots, "pdf_figs/ImpulseDE2_heatmap_case.pdf"), width = 10, height = 12)
draw(lsHeatmaps$complexHeatmapRaw)
dev.off()
```

### Heatmap of top 100 significant transient/transition UP genes normalized to T0

```{r}
impulse_sig_up <- impulse_sig %>% filter(Gene %in% transient_up | Gene %in% transition_up)
top_100_DE_genes <- impulse_sig_up %>% arrange(padj) %>% filter(Gene %in% rownames(vst)) %>% head(100) %>% arrange(response_type,padj)
row_annot <- top_100_DE_genes %>% select(response_type)
#col_annot <- meta[heat_samples,] %>% select(treatment,time)
col_annot <- meta[heat_samples,] %>% select(time)

T0_samples <- meta[heat_samples,] %>% filter(time == 0 ) %>% pull(sample)

vst_heat_top <- vst[top_100_DE_genes$Gene,heat_samples]
vst_heat_top$T0mean <- rowMeans(vst_heat_top[,T0_samples])
vst_heat_norm <- vst_heat_top - vst_heat_top$T0mean
vst_heat_norm <- vst_heat_norm %>% select(-T0mean)

vst_max <- max(abs(vst_heat_norm), na.rm = TRUE)
scale_colors <- colorRampPalette(c("#3c58a7","#3c58a7", "white", "#ee2e25","#ee2e25"))(100)
legend_breaks <- seq(-vst_max, vst_max, length.out = 101)

heatmap_up <- pheatmap(vst_heat_norm, 
         cluster_rows = FALSE, 
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE, 
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         breaks = legend_breaks,
         col = scale_colors,
         border_color = NA,
         gaps_col = cumsum(table(col_annot$time))[-length(table(col_annot$time))],
         annotation_row = row_annot,
         annotation_col = col_annot,
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors,
                                  "response_type" = c("Monotonous" = "#D8AF39FF", "Transient" = "#278B9AFF")))

png(file.path(outdir_plots,"ImpulseDE2_heatmap_UP.png"),  width = 7, height = 10, units = "in", res = 300)
heatmap_up
dev.off()

pdf(paste0(outdir_plots,"/pdf_figs/ImpulseDE2_heatmap_UP.pdf"), width = 7, height = 10)
heatmap_up
dev.off()

all_DE_genes <- impulse_sig_up %>% arrange(padj) %>% filter(Gene %in% rownames(vst)) %>% arrange(response_type,padj)
row_annot <- all_DE_genes %>% select(response_type)
#col_annot <- meta[heat_samples,] %>% select(treatment,time)
col_annot <- meta[heat_samples,] %>% select(time)

T0_samples <- meta[heat_samples,] %>% filter(time == 0 ) %>% pull(sample)

vst_heat_top <- vst[all_DE_genes$Gene,heat_samples]
vst_heat_top$T0mean <- rowMeans(vst_heat_top[,T0_samples])
vst_heat_norm <- vst_heat_top - vst_heat_top$T0mean
vst_heat_norm <- vst_heat_norm %>% select(-T0mean)

vst_max <- max(abs(vst_heat_norm), na.rm = TRUE)
scale_colors <- colorRampPalette(c("#3c58a7","#3c58a7", "white", "#ee2e25","#ee2e25"))(100)
legend_breaks <- seq(-vst_max*.9, vst_max*.9, length.out = 101)

heatmap_up_all <- pheatmap(vst_heat_norm, 
         cluster_rows = FALSE, 
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE, 
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         breaks = legend_breaks,
         col = scale_colors,
         border_color = NA,
         gaps_col = cumsum(table(col_annot$time))[-length(table(col_annot$time))],
         annotation_row = row_annot,
         annotation_col = col_annot,
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors,
                                  "response_type" = c("Monotonous" = "#D8AF39FF", "Transient" = "#278B9AFF")))

png(file.path(outdir_plots,"ImpulseDE2_heatmap_UP_ALL.png"),  width = 7, height = 10, units = "in", res = 300)
heatmap_up_all
dev.off()

pdf(paste0(outdir_plots,"/pdf_figs/ImpulseDE2_heatmap_UP_ALL.pdf"), width = 7, height = 10)
heatmap_up_all
dev.off()
```

### Heatmap of top 100 significant transient/transition downregulated genes normalized to T0

```{r}
impulse_sig_down <- impulse_sig %>% filter(Gene %in% transient_down | Gene %in% transition_down)
top_100_DE_genes <- impulse_sig_down %>% arrange(padj) %>% filter(Gene %in% rownames(vst)) %>% head(100) %>% arrange(response_type,padj)
row_annot <- top_100_DE_genes %>% select(response_type)
#col_annot <- meta[heat_samples,] %>% select(treatment,time)
col_annot <- meta[heat_samples,] %>% select(time)

vst_heat_top <- vst[top_100_DE_genes$Gene,heat_samples]
vst_heat_top$T0mean <- rowMeans(vst_heat_top[,T0_samples])
vst_heat_norm <- vst_heat_top - vst_heat_top$T0mean
vst_heat_norm <- vst_heat_norm %>% select(-T0mean)

vst_max <- max(abs(vst_heat_norm), na.rm = TRUE)
scale_colors <- colorRampPalette(c("#3c58a7","#3c58a7", "white", "#ee2e25","#ee2e25"))(100)
legend_breaks <- seq(-vst_max, vst_max, length.out = 101)

heatmap_down <- pheatmap(vst_heat_norm, 
         cluster_rows = FALSE, 
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE, 
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         breaks = legend_breaks,
         col = scale_colors,
         border_color = NA,
         gaps_col = cumsum(table(col_annot$time))[-length(table(col_annot$time))],
         annotation_row = row_annot,
         annotation_col = col_annot,
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors,
                                  "response_type" = c("Monotonous" = "#D8AF39FF", "Transient" = "#278B9AFF")))

png(file.path(outdir_plots,"ImpulseDE2_heatmap_DOWN.png"),  width = 7, height = 10, units = "in", res = 300)
heatmap_down
dev.off()

pdf(paste0(outdir_plots,"/pdf_figs/ImpulseDE2_heatmap_DOWN.pdf"), width = 7, height = 10)
heatmap_down
dev.off()

all_DE_genes <- impulse_sig_down %>% arrange(padj) %>% filter(Gene %in% rownames(vst)) %>% arrange(response_type,padj)
row_annot <- all_DE_genes %>% select(response_type)
#col_annot <- meta[heat_samples,] %>% select(treatment,time)
col_annot <- meta[heat_samples,] %>% select(time)

T0_samples <- meta[heat_samples,] %>% filter(time == 0 ) %>% pull(sample)

vst_heat_top <- vst[all_DE_genes$Gene,heat_samples]
vst_heat_top$T0mean <- rowMeans(vst_heat_top[,T0_samples])
vst_heat_norm <- vst_heat_top - vst_heat_top$T0mean
vst_heat_norm <- vst_heat_norm %>% select(-T0mean)

vst_max <- max(abs(vst_heat_norm), na.rm = TRUE)
scale_colors <- colorRampPalette(c("#3c58a7","#3c58a7", "white", "#ee2e25","#ee2e25"))(100)
legend_breaks <- seq(-vst_max*.9, vst_max*.9, length.out = 101)

heatmap_down_all <- pheatmap(vst_heat_norm, 
         cluster_rows = FALSE, 
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE, 
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         breaks = legend_breaks,
         col = scale_colors,
         border_color = NA,
         gaps_col = cumsum(table(col_annot$time))[-length(table(col_annot$time))],
         annotation_row = row_annot,
         annotation_col = col_annot,
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors,
                                  "response_type" = c("Monotonous" = "#D8AF39FF", "Transient" = "#278B9AFF")))

png(file.path(outdir_plots,"ImpulseDE2_heatmap_DOWN_ALL.png"),  width = 7, height = 10, units = "in", res = 300)
heatmap_down_all
dev.off()

pdf(paste0(outdir_plots,"/pdf_figs/ImpulseDE2_heatmap_DOWN_ALL.pdf"), width = 7, height = 10)
heatmap_down_all
dev.off()
```

### Top gene trajectories

```{r}
# Plot top 10 differentially expressed (by q-value) genes
top_genes <- impulse_sig %>% arrange(padj) %>% head(10) %>% pull(Gene)

plotGenes(
  vecGeneIDs = top_genes,
  objectImpulseDE2 = objectImpulseDE2,
  boolSimplePlot = TRUE,
  boolCaseCtrl     = TRUE,
  dirOut = outdir_plots,
  strFileName = "/top10_DE_genes.pdf",
  boolMultiplePlotsPerPage = FALSE)
```

---

# Mfuzz: Cluster ImpulseDE2-significant genes by trajectory

## 1. Preparing expression data

Filter vst transformed counts to keep significant genes and only heat-samples

```{r mfuzz-prep1}
sig_genes <- impulse_sig %>% filter(Gene %in% rownames(vst)) %>% pull(Gene)

# Get VST expression for case (heat) samples only for sig genes
heat_vst <- vst[sig_genes, heat_samples]
```

Average across replicates per timepoint, heat samples only

```{r mfuzz-prep2}
heat_avg <- vst[sig_genes, heat_samples] %>%
  as.data.frame() %>%
  rownames_to_column("Gene") %>%
  pivot_longer(-Gene, names_to = "sample", values_to = "expr") %>%
  left_join(meta %>% select(sample, time), by = "sample") %>%
  group_by(Gene, time) %>%
  summarize(mean_expr = mean(expr), .groups = "drop") %>%
  pivot_wider(names_from = time, values_from = mean_expr) %>%
  column_to_rownames("Gene")

# Reorder columns by time
heat_avg <- heat_avg[, order(as.numeric(colnames(heat_avg)))]
colnames(heat_avg) <- paste0("R", colnames(heat_avg))
```

Prepare data in eset format

```{r}
# Create ExpressionSet
heat_eset <- ExpressionSet(assayData = as.matrix(heat_avg))
heat_eset <- standardise(heat_eset)
```

## 2. Determine Mfuzz parameters

For fuzzy c-means clustering, the fuzzifier m and the number of clusters c has to be chosen in advance

```{r}
# Estimate fuzzifier
m <- mestimate(heat_eset)
cat("Estimated fuzzifier m:", round(m, 2), "\n")
```

```{r, eval=FALSE}
# Choose optimal cluster number based on elbow plot (do this once and add to species_parameters.R script)
Dmin(heat_eset, m = m, crange = seq(2, 12, 1), repeats = 3)

# adjust below to test what different numbers of clusters reveal in the data -- don't over-cluster
for (k in c(4,6,8)) {
  set.seed(global_params$seed)
  result <- mfuzz(heat_eset, c = k, m = m)
  mfuzz.plot(heat_eset, cl = result, mfrow = c(2, ceiling(k/2)), 
             new.window =FALSE,, time.labels =  c(0,1,3,12,24,72,120))
}
```

## 3. Run MFuzz

Run clustering with species-specific cluster number

```{r mfuzz}
k <-  config$n_clusters
set.seed(global_params$seed)
mfuzz_clusters <- mfuzz(heat_eset, c = k, m = m)
```

Extract cluster assignments and save results

```{r}
cluster_assignments <- data.frame(
  Gene = names(mfuzz_clusters$cluster),
  cluster = mfuzz_clusters$cluster,
  membership = apply(mfuzz_clusters$membership, 1, max)) %>%
  left_join(impulse_sig %>% select(Gene, response_type), by = "Gene") # join to impulseDE results

table(cluster_assignments$cluster, cluster_assignments$response_type)

# Save mfuzz objects and cluster assignments
saveRDS(mfuzz_clusters, file.path(outdir_mfuzz, "mfuzz_result.rds"))
saveRDS(heat_eset, file.path(outdir_mfuzz, "mfuzz_input_eset.rds"))
write.csv(cluster_assignments, file.path(outdir_mfuzz, "cluster_assignments.csv"), row.names = FALSE)
write.csv(as.data.frame(mfuzz_clusters$centers), file.path(outdir_mfuzz, "cluster_centers.csv"))
```

## 4. Characterize Clusters

```{r}
# Get cluster centers (average trajectory)
cluster_centers <- mfuzz_clusters$centers

# Identify peak and trough timepoint for each cluster
timepoints <- c(0, 1, 3, 12, 24, 72, 120)
peak_times <- timepoints[apply(cluster_centers, 1, which.max)]
trough_times <- timepoints[apply(cluster_centers, 1, which.min)]

cluster_info <- data.frame(
  cluster = 1:k,
  peak_time = peak_times,
  trough_time = trough_times,
  n_genes = as.numeric(table(mfuzz_clusters$cluster)))

print(cluster_info)
write.csv(cluster_info, file.path(outdir_mfuzz, "cluster_info.csv"), row.names = FALSE)
```

### Visualize all significant genes in their clusters

```{r}
mfuzz.plot(heat_eset, cl = mfuzz_clusters, new.window =FALSE, mfrow = c(2,k/2), time.labels =  c(0,1,3,12,24,72,120))

mfuzz.plot(heat_eset, cl = mfuzz_clusters, new.window =FALSE, mfrow = c(2,k/2), time.labels =  c(0,1,3,12,24,72,120), min.mem = 0.5)

# Visualize clusters
png(paste0(outdir_plots,"/temporal_clusters.png"), width = 12, height = 10, units = "in", res = 300)
mfuzz.plot2(heat_eset, cl = mfuzz_clusters, mfrow = c(2,k/2),
            time.labels = c("0", "1", "3", "12", "24", "72", "120"),
            xlab = "Time (hours)",x11=FALSE)
dev.off()

pdf(paste0(outdir_plots,"/pdf_figs/temporal_clusters.pdf"), width = 12, height = 10)
mfuzz.plot2(heat_eset, cl = mfuzz_clusters, mfrow = c(2,k/2),
            time.labels = c("0", "1", "3", "12", "24", "72", "120"),
            xlab = "Time (hours)",x11=FALSE)
dev.off()

png(paste0(outdir_plots,"/temporal_clusters_membership50.png"), width = 12, height = 10, units = "in", res = 300)
mfuzz.plot2(heat_eset, cl = mfuzz_clusters, mfrow = c(2,k/2),
            min.mem = 0.5,
            time.labels = c("0", "1", "3", "12", "24", "72", "120"),
            xlab = "Time (hours)",x11=FALSE)
dev.off()

pdf(paste0(outdir_plots,"/pdf_figs/temporal_clusters_membership50.pdf"), width = 12, height = 10)
mfuzz.plot2(heat_eset, cl = mfuzz_clusters, mfrow = c(2,k/2),
            min.mem = 0.5,
            time.labels = c("0", "1", "3", "12", "24", "72", "120"),
            xlab = "Time (hours)",x11=FALSE)
dev.off()
```


```{r}
png(paste0(outdir_plots,"/temporal_clusters_scaled.png"), width = 12, height = 10, units = "in", res = 300)
mfuzz.plot2(heat_eset, cl = mfuzz_clusters, mfrow = c(2,k/2),
            time.points = c(0, 1, 3, 12, 24, 72, 120),
            time.labels = c("0", "1", "3", "12", "24", "72", "120"),
            xlab = "Time (hours)",x11=FALSE)
dev.off()
```

```{r}
standardized_avg_heat_expr <- as_tibble(exprs(heat_eset),rownames="gene")
standardized_avg_heat_expr <- standardized_avg_heat_expr %>% pivot_longer(cols=starts_with("R"), names_to="timepoint",values_to="zscore")
standardized_avg_heat_expr <- standardized_avg_heat_expr %>% left_join(cluster_assignments, by=join_by("gene"=="Gene"))
standardized_avg_heat_expr <- standardized_avg_heat_expr %>%
  mutate(time=as.numeric(str_remove(timepoint,"R")),
         cluster=factor(cluster))

standardized_avg_heat_expr %>% ggplot(aes(x=time, y=zscore)) + 
  geom_line(aes(group=gene),alpha=0.01)+
  stat_summary(fun = mean, geom = "line", linewidth = 1,mapping=aes(color=cluster)) + theme_bw() + facet_grid(~cluster)

centers <- as_tibble(cluster_centers,rownames="cluster")
centers <- centers %>% pivot_longer(cols=starts_with("R"), names_to="timepoint",values_to="center")
centers <- centers %>%
  mutate(time=as.numeric(str_remove(timepoint,"R")),
         cluster=factor(cluster))

centers %>% ggplot(aes(x = as.numeric(factor(time)), y = center, group=cluster,color = cluster)) +
  geom_smooth(method = "loess",span = .7,se=FALSE,linewidth = 3) +
  theme_minimal() + facet_grid(~cluster) + theme_void() + theme(legend.position = "none")

save_ggplot(last_plot(),"cluster_centers_smooth", width=16, height=4)
```

---

## Exploring genes of interest

### Manually-curated heat stress genes by cluster

```{r}
# read in Manual heat stress genes annotation
HeatStressGenes <- read_csv(paste0(annot_dir,"/heatstress/HeatStressGenes_", species ,".csv")) %>%
  dplyr::select(-1) %>% dplyr::rename(query = paste0(species,"_gene")) %>% dplyr::select(query,everything())

HeatStressGenes_unique <- HeatStressGenes %>% group_by(query) %>% arrange(gene_id) %>%
  summarize(gene_id = paste(unique(gene_id), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")) 

HeatStressGenes_unique <- HeatStressGenes_unique %>% filter(query %in% rownames(vst))
 
stress_genes_ids <- unique(HeatStressGenes_unique$query) 
```

```{r}
HSPS <- impulse_results %>% filter(Gene %in% stress_genes_ids) %>% arrange(padj) %>% left_join(HeatStressGenes_unique, by = join_by(Gene==query)) %>% filter(grepl("HSP",gene_id)) %>% pull(Gene)

cluster_assignments %>% filter(Gene %in% HSPS)
```

```{r}
heat_clustered <- HeatStressGenes_unique %>% left_join(cluster_assignments, by = join_by(query==Gene)) %>%
  arrange(cluster, desc(cluster),gene_id)

# plot this to show which genes are in which cluster
heat_clustered %>% filter(!is.na(cluster)) %>% ggplot(aes(y=reorder(gene_id, cluster), x=factor(cluster), fill=cluster)) +
  geom_tile() +
  theme_bw() +
  labs(x="Mfuzz Cluster", y="Gene ID", title="Heat stress genes clustered by temporal expression pattern")
```

```{r}
sessionInfo()

detach(package:Mfuzz, unload=TRUE)
detach(package:ImpulseDE2, unload=TRUE)
detach(package:pheatmap, unload=TRUE)
detach(package:ComplexHeatmap, unload=TRUE)
```

