---
title: "WGCNA Analysis"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
params:
  species: "Mcap" # default for interactive use
  TestParams: FALSE # whether to run the code to determine WGCNA parameters, which takes a long time to run. Default is FALSE, will use parameters in species_parameters.R
  run_WGCNA: FALSE # whether to run the WGCNA module detection code, which takes a long time to run. Default is FALSE  and will load results from output_RNA/WGCNA/{species}/wgcna_results.RDS
output:
  github_document:
    toc: true
    html_preview: false
---

# Network analysis of Time Series bulk RNA-seq data

## Introduction

The goal of this script is to identify co-expressed gene modules from our time-course RNA-seq data. I hope to identify genes which respond to the heat stress similarly over time.

Helpful *general* WGCNA tutorial to help with parameter decisions can be found [here](https://alexslemonade.github.io/refinebio-examples/04-advanced-topics/network-analysis_rnaseq_01_wgcna.html#46_Determine_parameters_for_WGCNA) and another [here](https://bioinformaticsworkbook.org/tutorials/wgcna.html#gsc.tab=0).

And other example code here:
- https://github.com/fscucchia/HI_PhotoPhysio_TPC_geneExpr/blob/983b837e2dbd9bad2dfeda764dcc0b9da254073a/Gene_expression/scripts/WGCNA_Network_Analysis/WGCNA_Pacu.r

How to install packages:

```{r, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("impute", type = "source")
BiocManager::install("WGCNA",force = TRUE)
```

---

## 1. Load packages and functions

```{r setup}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath(paste0("../../output_RNA/reports/", params$species, "/")), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 10, fig.height = 8,
                      fig.path = "03_WGCNA_files/figure-gfm/")

#load packages
library(tidyverse)
library(WGCNA)

#load in parameters and functions
source("species_parameters.R")
source("utils.R")

sessionInfo() #provides list of loaded packages and version of R
```

## 2. Setup species-specific parameters

```{r get-species}
# get species
species <- params$species

# get parameters for this species
config <- get_params(species)
print_config(species)

# define preprocessing output directory (from 01_preprocessing.Rmd)
input_dir <- file.path("../../output_RNA/counts_filt_norm", species)

# set up necessary output directories if they don't exist
outdir <- file.path("../../output_RNA/WGCNA", species)
outdir_plots <- file.path(outdir,"plots")
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
if (!dir.exists(outdir_plots)) dir.create(outdir_plots, recursive = TRUE)

reportdir <- file.path("../../output_RNA/reports", params$species, "03_WGCNA_files/figure-gfm/")
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)
```
## 3. Setup WGCNA Parameters and plotting function

```{r params}
options(stringsAsFactors = FALSE)
enableWGCNAThreads(global_params$n_cores)
```

Create Heatmap function based heavily off of the following tutorial:

https://alexslemonade.github.io/refinebio-examples/04-advanced-topics/network-analysis_rnaseq_01_wgcna.html#46_Determine_parameters_for_WGCNA

```{r heatmap-function}
make_module_heatmap <- function(module_name,
                                expression_mat = normalized_counts,
                                metadata_df = meta,
                                gene_module_key_df = module_df,
                                module_eigengenes_df = module_eigengenes) {
  # Create a summary heatmap of a given module.
  # based on https://alexslemonade.github.io/refinebio-examples/04-advanced-topics/network-analysis_rnaseq_01_wgcna.html#46_Determine_parameters_for_WGCNA

  # Set up the module eigengene with its sample
  module_eigengene <- module_eigengenes_df %>%
    dplyr::select(all_of(module_name)) %>%
    tibble::rownames_to_column("sample")

  # Set up column annotation from metadata
  col_annot_df <- metadata_df %>%
    # Only select the treatment, time, and sample ID columns
    dplyr::select(sample, treatment, time) %>%
    # Add on the eigengene expression by joining with sample IDs
    dplyr::inner_join(module_eigengene, by = "sample") %>%
    # Arrange by treatment and time point
    dplyr::arrange(treatment, time, sample) %>%
    # Store sample
    tibble::column_to_rownames("sample")

  # Create the ComplexHeatmap column annotation object
  col_annot <- ComplexHeatmap::HeatmapAnnotation(
    # Supply treatment and time labels
    treatment = col_annot_df$treatment,
    time = col_annot_df$time,
    # Add annotation barplot
    module_eigengene = ComplexHeatmap::anno_barplot(dplyr::select(col_annot_df, module_name)),
    # Pick colors for each experimental group in treatment
    col = list(treatment = c("C" = "lightblue4", "H" = "#D55E00"),
               time = time_colors)
  )

  # Get a vector of the gene IDs that correspond to this module
  module_genes <- gene_module_key_df %>%
    dplyr::filter(module == module_name) %>%
    dplyr::pull(gene_id)

  # Set up the gene expression data frame
  mod_mat <- expression_mat %>%
    t() %>%
    as.data.frame() %>%
    # Only keep genes from this module
    dplyr::filter(rownames(.) %in% module_genes) %>%
    # Order the samples to match col_annot_df
    dplyr::select(rownames(col_annot_df)) %>%
    # Data needs to be a matrix
    as.matrix()

  # Normalize the gene expression values
  mod_mat <- mod_mat %>%
    # Scale can work on matrices, but it does it by column so we will need to
    # transpose first
    t() %>%
    scale() %>%
    # And now we need to transpose back
    t()

  # Create a color function based on standardized scale
  color_func <- circlize::colorRamp2(
    c(-1.5, 0, 1.5),
    c("#67a9cf", "#f7f7f7", "#ef8a62")
  )

  # Plot on a heatmap
  heatmap <- ComplexHeatmap::Heatmap(mod_mat,
    name = module_name,
    # Supply color function
    col = color_func,
    # Supply column annotation
    bottom_annotation = col_annot,
    # We don't want to cluster samples
    cluster_columns = FALSE,
    # We don't need to show sample or gene labels
    show_row_names = FALSE,
    show_column_names = FALSE
  )

  # Return heatmap
  return(heatmap)
}
```

## 4. Load in transformed counts and metadata

```{r load-data}
# load in vst-transformed counts
vst <- read.csv(file.path(input_dir, "vsd_expression_matrix.csv"))
vst <- vst %>% column_to_rownames(var = "X")

# transpose counts to format needed for WGCNA
normalized_counts <- t(vst)

# load in metadata
meta <- read.csv(paste0("../../output_RNA/",species,"_RNA_seq_metadata.csv"))
meta <- meta %>% column_to_rownames(var = "X") %>% select(-c(species, replicate))

cat("Input data:", nrow(vst), "genes x", ncol(vst), "samples")
```

## 5. Check for outliers

Ideally these were removed during 01_preprocessing.Rmd. See line above to confirm the number of samples matches what you expect after outlier removal. But confirm by clustering the samples and view as a tree.

```{r sample-tree}
sampleTree = hclust(dist(normalized_counts), method = "average")
plot(sampleTree, main = paste(species, "- Sample Clustering"), 
     xlab = "", sub = "", cex = 0.6)
```

## 6. Determine parameters for WGCNA

The below takes a long time to run, so it is only run if TestParams is set to TRUE. Otherwise, the pre-determined parameters for this species dataset are loaded in from the species_parameters.R script. This should be redone for any changes in data filtering and outlier removal.

```{r determine-params}
if(params$TestParams == TRUE) {
  sft <- pickSoftThreshold(normalized_counts,
                           networkType = "signed",
                           RsquaredCut = 0.8,
                           powerVector = c(seq(1, 12, by = 1), seq(14, 30, by = 2)),
                           verbose=3)
  sft_df <- data.frame(sft$fitIndices) %>% dplyr::mutate(model_fit = -sign(slope) * SFT.R.sq)
  
  fit_plot <- ggplot(sft_df, aes(x = Power, y = model_fit, label = Power)) +
    geom_point() +
    geom_text(nudge_y = 0.1) +
    # We will plot what WGCNA recommends as an R^2 cutoff
    geom_hline(yintercept = 0.80, col = "red") +
    ylim(c(min(sft_df$model_fit), 1.05)) +
    xlab("Soft Threshold (power)") +
    ylab("Scale Free Topology Model Fit, signed R^2") +
    ggtitle("Scale independence") +
    theme_classic()
  
  print(fit_plot)
  
  mean_plot <- ggplot(sft_df, aes(x = Power, y = mean.k., label = Power)) +
    geom_point() +
    geom_text(nudge_y = 500) +
    xlab("Soft Threshold (power)") +
    ylab("Mean Connectivity") +
    ggtitle("Mean Connectivity") +
    theme_classic()
  
  print(mean_plot)
  
  soft_power = sft$powerEstimate
  
  if (is.na(sft$powerEstimate)) {
    stop("Soft power could not be automatically determined. Potenitally test a greater range of powers.")
}
  
  if (soft_power != config$soft_power) {
    warning(paste0(" Calculated power (" , sft$powerEstimate, 
                  ") differs from config value (", config$soft_power,"). Consider updating species_parameters.R with new value. Examine graph and confirm if the calculated power matches visual examination of the data."))
  }
  
} else {
  soft_power = config$soft_power
}

cat("Soft Power for WGCNA:", soft_power)
```

## 7. WGCNA: One-step module detection

```{r wgcna}
if(params$run_WGCNA == TRUE) {
  temp_cor <- cor
  cor <- WGCNA::cor # Force it to use WGCNA cor function (fix a namespace conflict issue)
  netwk <- blockwiseModules(normalized_counts,
                            nThreads = global_params$n_cores,
  
                            # Adjacency Function
                            power = soft_power,
                            corType = "bicor",
                            networkType = "signed",
                            TOMType = "signed",
  
                            # Tree and Block Options
                            deepSplit = global_params$wgcna_default$deep_split,
                            pamRespectsDendro = F,
                            minModuleSize = global_params$wgcna_default$min_module_size,
                            maxBlockSize = 50000,
  
                            # topological overlap matrix, (TOM)
                            saveTOMs = TRUE,
                            saveTOMFileBase = file.path(outdir, "blockwiseTOM"),
                            #loadTOM = FALSE, #uncomment this if you are re-running with a previously saved TOM
  
                            # Output Options
                            mergeCutHeight = global_params$wgcna_default$merge_cut_height,
                            numericLabels = TRUE,
                            verbose = 3)
  
  cor <- temp_cor     # Return cor function to original namespace
  saveRDS(netwk, file.path(outdir, "wgcna_network.rds"))
}
```

### Load saved WGCNA results

```{r load-wgcna}
netwk <- readRDS(file.path(outdir, "wgcna_network.rds"))

# what is stored in this object?
names(netwk)

# save the module labels
moduleLabels <- netwk$colors

# how many modules are there?
paste("There are", length(unique(moduleLabels)), "modules in our current analysis.")

# see the distribution of genes across these labelled modules
table(moduleLabels)

# Convert labels to colors for plotting
moduleColors <- labels2colors(moduleLabels)
```

### Visualize the network

```{r wgcna-dendro}
# Plot the dendrogram and the module colors underneath
plotDendroAndColors(
  netwk$dendrograms[[1]],
  moduleColors,
  "Module colors",
  dendroLabels = FALSE,
  hang = 0.03,
  addGuide = TRUE,
  guideHang = 0.05, main = "Consensus gene dendrogram and module colors")
```

<!-- ### Treatment and Time Module Correlation -->

<!-- ```{r} -->
<!-- # save the module info as a dataframe and txt file -->
<!-- module_df <- data.frame( -->
<!--   gene_id = names(netwk$colors), -->
<!--   module = paste0("ME", netwk$colors), -->
<!--   color = labels2colors(netwk$colors) -->
<!-- ) -->

<!-- head(module_df) -->
<!-- write_delim(module_df, file = paste0(outdir,"gene_modules.txt"), delim = "\t") -->

<!-- # get the module eigengenes -->
<!-- module_eigengenes <- netwk$MEs -->
<!-- head(module_eigengenes) -->

<!-- # get a list of all the genes in a module -->
<!-- gene_module_key <- tibble::enframe(netwk$colors, name = "gene", value = "module") %>% -->
<!--   # Let's add the `ME` part so its more clear what these numbers are and it matches elsewhere -->
<!--   dplyr::mutate(module = paste0("ME", module)) -->

<!-- # confirm that the sample metadata and sample labels for the module eigengenes are matching -->
<!-- all.equal(meta$sample, rownames(module_eigengenes)) -->
<!-- ``` -->

<!-- #### Time+Treatment-Module Correlation Heatmaps -->

<!-- ```{r} -->
<!-- nSamples = nrow(normalized_counts) -->

<!-- time_treat_factorial <- meta %>% -->
<!--   mutate(group = paste0(time, "hr-", ifelse(treatment == "C", "Control", "Heat"))) %>% -->
<!--   select(sample, group) %>%  -->
<!--   mutate(value = 1) %>%  -->
<!--   tidyr::pivot_wider(names_from = group, values_from = value, values_fill = 0) %>% -->
<!--   column_to_rownames(var = "sample") %>% -->
<!--   relocate(contains("Control"), contains("Heat")) %>% -->
<!--    mutate(across(everything(), as.factor)) -->

<!-- # Reorder modules so similar modules are next to each other -->
<!-- module_eigengenes_ordered <- orderMEs(module_eigengenes) -->
<!-- module_order = names(module_eigengenes_ordered)  -->

<!-- moduleTraitCor =  WGCNA::cor(module_eigengenes_ordered, time_treat_factorial, use = "p"); -->
<!-- moduleTraitPvalue = corPvalueStudent(moduleTraitCor, nSamples); -->

<!-- textMatrix <- ifelse(moduleTraitPvalue < 0.05, -->
<!--                      paste0(signif(moduleTraitCor, 2), "\n(", -->
<!--                             signif(moduleTraitPvalue, 2), ")"),"") -->

<!-- pdf(paste0(outdir,"/all_heatmap.pdf"),width=8, height=8) -->
<!-- # Will display correlations and their p-values -->

<!-- par(mar = c(4, 3, 2, 2)) -->
<!-- labeledHeatmap(Matrix = moduleTraitCor, -->
<!--                textMatrix = textMatrix, -->
<!--                xLabels = names(time_treat_factorial), -->
<!--                yLabels = names(module_eigengenes_ordered), -->
<!--                ySymbols = names(module_eigengenes_ordered), -->
<!--                colorLabels = TRUE, -->
<!--                colors = blueWhiteRed(100), -->
<!--                setStdMargins = FALSE, -->
<!--                cex.text = 0.5, -->
<!--                cex.lab = 0.7, -->
<!--                cex.colorLabels = 0.7, -->
<!--                zlim = c(-1,1), -->
<!--                main = paste("Module-trait relationships - all")) -->

<!-- dev.off() -->
<!-- ``` -->

<!-- ##### ggplot version -->

<!-- ```{r} -->
<!-- # Add treatment names -->
<!-- module_eigengenes_ordered$treatment_time = paste0(meta$time,"hr","-",ifelse(meta$treatment == "C", "Control", "Heat")) -->
<!-- module_eigengenes_ordered$treatment = meta$treatment -->
<!-- module_eigengenes_ordered <- module_eigengenes_ordered %>% arrange(treatment) -->

<!-- mmPval = moduleTraitPvalue %>% as.data.frame() %>% rownames_to_column("module") %>% -->
<!--   pivot_longer(-module, names_to = "treatment_time", values_to = "pvalue") -->

<!-- mmCor = moduleTraitCor %>% as.data.frame() %>% rownames_to_column("module") %>% -->
<!--   pivot_longer(-module, names_to = "treatment_time", values_to = "correlation") %>% -->
<!--   left_join(mmPval, by = c("module", "treatment_time")) %>% -->
<!--   mutate( -->
<!--     label = ifelse(pvalue < 0.05, -->
<!--                    paste0(signif(correlation, 2)), ""), -->
<!--     # use this if you want the p-value also plotted -->
<!--     #label = ifelse(pvalue < 0.05, -->
<!--     #               paste0(signif(correlation, 2), "\n(", signif(pvalue, 2), ")"), ""), -->
<!--     module = factor(module, levels = rev(module_order)), -->
<!--     treatment_time = factor(treatment_time, levels = unique(module_eigengenes_ordered$treatment_time))) -->

<!-- ggplot(mmCor, aes(x=treatment_time, y=module, fill=correlation)) + -->
<!--   geom_tile(color = "white", linewidth = 0.3) + -->
<!--   geom_text(aes(label = label), size = 3, color = "black") + -->
<!--   theme_minimal(base_size = 12) + -->
<!--   scale_fill_gradient2( -->
<!--     low = "#4575B4", -->
<!--     high = "#D73027", -->
<!--     mid = "white", -->
<!--     midpoint = 0, -->
<!--     limits = c(-1, 1)) + -->
<!--   labs(title = "Module-trait Relationships", y = "Modules", fill="Correlation")+  -->
<!--   theme( -->
<!--     axis.text.x = element_text(angle = 45, hjust = 1), -->
<!--     panel.grid = element_blank(), -->
<!--     axis.ticks = element_blank() -->
<!--   ) +coord_fixed(ratio = 0.7)  -->

<!-- save_ggplot(plot = last_plot(), filename = "all_heatmap_ggplot", width = 8, height = 8) -->
<!-- ``` -->

<!-- ##### ID peak times based on correlation -->

<!-- ```{r} -->
<!-- module_peak_times <- mmCor %>% -->
<!--   filter(pvalue < 0.05, grepl("Heat",treatment_time)) %>% -->
<!--   group_by(treatment_time) %>% -->
<!--   summarize( -->
<!--     n_modules = n(), -->
<!--     mean_abs_cor = mean(abs(correlation)) -->
<!--   ) %>% -->
<!--   extract(treatment_time, "time", "([0-9]+)hr", convert = TRUE) -->
<!-- ``` -->


<!-- #### Run linear model on each module vs. treatment -->

<!-- ```{r} -->
<!-- # Create the design matrix for full (with interaction) models, use factor for time since non-evenly spaced intervals -->

<!-- meta$time_factor <- factor(meta$time) -->
<!-- des_mat_full <- model.matrix(~ treatment*time_factor, data = meta) -->
<!-- head(des_mat_full) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- # lmFit() needs a transposed version of the matrix -->
<!-- fit_full <- limma::lmFit(t(module_eigengenes), design = des_mat_full) -->

<!-- # Apply empirical Bayes to smooth standard errors -->
<!-- fit_full <- limma::eBayes(fit_full) -->

<!-- # Apply multiple testing correction and obtain stats -->

<!-- ## interaction <- treatment effect differs by time -->
<!-- interaction_coefs <- grep("treatment.*:time", colnames(des_mat_full), value = TRUE) -->

<!-- stats_interaction <- limma::topTable(fit_full, coef = interaction_coefs, number = ncol(module_eigengenes)) %>% -->
<!--   tibble::rownames_to_column("module") -->

<!-- stats_df_full <- limma::topTable(fit_full, number = ncol(module_eigengenes)) %>% -->
<!--   tibble::rownames_to_column("module") -->

<!-- # we care most about the interaction, the full model will pull out modules that vary in both treatments by time also (like a circadian rhythm 0hr vs 12hr difference) -->

<!-- # almost all of our modules significant by this model: -->
<!-- stats_interaction %>% filter(adj.P.Val < 0.05)  %>% nrow() -->

<!-- #save these as a vector -->
<!-- top_mod_sig_interaction <- stats_interaction %>% filter(adj.P.Val < 0.05)  %>% pull(module) -->

<!-- # print the top 5: -->
<!-- stats_interaction %>% filter(adj.P.Val < 0.05)  %>% head(5) -->
<!-- ``` -->

<!-- Module 1 is the most differentially expressed across treatments + in the full model. -->

<!-- #### Plot example module over time -->

<!-- ```{r} -->
<!-- eigengenes_treatment_df <- module_eigengenes %>% -->
<!--   tibble::rownames_to_column("sample") %>% -->
<!--   dplyr::inner_join(meta %>% -->
<!--     dplyr::select(sample, treatment,time), -->
<!--   by = c("sample" = "sample")) -->

<!-- ggplot(eigengenes_treatment_df, aes(x = factor(time), y = ME1,color = treatment)) + -->
<!--   geom_boxplot(outlier.shape = NA) + -->
<!--   ggforce::geom_sina(size=1, alpha = 0.5) + -->
<!--   scale_color_manual(values = treat_colors) + -->
<!--   theme_classic() -->
<!-- ``` -->

<!-- #### Trajectory plots for all modules -->

<!-- ```{r} -->
<!-- eigengenes_treatment_df_long <- eigengenes_treatment_df %>% -->
<!--   pivot_longer(cols = starts_with("ME"), -->
<!--                names_to = "module", -->
<!--                values_to = "eigengene_value") %>% -->
<!--   mutate(module = factor(module, levels = module_order)) %>% -->
<!--   mutate(module_label = ifelse(module %in% top_mod_sig_interaction,  -->
<!--                                 paste0("*",module),  -->
<!--                                 as.character(module))) -->

<!-- ggplot(eigengenes_treatment_df_long, aes(x = factor(time), y = eigengene_value,color = treatment)) + -->
<!--   geom_boxplot(outlier.shape = NA) + -->
<!--   ggforce::geom_sina(size=1, alpha = 0.5) + -->
<!--   scale_color_manual(values = treat_colors) + -->
<!--   facet_wrap(~module_label, ncol = 5) + -->
<!--   theme_classic() + theme( -->
<!--     strip.text = element_text(size = 8, face = "bold"), -->
<!--     axis.text = element_text(size = 7), -->
<!--     legend.position = "bottom") + -->
<!--   labs(x = "Time (hours)", y = "Module Eigengene") -->

<!-- save_ggplot(plot = last_plot(), filename = "all_modules", width = 14, height = 12) -->

<!-- eigengenes_summary <- eigengenes_treatment_df_long %>% -->
<!--   group_by(module, module_label, time, treatment) %>% -->
<!--   summarize(mean_value = mean(eigengene_value), -->
<!--             se = sd(eigengene_value) / sqrt(n()), -->
<!--             .groups = "drop") -->

<!-- ggplot(eigengenes_summary, aes(x = factor(time), y = mean_value, color = treatment, group = treatment)) + -->
<!--   geom_line(linewidth = 0.5) + -->
<!--   geom_errorbar(aes(ymin = mean_value - se, ymax = mean_value + se), width = 0.2) + -->
<!--   scale_color_manual(values = treat_colors) + -->
<!--   facet_wrap(~module_label, ncol = 5, scales = "free_y") + -->
<!--   theme_classic() + -->
<!--   theme( -->
<!--     strip.text = element_text(size = 8, face = "bold"), -->
<!--     axis.text = element_text(size = 7), -->
<!--     legend.position = "bottom") + -->
<!--   labs(x = "Time (hours)", y = "Module Eigengene") -->

<!-- save_ggplot(plot = last_plot(), filename = "all_modules_lines", width = 14, height = 12) -->
<!-- ``` -->

<!-- #### Individual module heatmaps -->

<!-- ```{r} -->
<!-- make_module_heatmap(module_name = "ME1") -->
<!-- make_module_heatmap(module_name = "ME0") -->
<!-- make_module_heatmap(module_name = "ME3") -->
<!-- ``` -->
