---
title: "Transcription Factor Binding Sites Analysis"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
params:
  species: "Pcomp" # default for interactive use
output:
  github_document:
    toc: true
    html_preview: false
---

# Analysis of Time Series bulk RNA-seq data: Transcription Factor Binding Sites (TFBS) Analysis

## Introduction

In my annotation [repository](https://github.com/zdellaert/HI_genome_annotations.git) (read about this is my [README.md](https://github.com/zdellaert/TimeSeries/blob/main/4-multi-species/code/RNA/README.md)), I identified the putative promoter regions (500 bp upstream) of all genes for the three species analyzed in this project. I used the program [FIMO](https://meme-suite.org/meme/tools/fimo) to identify putative binding sites for 3 stress-related transcription factors (FOXO3, HSF1, and NRF2/NFE2L2) in these promoter regions. Here, I load in the FIMO output and quantify the number of putative binding sites for each TF in the promoter region of expressed genes in my RNA-seq dataset.

TFs analyzed:

  - **HSF1** - Heat Shock Factor 1 (primary regulator of heat shock response)
    - Human binding site motif used: `wget "https://jaspar.elixir.no/api/v1/matrix/MA0486.1.meme"`
  - **FOXO3** - Forkhead box O3 (potentially involved in coral apoptosis/autophagy response)
    - Human binding site motif used: `wget "https://jaspar.elixir.no/api/v1/matrix/MA0157.2.meme"`
  - **NFE2L2** (Nrf2) - Nuclear factor erythroid 2-related factor 2 (oxidative stress response)
    - Human binding site motif used: `wget "https://jaspar.elixir.no/api/v1/matrix/MA0150.1.meme"`
  
This analysis was originally inspired by the paper *Cleves PA, Krediet CJ, Lehnert EM, Onishi M, Pringle JR. Insights into coral bleaching under heat stress from analysis of gene expression in a sea anemone model system. Proceedings of the National Academy of Sciences. 2020 Nov 17;117(46):28906–17. (https://www.pnas.org/doi/10.1073/pnas.2015737117)* and the updated analysis in the paper *Swinhoe N, Tinoco AI, Sarfati DN, Henderson CF, Kowalewski GP, Meier EK, et al. CRISPR/Cas9-mutagenesis reveals that varying dependence on HSF1 is associated with differences in coral heat tolerance. bioRxiv; 2026. p. 2026.04.01.714264. doi:10.64898/2026.04.01.714264* by the same lab.

### Fimo code used (for reference, not run in this Rmd since it was run in my annotation [repository](https://github.com/zdellaert/HI_genome_annotations.git)):

Fimo was run using the docker image of MEMEsuite version 5.5.9 in March 2026.

```{bash, eval=FALSE}
SINGULARITY_IMAGE="docker://memesuite/memesuite:latest"

    # run FIMO with default settings (--thresh 0.0001, --max-stored-scores 100000)
    singularity exec --cleanenv $SINGULARITY_IMAGE fimo \
        -oc "fimo_output/${sp}_stress_TFs" \
        --thresh 0.0001 \
        --max-stored-scores 100000 \
        ../../references/motif_databases/stress_TFs.meme \
        "$fasta"
```

---

## 1. Load packages and functions

```{r setup}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath(paste0("../../output_RNA/reports/", params$species, "/")), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 10, fig.height = 8,
                      fig.path = "04_TFBS_files/figure-gfm/")

#load packages
library(tidyverse)

#load in parameters and functions
source("species_parameters.R")
source("utils.R")

sessionInfo() #provides list of loaded packages and version of R
```

## 2. Setup species-specific parameters and define directories

```{r get-species}
# get species
species <- params$species

# get parameters for this species
config <- get_params(species)
print_config(species)

# define preprocessing output directory (from 01_preprocessing.Rmd)
input_dir <- file.path("../../output_RNA/counts_filt_norm", species)

# FIMO path (from annotation repo)
fimo_path <- file.path(annot_dir, "promoters/fimo_output", paste0(species, "_stress_TFs"))
cat("\nFIMO path:", fimo_path, "\n")

# set up necessary output directories if they don't exist
outdir <- file.path("../../output_RNA/TFBS", species)
outdir_plots <- file.path(outdir,"plots")
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
if (!dir.exists(outdir_plots)) dir.create(outdir_plots, recursive = TRUE)

reportdir <- file.path("../../output_RNA/reports", params$species, "04_TFBS_files/figure-gfm/")
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)
```

## 3. Load in filtered counts and SwissProt annotations

```{r}
# load in filtered counts data
filtered_counts <- read.csv(file.path(input_dir, "filtered_counts.csv"), row.names = 1)

# SwissProt annotations
SwissProt <- read.delim(file.path(annot_dir,config$SwissProt))
cat("Annotations:", nrow(SwissProt), "Swissprot-annotated genes")
```

## 4. Load in Stress Transcription Factor Binding Site Data

```{r}
Stress_TFs <- read.delim(file.path(fimo_path,"fimo.tsv"), comment.char = "#")

# only keep genes in the filtered dataset
Stress_TFs <- Stress_TFs %>% filter(sequence_name %in% rownames(filtered_counts))
  
#get list of motifs/TFs present in the data
all_tfs <- unique(Stress_TFs$motif_alt_id)
cat("Transcription factor binding sites found for:", paste(unique(Stress_TFs$motif_alt_id), collapse = ", "), "\n")
```


```{r}
Stress_TFs_quantification <- Stress_TFs %>%
  group_by(motif_alt_id,sequence_name) %>%
  summarize(count=n(), .groups = "drop") 

Foxo3_quantification <- Stress_TFs_quantification %>% filter(motif_alt_id =="FOXO3") %>% left_join(SwissProt, by = join_by("sequence_name"=="query")) 
Hsf1_quantification <- Stress_TFs_quantification %>% filter(motif_alt_id =="HSF1") %>% left_join(SwissProt, by = join_by("sequence_name"=="query")) 
Nrf2_quantification <- Stress_TFs_quantification %>% filter(motif_alt_id =="NFE2L2") %>% left_join(SwissProt, by = join_by("sequence_name"=="query")) 

cat("Transcription Factor", "FOXO3:", nrow(Foxo3_quantification), "genes with putative binding sites.\nGene(s) with most binding sites:",
    Foxo3_quantification %>% filter(count == max(count)) %>% pull(sequence_name), "( Best Annotation:", Foxo3_quantification %>% filter(count == max(count)) %>% arrange(evalue) %>% head(1) %>% pull(ProteinNames)  %>% word(1:4), ") with", max(Foxo3_quantification$count), "binding sites.\n\n")

cat("Transcription Factor", "HSF1:", nrow(Hsf1_quantification), "genes with putative binding sites.\nGene(s) with most binding sites:",
    Hsf1_quantification %>% filter(count == max(count)) %>% pull(sequence_name), "( Best Annotation:", Hsf1_quantification %>% filter(count == max(count)) %>% arrange(evalue) %>% head(1) %>% pull(ProteinNames)  %>% word(1:4), ") with", max(Hsf1_quantification$count), "binding sites.\n\n")

cat("Transcription Factor", "NFE2L2:", nrow(Nrf2_quantification), "genes with putative binding sites.\nGene(s) with most binding sites:",
    Nrf2_quantification %>% filter(count == max(count)) %>% pull(sequence_name), "( Best Annotation:", Nrf2_quantification %>% filter(count == max(count)) %>% arrange(evalue) %>% head(1) %>% pull(ProteinNames)  %>% word(1:4), ") with", max(Nrf2_quantification$count), "binding sites.")
```

Make dataframe that collapses genes with sites from more than one TF in Stress_TFs_quantification so that there is only one row per gene.

```{r}
TFBS_by_gene <- Stress_TFs_quantification %>%
  pivot_wider(names_from = motif_alt_id, values_from = count, values_fill = 0, names_prefix = "count_") 

head(TFBS_by_gene %>% mutate(total_motifs=(`count_FOXO3` + `count_HSF1` + `count_NFE2L2`)) %>% arrange(desc(total_motifs)) %>% left_join(SwissProt %>% dplyr::select(query,ProteinNames,blast_hit), by = join_by("sequence_name"=="query")) )

write.csv(TFBS_by_gene, file.path(outdir, "TFBS_counts.csv"), row.names = FALSE)
```

```{r}
sessionInfo()
```

