---
title: "Transcription Factor Binding Sites Analysis"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
params:
  species: "Pcomp" # default for interactive use
output:
  github_document:
    toc: true
    html_preview: false
---

# Analysis of Time Series bulk RNA-seq data: Transcription Factor Binding Sites (TFBS) Analysis

## Introduction

In my annotation [repository](https://github.com/zdellaert/HI_genome_annotations.git) (read about this is my [README.md](https://github.com/zdellaert/TimeSeries/blob/main/4-multi-species/code/RNA/README.md)), I identified the putative promoter regions (500 bp upstream) of all genes for the three species analyzed in this project. I used the program [FIMO](https://meme-suite.org/meme/tools/fimo) to identify putative binding sites for 3 stress-related transcription factors (FOXO3, HSF1, and NRF2/NFE2L2) in these promoter regions. Here, I load in the FIMO output and quantify the number of putative binding sites for each TF in the promoter region of expressed genes in my RNA-seq dataset.

TFs analyzed:

  - **HSF1** - Heat Shock Factor 1 (primary regulator of heat shock response)
    - Human binding site motif used: `wget "https://jaspar.elixir.no/api/v1/matrix/MA0486.1.meme"`
  - **FOXO3** - Forkhead box O3 (potentially involved in coral apoptosis/autophagy response)
    - Human binding site motif used: `wget "https://jaspar.elixir.no/api/v1/matrix/MA0157.2.meme"`
  - **NFE2L2** (Nrf2) - Nuclear factor erythroid 2-related factor 2 (oxidative stress response)
    - Human binding site motif used: `wget "https://jaspar.elixir.no/api/v1/matrix/MA0150.1.meme"`
  
This analysis was originally inspired by the paper *Cleves PA, Krediet CJ, Lehnert EM, Onishi M, Pringle JR. Insights into coral bleaching under heat stress from analysis of gene expression in a sea anemone model system. Proceedings of the National Academy of Sciences. 2020 Nov 17;117(46):28906–17. (https://www.pnas.org/doi/10.1073/pnas.2015737117)* and the updated analysis in the paper *Swinhoe N, Tinoco AI, Sarfati DN, Henderson CF, Kowalewski GP, Meier EK, et al. CRISPR/Cas9-mutagenesis reveals that varying dependence on HSF1 is associated with differences in coral heat tolerance. bioRxiv; 2026. p. 2026.04.01.714264. doi:10.64898/2026.04.01.714264* by the same lab.

### Fimo code used (for reference, not run in this Rmd since it was run in my annotation [repository](https://github.com/zdellaert/HI_genome_annotations.git)):

Fimo was run using the docker image of MEMEsuite version 5.5.9 in March 2026.

```{bash, eval=FALSE}
SINGULARITY_IMAGE="docker://memesuite/memesuite:latest"

    # run FIMO with default settings (--thresh 0.0001, --max-stored-scores 100000)
    singularity exec --cleanenv $SINGULARITY_IMAGE fimo \
        -oc "fimo_output/${sp}_stress_TFs" \
        --thresh 0.0001 \
        --max-stored-scores 100000 \
        ../../references/motif_databases/stress_TFs.meme \
        "$fasta"
```

---

## 1. Load packages and functions

```{r setup}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath(paste0("../../output_RNA/reports/", params$species, "/")), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 10, fig.height = 8,
                      fig.path = "04_TFBS_files/figure-gfm/")

#load packages
library(tidyverse)

#load in parameters and functions
source("species_parameters.R")
source("utils.R")

sessionInfo() #provides list of loaded packages and version of R
```

## 2. Setup species-specific parameters and define directories

```{r get-species}
# get species
species <- params$species

# get parameters for this species
config <- get_params(species)
print_config(species)

# define preprocessing output directory (from 02_ImpulseDE2.Rmd & 03_WGCNA.Rmd)
impulse_dir <- file.path("../../output_RNA/ImpulseDE2", species)
mfuzz_dir <- file.path(impulse_dir, "Mfuzz")
wgcna_dir <- file.path("../../output_RNA/WGCNA", species)

# FIMO path (from annotation repo)
fimo_path <- file.path(annot_dir, "promoters/fimo_output", paste0(species, "_stress_TFs"))
cat("\nFIMO path:", fimo_path, "\n")

# set up necessary output directories if they don't exist
outdir <- file.path("../../output_RNA/TFBS", species)
outdir_plots <- file.path(outdir,"plots")
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
if (!dir.exists(outdir_plots)) dir.create(outdir_plots, recursive = TRUE)

reportdir <- file.path("../../output_RNA/reports", params$species, "04_TFBS_files/figure-gfm/")
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)
```

## 3. Load in expression results (ImpuseDE, Mfuzz, WGCNA) and SwissProt annotations

```{r}
# ImpulseDE2 results
impulse_results <- read.csv(file.path(impulse_dir, "ImpulseDE2_results.csv"))
impulse_sig <- read.csv(file.path(impulse_dir, "ImpulseDE2_significant.csv"))

# Mfuzz cluster assignments
mfuzz_clusters <- read.csv(file.path(mfuzz_dir, "cluster_assignments.csv"))
mfuzz_clusters_mem50 <- mfuzz_clusters %>% filter(membership > 0.5)

# WGCNA module assignments
wgcna_modules <- read.csv(file.path(wgcna_dir, "gene_modules.csv"))

# Module interaction stats (to ID significant modules)
module_stats <- read.csv(file.path(wgcna_dir, "module_interaction_stats.csv"))
sig_modules <- module_stats %>% filter(adj.P.Val < 0.05) %>% pull(module)

# Combine all analyzed genes across analysis
all_genes <- unique(c(impulse_results$Gene, wgcna_modules$gene_id))

cat("ImpulseDE2 significant genes:", nrow(impulse_sig), "\n")
cat("Mfuzz clustered genes:", nrow(mfuzz_clusters), "\n")
cat("WGCNA modules:", n_distinct(wgcna_modules$module), "\n")
cat("Significant modules (treatment*time):", length(sig_modules), "\n")
```

```{r}
# SwissProt annotations
SwissProt <- read.delim(file.path(annot_dir,config$SwissProt))
cat("Annotations:", nrow(SwissProt), "Swissprot-annotated genes")
```

## 4. Load in Stress Transcription Factor Binding Site Data

```{r}
Stress_TFs <- read.delim(file.path(fimo_path,"fimo.tsv"), comment.char = "#")

#get list of motifs/TFs present in the data
all_tfs <- unique(Stress_TFs$motif_alt_id)
cat("Transcription factor binding sites found for:", paste(unique(Stress_TFs$motif_alt_id), collapse = ", "), "\n")
```


```{r}
Stress_TFs_quantification <- Stress_TFs %>%
  group_by(motif_alt_id,sequence_name) %>%
  summarize(count=n(), .groups = "drop") 

Foxo3_quantification <- Stress_TFs_quantification %>% filter(motif_alt_id =="FOXO3") %>% left_join(SwissProt, by = join_by("sequence_name"=="query")) 
Hsf1_quantification <- Stress_TFs_quantification %>% filter(motif_alt_id =="HSF1") %>% left_join(SwissProt, by = join_by("sequence_name"=="query")) 
Nrf2_quantification <- Stress_TFs_quantification %>% filter(motif_alt_id =="NFE2L2") %>% left_join(SwissProt, by = join_by("sequence_name"=="query")) 

cat("Transcription Factor", "FOXO3:", nrow(Foxo3_quantification), "genes with putative binding sites.\nGene(s) with most binding sites:",
    Foxo3_quantification %>% filter(count == max(count)) %>% pull(sequence_name), "( Best Annotation:", Foxo3_quantification %>% filter(count == max(count)) %>% arrange(evalue) %>% head(1) %>% pull(ProteinNames)  %>% word(1:4), ") with", max(Foxo3_quantification$count), "binding sites.\n\n")

cat("Transcription Factor", "HSF1:", nrow(Hsf1_quantification), "genes with putative binding sites.\nGene(s) with most binding sites:",
    Hsf1_quantification %>% filter(count == max(count)) %>% pull(sequence_name), "( Best Annotation:", Hsf1_quantification %>% filter(count == max(count)) %>% arrange(evalue) %>% head(1) %>% pull(ProteinNames)  %>% word(1:4), ") with", max(Hsf1_quantification$count), "binding sites.\n\n")

cat("Transcription Factor", "NFE2L2:", nrow(Nrf2_quantification), "genes with putative binding sites.\nGene(s) with most binding sites:",
    Nrf2_quantification %>% filter(count == max(count)) %>% pull(sequence_name), "( Best Annotation:", Nrf2_quantification %>% filter(count == max(count)) %>% arrange(evalue) %>% head(1) %>% pull(ProteinNames)  %>% word(1:4), ") with", max(Nrf2_quantification$count), "binding sites.")
```

Make dataframe that collapses genes with sites from more than one TF in Stress_TFs_quantification so that there is only one row per gene.

```{r}
TFBS_by_gene_orig <- Stress_TFs_quantification %>%
  pivot_wider(names_from = motif_alt_id, values_from = count, values_fill = 0, names_prefix = "count_")

# the above dataframe only has the genes that have at least 1 binding site for at least one of the three TFs, but I want to make sure to include all genes in my dataset in the downstream analysis, so I will add rows for genes that don't have any binding sites for these TFs with counts of 0. This also removes any genes from the TFBS dataset that aren't in our filtered gene set.

TFBS_by_gene <- data.frame(gene_id = all_genes) %>%
  left_join(TFBS_by_gene_orig, by = join_by(gene_id == sequence_name)) %>%
  replace_na(list(count_FOXO3 = 0, count_HSF1 = 0, count_NFE2L2 = 0))

TFBS_by_gene_min1BS <- TFBS_by_gene[rowSums(TFBS_by_gene %>% select(-gene_id)) > 0,]
```

## 5. Visualize relationship between TFBS count and gene expression results 

```{r}
# combine impulseDE results, Mfuzz clusters, WGCNA modules, and TFBS quantification into one master dataframe for downstream analysis and visualization
# results are only included for significant Impulse results, genes that have a membership of > 0.5 in their Mfuzz clusters, but all genes are included in the table

master_table <- TFBS_by_gene %>%
  left_join(impulse_sig %>% select(Gene, padj, response_type), by = join_by(gene_id == Gene)) %>%
  left_join(mfuzz_clusters_mem50 %>% select(Gene, cluster, membership), by = join_by(gene_id == Gene)) %>%
  left_join(wgcna_modules %>% select(gene_id, module), by = "gene_id") %>%
  dplyr::rename(ImpulseDE2_padj = padj,
         ImpulseDE2_response_type = response_type,
         Mfuzz_cluster = cluster,
         Mfuzz_membership = membership,
         WGCNA_module = module)
```

```{r}
master_table_min1BS <- master_table %>% filter(gene_id %in% TFBS_by_gene_orig$sequence_name)
```

## 6. Save outputs

```{r}
write.csv(master_table, file.path(outdir, "all_genes_combined_results.csv"), row.names = FALSE)
write.csv(master_table_min1BS, file.path(outdir, "genes_withTFBS_combined_results.csv"), row.names = FALSE)
```

