---
title: "Integrate across analyses"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
params:
  species: "Pacuta" # default for interactive use
output:
  github_document:
    toc: true
    html_preview: false
---

# Analysis of Time Series bulk RNA-seq data: Combine results from all analyses

## 1. Load packages and functions

```{r setup}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath(paste0("../../output_RNA/reports/", params$species, "/")), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 10, fig.height = 8,
                      fig.path = "05_Integration_files/figure-gfm/")

#load packages
library(tidyverse)
library(knitr)
library(ComplexHeatmap)
library(igraph)
library(ggraph)
library(ggarrow)
library(patchwork)

#load in parameters and functions
source("species_parameters.R")
source("utils.R")

sessionInfo() #provides list of loaded packages and version of R
```

## 2. Setup species-specific parameters and define directories

```{r get-species}
# get species
species <- params$species

# get parameters for this species
config <- get_params(species)
print_config(species)

# define preprocessing and analysis output directories (from 01_preprocessing.Rmd, 02_ImpulseDE2.Rmd, 03_WGCNA.Rmd, & 04_TFBS.Rmd)
input_dir <- file.path("../../output_RNA/counts_filt_norm", species)
impulse_dir <- file.path("../../output_RNA/ImpulseDE2", species)
mfuzz_dir <- file.path(impulse_dir, "Mfuzz")
wgcna_dir <- file.path("../../output_RNA/WGCNA", species)
tfbs_dir <- file.path("../../output_RNA/TFBS", species)

# set up necessary output directories if they don't exist
outdir <- file.path("../../output_RNA/analysis_integration", species)
outdir_plots <- file.path(outdir,"plots")
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
if (!dir.exists(outdir_plots)) dir.create(outdir_plots, recursive = TRUE)

reportdir <- file.path("../../output_RNA/reports", params$species, "05_Integration_files/figure-gfm/")
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)
```

## 3. Load in metadata, filtered counts, transformed counts, and expression results (ImpuseDE, Mfuzz, WGCNA, TFBS) and SwissProt annotations

```{r}
# load in species metadata
meta <- read.csv(paste0("../../output_RNA/",species,"_RNA_seq_metadata.csv"), row.names = 1)

# load in filtered counts data
filtered_counts <- read.csv(file.path(input_dir, "filtered_counts.csv"), row.names = 1)
all_genes <- rownames(filtered_counts)

# load in transformed counts data
vst_counts <- read.csv(file.path(input_dir, "vst_expression_matrix.csv"), row.names = 1)

# ImpulseDE2 results
impulse_results <- read.csv(file.path(impulse_dir, "ImpulseDE2_results.csv"))
impulse_sig <- impulse_results %>%
  filter(padj < global_params$padj_threshold)

# Mfuzz cluster assignments
mfuzz_clusters <- read.csv(file.path(mfuzz_dir, "cluster_assignments.csv"))
mfuzz_clusters_mem50 <- mfuzz_clusters %>% filter(membership > 0.5)

# WGCNA module assignments
wgcna_modules <- read.csv(file.path(wgcna_dir, "gene_modules.csv"))
hub_genes <- read.csv(file.path(wgcna_dir, "hub_genes.csv"))

# Module interaction stats (to ID significant modules)
module_stats <- read.csv(file.path(wgcna_dir, "module_interaction_stats.csv"))
sig_modules <- module_stats %>% filter(adj.P.Val < 0.05) %>% pull(module)

# TFBS Count data
tfbs <- read.csv(file.path(tfbs_dir, "TFBS_counts.csv"))

cat("Number of genes after filtering: ", nrow(filtered_counts), "\n")
cat("ImpulseDE2 significant genes:", nrow(impulse_sig), "\n")
cat("Mfuzz clustered genes:", nrow(mfuzz_clusters), "\n")
cat("Mfuzz clustered genes with membership > 0.5:", nrow(mfuzz_clusters_mem50), "\n")
cat("WGCNA modules:", n_distinct(wgcna_modules$module), "\n")
cat("Significant modules (treatment*time):", length(sig_modules), "\n")
cat("Genes with putative HSF1, FOXO3, or NRF2 transcription factor binding sites:", nrow(tfbs), "\n")
```

```{r}
# SwissProt annotations
SwissProt <- read.delim(file.path(annot_dir,config$SwissProt))
cat("Annotations:", nrow(SwissProt), "Swissprot-annotated genes")
```

```{r}
#loads the pattern mapping assessed by me after running ImpulseDE2 and comparing across species (see ../../output_RNA/ImpulseDE2/cluster_patterns.md)

Mfuzz_pattern_mapping <- NULL
source("../../output_RNA/ImpulseDE2/cluster_patterns.R")

Mfuzz_pattern_mapping <- pattern_mapping %>% filter(species ==  params$species) %>% dplyr::select(-species)
```

## 4. Combine results across analyses

```{r}
# combine impulseDE results, Mfuzz clusters, WGCNA modules, and TFBS quantification into one master dataframe for downstream analysis and visualization

master_table <- data.frame(gene_id=all_genes) %>%
  left_join(impulse_results %>% select(Gene, padj, response_type,classification), by = join_by(gene_id == Gene)) %>%
  mutate(is_DE = padj < 0.05) %>%
  left_join(mfuzz_clusters %>% select(Gene, cluster, membership), by = join_by(gene_id == Gene)) %>%
  mutate(Mfuzz_highconf = membership > 0.5) %>%
  left_join(Mfuzz_pattern_mapping, by = "cluster") %>%
  left_join(wgcna_modules %>% select(gene_id, module,kME_own), by = "gene_id") %>%
  mutate(is_hub = gene_id %in% hub_genes$gene_id) %>% 
  left_join(tfbs, by = join_by(gene_id == sequence_name)) %>%
  mutate(across(starts_with("count_"), ~ replace_na(.x, 0))) %>%
  mutate(has_HSF1 = count_HSF1 > 0, has_FOXO3 = count_FOXO3 > 0, has_NFE2L2 = count_NFE2L2 > 0) %>% 
  left_join(SwissProt, by = join_by(gene_id == query)) %>%
  dplyr::rename(ImpulseDE2_padj = padj,
         ImpulseDE2_response_type = response_type,
         ImpulseDE2_response_class = classification,
         Mfuzz_cluster = cluster,
         Mfuzz_membership = membership,
         Mfuzz_pattern = pattern,
         WGCNA_module = module,
         SwissProt_BlastHit = blast_hit,
         SwissProt_BlastEval = evalue,
         SwissProt_ProteinName = ProteinNames)
```

## 5. Visualize comparisons and overlaps across analyses

### Summary table

```{r}
summary_stats <- tibble(
  metric = c(
    "Total genes (post-filtering)",
    "SwissProt-annotated genes",
    "Differentially expressed (ImpulseDE2, padj < 0.05)",
    "DE genes temporally clustered (Mfuzz, membership > 0.5)",
    "WGCNA assigned",
    "WGCNA modules",
    "WGCNA modules Sig. by Time*Treatment",
    "Hub genes (WGCNA, top 10% kME)",
    "Hub genes also differentially expressed (ImpulseDE2)",
    "Genes with HSF1 binding sites",
    "Genes with FOXO3 binding sites",
    "Genes with NRF2/NFE2L2 binding sites"
  ),
  count = c(
    nrow(master_table),
    sum(!is.na(master_table$SwissProt_ProteinName)),
    sum(master_table$is_DE, na.rm = TRUE),
    sum(master_table$Mfuzz_highconf, na.rm = TRUE),
    sum(!is.na(master_table$WGCNA_module)),
    length(unique(master_table$WGCNA_module)),
    sum(unique(master_table$WGCNA_module) %in% sig_modules),
    sum(master_table$is_hub),
    sum(master_table$is_DE & master_table$is_hub, na.rm = TRUE),
    sum(master_table$has_HSF1),
    sum(master_table$has_FOXO3),
    sum(master_table$has_NFE2L2)
  )
)

summary_stats %>% kable(format = "markdown")
```

### Visualize overlaps with upset plot

First make a true/false matrix for columns of interest

- is_DE = is differentially expressed (padj < 0.05) by ImpulseDE2
- Mfuzz_highconf = of the DE genes, these had a membership of > 0.5 in their Mfuzz cluster
- WGCNA_sig_module = member of a WGCNA module that was signfiicant for the interaction between temperature and timepoint by limma
- is_hub = is a hub gene for its module
- Has_TF_binding = has at least one match for the searched transcription factor bindinf motifs (HSF1, FOXO3, or NFE2L2/NRF2)
- SwissProt_annot = has a Swissprot annotation match

```{r}
overlap_matrix <- master_table %>%
  mutate(Mfuzz_highconf = replace_na(Mfuzz_highconf, FALSE),
         WGCNA_sig_module = WGCNA_module %in% sig_modules,
         Has_TF_binding = has_HSF1 | has_FOXO3 | has_NFE2L2,
         SwissProt_annot = !is.na(SwissProt_BlastHit)) %>%
  select(gene_id, is_DE, Mfuzz_highconf,
         WGCNA_sig_module, is_hub,
         Has_TF_binding, SwissProt_annot) 
```

Upset plot from ComplexHeatmap package, documentation [here](https://github.com/jokergoo/ComplexHeatmap-reference/blob/master/book/08-upset.md)

```{r}
m <- make_comb_mat(as.matrix(overlap_matrix[,-1]))

pdf(file.path(outdir_plots, "pdf_figs/analysis_overlap_upset.pdf"), width=12, height=6)
UpSet(m, 
      top_annotation = upset_top_annotation(m, add_numbers = TRUE),
      right_annotation = upset_right_annotation(m, add_numbers = TRUE),
      comb_order = order(comb_size(m), decreasing = TRUE))
dev.off()

png(file.path(outdir_plots, "analysis_overlap_upset.png"),width = 3000, height = 1500, res = 300)
UpSet(m, 
      top_annotation = upset_top_annotation(m, add_numbers = TRUE),
      right_annotation = upset_right_annotation(m, add_numbers = TRUE),
      comb_order = order(comb_size(m), decreasing = TRUE))
dev.off()

UpSet(m, 
      top_annotation = upset_top_annotation(m, add_numbers = TRUE),
      right_annotation = upset_right_annotation(m, add_numbers = TRUE),
      comb_order = order(comb_size(m), decreasing = TRUE))
```

```{r}
# Fisher's exact test: DE vs TF binding
contingency_DE_TF <- table(
  DE = overlap_matrix$is_DE,
  TF = overlap_matrix$Has_TF_binding)

fisher.test(contingency_DE_TF)

# DE vs Hub genes  
contingency_DE_Hub <- table(
  DE = overlap_matrix$is_DE,
  Hub = overlap_matrix$is_hub)

fisher.test(contingency_DE_Hub)

# DE vs Significant module
contingency_DE_SigMod <- table(
  DE = overlap_matrix$is_DE,
  SigModule = overlap_matrix$WGCNA_sig_module)

fisher.test(contingency_DE_SigMod)

# TF binding vs Hub genes
contingency_TF_Hub <- table(
  TF = overlap_matrix$Has_TF_binding,
  Hub = overlap_matrix$is_hub)

fisher.test(contingency_TF_Hub)
```

## 6. Save results

```{r}
# Most important: master gene table, one row per gene with all info
write.csv(master_table, file.path(outdir,"master_gene_table.csv"),row.names = FALSE)

# Also helpful: make lists of gene ids that may be useful downstream

gene_sets <- list(
  all_genes = master_table$gene_id,
  
  # ImpulseDE
  DE_all = master_table %>% filter(is_DE) %>% pull(gene_id),
  DE_transient = master_table %>% filter(is_DE, ImpulseDE2_response_type == "Transient") %>% pull(gene_id),
  
  # Mfuzz patterns (high membership)
  early_response = master_table %>% 
    filter(grepl("early", Mfuzz_pattern,ignore.case = TRUE), Mfuzz_membership > 0.5) %>% 
    pull(gene_id),
  sustained_response = master_table %>% 
    filter(grepl("sustained", Mfuzz_pattern,ignore.case = TRUE), Mfuzz_membership > 0.5) %>% 
    pull(gene_id),
  cluster_1 = master_table %>% 
    filter(Mfuzz_cluster=="1", Mfuzz_membership > 0.5) %>% pull(gene_id),
  cluster_2 = master_table %>% 
    filter(Mfuzz_cluster=="2", Mfuzz_membership > 0.5) %>% pull(gene_id),
  cluster_3 = master_table %>% 
    filter(Mfuzz_cluster=="3", Mfuzz_membership > 0.5) %>% pull(gene_id),
  cluster_4 = master_table %>% 
    filter(Mfuzz_cluster=="4", Mfuzz_membership > 0.5) %>% pull(gene_id),
  cluster_5 = master_table %>% 
    filter(Mfuzz_cluster=="5", Mfuzz_membership > 0.5) %>% pull(gene_id),
  cluster_6 = master_table %>% 
    filter(Mfuzz_cluster=="6", Mfuzz_membership > 0.5) %>% pull(gene_id),
  
  # WGCNA genes
  hub_genes = master_table %>% filter(is_hub) %>% pull(gene_id),
  sig_module_genes = master_table %>% filter(WGCNA_module %in% sig_modules) %>% pull(gene_id),
  
  # TFBS analysis
  HSF1_targets = master_table %>% filter(has_HSF1) %>% pull(gene_id),
  FOXO3_targets = master_table %>% filter(has_FOXO3) %>% pull(gene_id),
  NFE2L2_targets = master_table %>% filter(has_NFE2L2) %>% pull(gene_id),
  
  # Combined interesting sets
  DE_hub = master_table %>% filter(is_DE, is_hub) %>% pull(gene_id),
  DE_with_TF = master_table %>% 
    filter(is_DE, has_HSF1 | has_FOXO3 | has_NFE2L2) %>% 
    pull(gene_id),
  hub_with_TF = master_table %>% 
    filter(is_hub, has_HSF1 | has_FOXO3 | has_NFE2L2) %>% 
    pull(gene_id)
)

# Save gene sets
saveRDS(gene_sets, file.path(outdir, "gene_sets.rds"))
```

## 7. Heat stress genes

```{r}
# make time and treatment factors
meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
meta$treatment <- factor(meta$treatment)
```

```{r}
HeatStressGenes <- read_csv(paste0(annot_dir,"/heatstress/HeatStressGenes_", species ,".csv")) %>%
    dplyr::select(-1) %>%
    dplyr::rename(query = paste0(species,"_gene")) %>%
    dplyr::select(query,everything())
  
  # rename columns
  HeatStressGenes <- HeatStressGenes %>%
    dplyr::rename(ref_species=species) %>% 
    dplyr::rename(gene_sym=gene_id) %>% 
    dplyr::rename(gene_id=query)
  
  # keep only expressed genes
  HeatStressGenes <- HeatStressGenes %>% filter(gene_id %in% all_genes)

  assign(paste0("heatstress_",species),HeatStressGenes)
  
  HeatStressGenes_unique <- HeatStressGenes %>% group_by(gene_id) %>%
  summarize(gene_sym = paste(unique(gene_sym), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")
            ) 
  
  HeatStressGenes_unique <- HeatStressGenes_unique %>% mutate(gene_sym=str_replace(str_replace(gene_sym,"Hsc71,HSP70","HSP70"),"HSP70,Hsc71","HSP70"))
```

```{r}
stress_genes_ids <- unique(HeatStressGenes_unique$gene_id) 
stress_genes_counts <- vst_counts[stress_genes_ids, ]

plot_df <- as.data.frame(t(stress_genes_counts)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(stress_genes_ids), names_to="gene_id", values_to="expression") %>%
  left_join(HeatStressGenes_unique)

plot_df %>% ggplot(aes(x=time, y=expression, color=gene_sym, group=gene_sym)) +
  stat_summary(fun="mean", geom="line") +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(treatment~response_type) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("Type1", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,gene_id))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_sym, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 1 Expressed Response genes")

save_ggplot(last_plot(), "All_Type1")

plot_df %>% filter(grepl("Type2", response_type)) %>%
  ggplot(aes(x=time, y=expression, color=treatment, group=interaction(treatment,gene_id))) +
  stat_summary(fun="mean", geom="line", alpha=0.6) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2,linewidth=0.5, alpha=0.6) +
  scale_color_manual(values = treat_colors) +
  facet_wrap(~gene_sym, ncol= 3,scales="free_y") +
  theme_bw() +
  theme(
    strip.text = element_text(face="bold", size=8),panel.spacing = unit(0.4, "lines")
  ) +
  labs(y="VST expression", x="Timepoint", title = "Type 2 Expressed Response genes")

save_ggplot(plot = last_plot(), filename = "All_Type2")

plot_df %>% filter(grepl("HSP",gene_sym)|grepl("Nrf2",gene_sym)|grepl("HSF1",gene_sym)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(gene_id,"Pocillopora_acuta_HIv2___",""), ": ", gene_sym)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")

plot_df %>% filter(grepl("BNIP",gene_sym)|grepl("HIF",gene_sym)) %>% ggplot(aes(x=time, y=expression, color=treatment, group=treatment)) +
  stat_summary(fun="mean", geom="line") +
  scale_color_manual(values = treat_colors) +
  stat_summary(fun.data=mean_se, geom="errorbar", width=0.2) +
  facet_wrap(~paste0(str_replace(gene_id,"Pocillopora_acuta_HIv2___",""), ": ", gene_sym)) +
  theme_bw() +
  labs(y="VST expression", x="Timepoint")
```

### heat stress genes stats

```{r}
heat_res <- HeatStressGenes_unique %>% left_join(master_table, by="gene_id")

heat_res <- heat_res %>% mutate(gene_sym=str_replace(str_replace(gene_sym,"Hsc71,HSP70","HSP70"),"HSP70,Hsc71","HSP70"))
```

### Summary table

```{r}
summary_stats <- tibble(
  metric = c(
    "Total genes (post-filtering)",
    "SwissProt-annotated genes",
    "Differentially expressed (ImpulseDE2, padj < 0.05)",
    "DE genes temporally clustered (Mfuzz, membership > 0.5)",
    "WGCNA assigned",
    "WGCNA modules",
    "WGCNA modules Sig. by Time*Treatment",
    "Hub genes (WGCNA, top 10% kME)",
    "Hub genes also differentially expressed (ImpulseDE2)",
    "Genes with HSF1 binding sites",
    "Genes with FOXO3 binding sites",
    "Genes with NRF2/NFE2L2 binding sites"
  ),
  count = c(
    nrow(heat_res),
    sum(!is.na(heat_res$SwissProt_ProteinName)),
    sum(heat_res$is_DE, na.rm = TRUE),
    sum(heat_res$Mfuzz_highconf, na.rm = TRUE),
    sum(!is.na(heat_res$WGCNA_module)),
    length(unique(heat_res$WGCNA_module)),
    sum(unique(heat_res$WGCNA_module) %in% sig_modules),
    sum(heat_res$is_hub),
    sum(heat_res$is_DE & heat_res$is_hub, na.rm = TRUE),
    sum(heat_res$has_HSF1),
    sum(heat_res$has_FOXO3),
    sum(heat_res$has_NFE2L2)
  )
)

summary_stats %>% kable(format = "markdown")
```

## 8. Pathway visualization

### Set up pathway

```{r}
# define edges - each gene needs to be paired with a target
edges <- data.frame(
  from = c("HSF1",  "HSF1",  "HSP90",    "HIF-a", "BNIP3", "Bcl-2", "Bcl-2"),
  to   = c("HSP90", "HSP70", "HIF-a", "BNIP3",    "Bcl-2",  "BAK",  "BAX"),
  int_type   = c("activate", "activate", "activate", "activate",    "inhibit",  "inhibit",  "inhibit")
)

# Create graph
g <- graph_from_data_frame(edges, directed = TRUE)

# use ggraph to get coordinates for basic tree layout of graph
basic_layout <- ggraph(g, layout = "tree")$data %>% select(x, y, name)

edges_detailed <- edges %>% 
  mutate(edge_id = paste0(from, "-", to)) %>% 
  mutate(from = as.character(from)) %>% 
  mutate(to = as.character(to)) %>% 
  inner_join(basic_layout, by = c("from" = "name")) %>% 
  rename(
    temx = x, 
    temy = y,
  ) %>% 
  inner_join(basic_layout, by = c("to" = "name")) %>% 
  rename(
    x = temx,
    y = temy,
    xend = x,
    yend = y
  )
```

#### VST H vs. C

```{r}
pathway_genes <- HeatStressGenes_unique %>%
  filter(gene_sym %in%
           c("HSF1", "HSP90","HSP70", "HIF-a","BNIP3","Bcl-2", "BAK","BAX"))

pathway_genes_ids <- unique(pathway_genes$gene_id) 
pathway_genes_counts <- vst_counts[pathway_genes_ids, ]

plot_df <- as.data.frame(t(pathway_genes_counts)) %>%
  rownames_to_column(var="sample") %>%
  left_join(meta, by=c("sample"="sample")) %>%
  pivot_longer(cols = all_of(pathway_genes_ids), names_to="gene_id", values_to="expression") %>%
  left_join(HeatStressGenes_unique)
```

```{r}
plot_df_means <- plot_df %>% group_by(gene_id, gene_sym, time, treatment) %>%
  summarize(mean_expr = mean(expression), .groups = "drop")
```

```{r}
timepoints <- c("1", "3", "12")
treatments <- c("C", "H")

plot_list <- list()

for (plot_time in timepoints) {
  for (plot_treat in treatments) {
    filtered_expr <- plot_df_means %>% filter(time==plot_time & treatment==plot_treat)
    
    nodes <- basic_layout %>% left_join(filtered_expr, by=join_by(name==gene_sym))
    
    edges_plot <- edges_detailed %>%
      mutate(int_type = factor(int_type, levels = c("activate", "inhibit")))
    
    p <- ggplot() +
      geom_label(data = nodes,
                aes(x = x, y = y,label = name,fill=mean_expr),
                size = 4) +
      geom_arrow_segment(data=edges_plot, aes(x = x, y = y-0.2,
                             xend = xend, yend = yend+0.3,
                             arrow_head = int_type, color = int_type),
                 linewidth = 1.1, alpha = 0.8) +
      scale_arrow_head_discrete(values = list(
        arrow_head_wings(offset = 30, inset = 70),
        arrow_head_line(90, lineend = "butt")),guide="none") +
      scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                           midpoint = median(plot_df_means$mean_expr),
                           limits=c(min(plot_df_means$mean_expr),
                                   max(plot_df_means$mean_expr))) +
      scale_color_manual(values = c("green3", "tomato1"),guide="none") +
      labs(arrow_head = "",
           color = "",
           fill= "VST\nExpr.",
           title = paste0(plot_time," hours, ", plot_treat)) +
      theme_void() +
      scale_x_continuous(limits = c(-1.5, 1.5)) +
      theme(legend.position = c(0.75, 0.5))
    
    plot_list[[paste0(plot_time, plot_treat)]] <- p
  }
}

wrap_plots(plot_list, ncol = 2, nrow = 3) +
  plot_layout(guides = 'collect')
```


#### VST Heat - Control 

```{r}
plot_df_diff <- plot_df_means %>%
  pivot_wider(names_from = treatment, 
              values_from = mean_expr,
              values_fill = 0) %>%
  mutate(diff = H - C) %>%
  pivot_longer(cols = c(H, C, diff),
               names_to = "treatment",
               values_to = "mean_expr")  %>% filter(treatment=="diff")
```

```{r}
timepoints <- unique(meta$time)
treatments <- c("diff")

plot_list <- list()

for (plot_time in timepoints) {
  for (plot_treat in treatments) {
    filtered_expr <- plot_df_diff %>% filter(time==plot_time & treatment==plot_treat)
    
    nodes <- basic_layout %>% left_join(filtered_expr, by=join_by(name==gene_sym))
    
    edges_plot <- edges_detailed %>%
      mutate(int_type = factor(int_type, levels = c("activate", "inhibit")))
    
    p <- ggplot() +
      geom_label(data = nodes,
                aes(x = x, y = y,label = name,fill=mean_expr),
                size = 4) +
      geom_arrow_segment(data=edges_plot, aes(x = x, y = y-0.2,
                             xend = xend, yend = yend+0.3,
                             arrow_head = int_type, color = int_type),
                 linewidth = 1.1, alpha = 0.8) +
      scale_arrow_head_discrete(values = list(
        arrow_head_wings(offset = 30, inset = 70),
        arrow_head_line(90, lineend = "butt")),guide="none") +
      scale_fill_gradient2(low = "blue", mid = "gray90", high = "red",
                           midpoint = 0,
                           limits=c(min(plot_df_diff$mean_expr)-1,max(plot_df_diff$mean_expr))) +
      scale_color_manual(values = c("green3", "tomato1"),guide="none") +
      labs(arrow_head = "",
           color = "",
           fill= "VST\nExpr.",
           title = paste0(plot_time," hours, ", plot_treat)) +
      theme_void() +
      scale_x_continuous(limits = c(-1.5, 1.5)) +
      theme(legend.position = c(0.75, 0.5))
    
    plot_list[[paste0(plot_time, plot_treat)]] <- p
  }
}

final <- wrap_plots(plot_list, ncol = 4, nrow = 2) +
  plot_layout(guides = 'collect')

final

save_ggplot(final, "HSF1_Pathway_VST_Diff")
```

