---
title: "Cross Species Analysis"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    html_preview: false
---

# Analysis of Time Series bulk RNA-seq data: Cross-Species Analysis

## Introduction

This analysis integrates expression data across three coral species (*Pocillopora acuta*, *Montipora capitata*, and *Porites compressa*) using orthologous relationships identified using the program [Broccoli](https://github.com/rderelle/Broccoli.git). In my annotation [repository](https://github.com/zdellaert/HI_genome_annotations.git) (read about this is my [README.md](https://github.com/zdellaert/TimeSeries/blob/main/4-multi-species/code/RNA/README.md)), I identified orthologous genes between these genomes, which I will use to compare expression patterns across species.

## Inputs

| Source | File |
|--------|------|
| Cross-Species Orthologs | `HI_genome_annotations/annotation/orthologs_3sp.csv` |
| Per-species | `output_RNA/analysis_integration/{species}/master_gene_table.csv` |

---

## 1. Load packages and functions

```{r setup}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath("../../output_RNA/reports/"), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 10, fig.height = 8,
                      fig.path = "06_CrossSpecies_files/figure-gfm/")

#load packages
library(tidyverse)
library(knitr)
library(ComplexHeatmap)
library(pheatmap)

#load in parameters and functions
source("species_parameters.R")
source("utils.R")

sessionInfo() #provides list of loaded packages and version of R
```

## 2. Define directories

```{r directories}
# set up necessary output directories if they don't exist
outdir <- file.path("../../output_RNA/cross_species")
outdir_plots <- file.path(outdir,"plots")
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
if (!dir.exists(outdir_plots)) dir.create(outdir_plots, recursive = TRUE)

reportdir <- "../../output_RNA/reports/06_CrossSpecies_files/figure-gfm/"
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)
```

## 3. Load in ortholog data

```{r}
orthologs <- read_csv(file.path(annot_dir,"orthologs_3sp.csv"))

orthologs %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    n_in_all_3 = sum(OG_in_all_3),
    pct_in_all_3 = 100*mean(OG_in_all_3),
    n_1to1to1 = sum(OG_1to1to1),
    pct_1to1to1 = 100*mean(OG_1to1to1),
    .groups = 'drop'
  ) %>% kable(format = "markdown")
```



## 4. Load in integrated expression results for all three species

```{r}
for (species in species_list){
  integration_dir <- file.path("../../output_RNA/analysis_integration", species)
  master_table <- read.csv(file.path(integration_dir, "master_gene_table.csv")) %>% mutate(species=species)
  
  assign(paste0(species,"_master"),master_table)
  rm(master_table)
}

tables <- paste0(species_list,"_master")

# Combine
all_master <- bind_rows(Pacuta_master,Mcap_master,Pcomp_master)

cat("Genes per species:\n")
print(all_master %>% dplyr::count(species))
```

## 5. Join orthologs with expression data

```{r}
cross_species <- all_master %>%
  left_join(orthologs, by = c("gene_id", "species"))

write_csv(cross_species, file.path(outdir, "cross_species_integrated_results.csv"))
```

### Species-level summaries of orthogroups:

- n_genes = how many genes for this species are represented in the cross_species dataset (should be the same number as in master_gene_table.csv for each species since we are only adding orthogroup data for genes we have expression data for)
- n_in_OG = of these genes, how many are in an orthogroup
- n_in_all_3 = of these genes, how many are in an orthogroup that has genes from all three species in it
- n_1to1to1 = of these genes, how many are in an orthogroup that has exactly one gene from all three species in it (though, all of these might not be in this table if one of the species had this gene filtered out during expression filtering)

```{r}
# quick summary
cross_species %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    n_in_OG = sum(!is.na(OG)),
    pct_in_OG = 100*mean(!is.na(OG)),
    n_in_all_3 = sum(OG_in_all_3, na.rm = TRUE),
    pct_in_all_3 = 100*mean(OG_in_all_3, na.rm = TRUE),
    n_1to1to1 = sum(OG_1to1to1, na.rm = TRUE),
    pct_1to1to1 = 100*mean(OG_1to1to1, na.rm = TRUE)
  ) %>% kable(format = "markdown")
```

## 6. Filter to complete 1:1:1 ortholog groups

Get list of one-to-one-to-one orthogroups for which all three species are represented in the *filtered* data

```{r}
complete_1to1_ogs <- cross_species %>%
  filter(OG_1to1to1) %>%
  group_by(OG) %>%
  filter(n_distinct(species) == 3) %>%
  ungroup()

write_csv(complete_1to1_ogs, file.path(outdir, "filtered_orthogroups_1to1to1.csv"))
```

### Species-level summaries of 1:1:1 orthogroups:

- n_genes = how many genes for this species are represented in the 1:1:1 ortholog dataset (post-expression filtering)
- n_DE = of these genes, how many are DE by ImpulseDE2
- n_in_clusters = of these genes, how many showed membership > 0.5 in their assigned Mfuzz cluster
- n_hubs = of these genes, how many are WGCNA hub genes

```{r}
# quick summary
sp_summary_1to1_ogs <- complete_1to1_ogs %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    n_DE = sum(is_DE, na.rm = TRUE),
    pct_DE = 100*mean(is_DE, na.rm = TRUE),
    n_in_clusters = sum(Mfuzz_highconf, na.rm = TRUE),
    n_hubs = sum(is_hub, na.rm = TRUE),
    .groups = 'drop'
  )

kable(sp_summary_1to1_ogs,format = "markdown")
```

#### Upset plot: 1:1:1 orthologs DE across species

Upset plot from ComplexHeatmap package, documentation [here](https://github.com/jokergoo/ComplexHeatmap-reference/blob/master/book/08-upset.md)

```{r}
de_matrix <- complete_1to1_ogs %>%
  select(OG, species, is_DE) %>%
  mutate(is_DE = replace_na(is_DE, FALSE)) %>%
  pivot_wider(names_from = species, values_from = is_DE, values_fill = FALSE) %>%
  column_to_rownames("OG") %>%
  as.matrix()

m <- make_comb_mat(de_matrix)
plot <- UpSet(m, pt_size = unit(5, "mm"), lwd = 3,
      top_annotation = upset_top_annotation(m, add_numbers = TRUE),
      right_annotation = upset_right_annotation(m, add_numbers = TRUE),
      comb_col = c("gray","#00A087","#4DBBD5","#E64B35")[comb_degree(m)+1],
      column_title = "Differential Expression of 1:1:1 Orthologs")

print(plot)

pdf(file.path(outdir_plots, "pdf_figs/DE_overlap_upset.pdf"), width=10, height=6)
  print(plot)
dev.off()

png(file.path(outdir_plots, "DE_overlap_upset.png"), width=10, height=6, units="in", res=300)
  print(plot)
dev.off()
```

## 7. Core response in 1:1:1 orthologs

This does not include any genes that have more than one paralog in a species.

```{r}
complete_1to1_ogs_summary <- complete_1to1_ogs %>%
  group_by(OG) %>%
  arrange(SwissProt_BlastEval, .by_group = TRUE) %>% 
  summarise(
    # how many of the which 1:1:1 orthologs are differentially expressed in all three species
    n_DE_species = sum(is_DE, na.rm = TRUE),
    all_species_DE = n_DE_species == 3,
    
    # what mfuzz patterns are represented
    Mfuzz_patterns = paste(sort(unique(na.omit(Mfuzz_pattern))), collapse = " | "),
    Mfuzz_patterns_unique = length(unique(na.omit(Mfuzz_pattern))),
    
    # get annotation info (we pre-sorted by evalue above, so the best annotation will be chosen)
    annotation = first(na.omit(SwissProt_ProteinName)),
    annotation_evalue = first(na.omit(SwissProt_BlastEval)),
    has_annotation = !all(is.na(SwissProt_ProteinName)),
    
    # list the gene for each species (again, these are only OGs where there is exactly one gene per species)
    gene_Pacuta = gene_id[species == "Pacuta"],
    gene_Mcap = gene_id[species == "Mcap"],
    gene_Pcomp = gene_id[species == "Pcomp"],
    
    # and also gather details for each species to go along with the summary info above
    # actual impulseDE2 p-values
    padj_Pacuta = ImpulseDE2_padj[species == "Pacuta"],
    padj_Mcap = ImpulseDE2_padj[species == "Mcap"],
    padj_Pcomp = ImpulseDE2_padj[species == "Pcomp"],
    
    # Mfuzz membership
    MfuzzM_Pacuta = Mfuzz_membership[species == "Pacuta"],
    MfuzzM_Mcap = Mfuzz_membership[species == "Mcap"],
    MfuzzM_Pcomp = Mfuzz_membership[species == "Pcomp"],
    
    # are any members of the 1:1:1 OG hub genes of their WGCNA module
    n_species_hub = sum(is_hub, na.rm = TRUE),
    hub_in_any = n_species_hub > 0,
    hub_in_all = n_species_hub == 3,
    
    # do any of the 1:1:1 OG genes have TF binding sites and are they conserved across species?
    n_species_HSF1 = sum(has_HSF1, na.rm = TRUE),
    n_species_FOXO3 = sum(has_FOXO3, na.rm = TRUE),
    n_species_NFE2L2 = sum(has_NFE2L2, na.rm = TRUE),
    HSF1_conserved = n_species_HSF1 == 3,
    FOXO3_conserved = n_species_FOXO3 == 3,
    NFE2L2_conserved = n_species_NFE2L2 == 3,
    any_TF_conserved = HSF1_conserved | FOXO3_conserved | NFE2L2_conserved,
    
    .groups = 'drop'
  ) %>%
  #finally, sort by some columns of interest so that the top rows will show 1:1:1 orthologs which are all DE, share an Mfuzz pattern, and show potential transcription factor binding site conservation
  arrange(desc(all_species_DE), Mfuzz_patterns_unique, desc(any_TF_conserved))

kable(head(complete_1to1_ogs_summary),format = "markdown")
```

### Define response categories for 1:1:1 orthologs

```{r}
# Core conserved: DE in all 3, same Mfuzz pattern
core_conserved_response <- complete_1to1_ogs_summary %>%
  filter(all_species_DE, Mfuzz_patterns_unique == 1) 

# Core divergent: DE in all 3, different Mfuzz patterns  
core_divergent_response <- complete_1to1_ogs_summary %>%
  filter(all_species_DE, Mfuzz_patterns_unique > 1)

# Variable response: DE in 1-2 species only
variable_response <- complete_1to1_ogs_summary %>%
  filter(!all_species_DE, n_DE_species >= 1)

# Not responsive
not_responsive <- complete_1to1_ogs_summary %>%
  filter(n_DE_species == 0)
```

#### Summarize the response categories:

```{r}
responses_1to1_ogs_summary <- data.frame(
  Category = c("Core Conserved", 
               "Core Divergent", 
               "Variable Response", 
               "Not Responsive"),
  Count = c(nrow(core_conserved_response), nrow(core_divergent_response), 
            nrow(variable_response), nrow(not_responsive))
) %>%
  mutate(
    Percentage = 100 * Count / sum(Count)
  )
```

1:1:1 Ortholog Response Categories:

- Core conserved: DE in all 3, same Mfuzz pattern
- Core divergent: DE in all 3, different Mfuzz patterns  
- Variable response: DE in 1-2 species only
- Not responsive: DE in 0 species

```{r}
kable(responses_1to1_ogs_summary,format = "markdown")
```

### Are TFBS conserved across DE 1:1:1 orthologs?

```{r}
core_conserved_response %>% filter(HSF1_conserved == TRUE) %>% kable(format = "markdown")

core_conserved_response %>% filter(FOXO3_conserved == TRUE) %>% kable(format = "markdown")

core_conserved_response %>% filter(NFE2L2_conserved == TRUE) %>% kable(format = "markdown")
```

### Visualizations

```{r}
pie_data <- responses_1to1_ogs_summary %>%
  mutate(Label = paste0(Category, "\nn = ", Count, " (", round(Percentage, 1), "%)"))

p_pie <- ggplot(pie_data, aes(x = "", y = Count, fill = Category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = c("#E64B35", "#F39B7F","#8491B4","gray"),
                    labels = pie_data$Label) +
  labs(title = "Composition of 1:1:1 Orthologs") +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    legend.position = "bottom",
    legend.title = element_blank(),legend.box.margin = margin_auto(10)
  )
save_ggplot(p_pie,"all_response_pie", width=8, height=8)

p_pie

pie_data_allDE <- responses_1to1_ogs_summary %>% filter(grepl("Core",Category)) %>%
  mutate(Percentage = 100 * Count / sum(Count),
         Label = paste0(Category, "\nn=", Count, " (", round(Percentage, 1), "%)"))

p_pie_allDE <- ggplot(pie_data_allDE, aes(x = "", y = Count, fill = Category)) +
  geom_bar(stat = "identity", width = 1, color = "white") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = c("#E64B35", "#F39B7F"),
                    labels = pie_data$Label) +
  labs(title = "Composition of 1:1:1 Orthologs: DE in 3 Species") +
  theme_void(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5, size = 16),
    legend.position = "bottom",
    legend.title = element_blank(),legend.box.margin = margin_auto(10)
  )

save_ggplot(p_pie_allDE,"core_response_pie", width=8, height=8)

p_pie_allDE
```

## 8. Mfuzz-pattern conserved responses (not exclusively 1:1:1)

### Sustained Up (3h) genes

```{r}
sus_up_3hr <- cross_species %>% filter(Mfuzz_pattern == "Sustained Up (3h)")

# ortholog summary
sus_up_3hr %>% group_by(species) %>%
    summarise(
    n_genes = n(),
    n_in_OG = sum(!is.na(OG)),
    pct_in_OG = 100*mean(!is.na(OG)),
    n_in_all_3 = sum(OG_in_all_3, na.rm = TRUE),
    pct_in_all_3 = 100*mean(OG_in_all_3, na.rm = TRUE),
    n_1to1to1 = sum(OG_1to1to1, na.rm = TRUE),
    pct_1to1to1 = 100*mean(OG_1to1to1, na.rm = TRUE)
  ) %>% kable(format = "markdown")

sus_up_3hr %>% group_by(species) %>%
    summarise(
    n_genes = n(),
    n_DE = sum(is_DE, na.rm = TRUE),
    pct_DE = 100*mean(is_DE, na.rm = TRUE),
    Mfuzz_cluster = paste0(unique(Mfuzz_cluster, na.rm = TRUE)),
    n_high_conf = sum(Mfuzz_highconf, na.rm = TRUE),
    pct_high_conf = 100*mean(Mfuzz_highconf, na.rm = TRUE),
    n_Modules = length(unique(na.omit(WGCNA_module))),
    largest_module = names(table(WGCNA_module))[which.max(table(WGCNA_module))],
    n_hubs = sum(is_hub, na.rm = TRUE),
    has_HSF1 = sum(has_HSF1, na.rm = TRUE),
    has_FOXO3 = sum(has_FOXO3, na.rm = TRUE),
    has_NFE2L2 = sum(has_NFE2L2, na.rm = TRUE),
    .groups = 'drop'
  ) %>% kable(format = "markdown")
```

### Sustained Up (12h) genes

```{r}
sus_up_12hr <- cross_species %>% filter(Mfuzz_pattern == "Sustained Up (12h)")

# ortholog summary
sus_up_12hr %>% group_by(species) %>%
    summarise(
    n_genes = n(),
    n_in_OG = sum(!is.na(OG)),
    pct_in_OG = 100*mean(!is.na(OG)),
    n_in_all_3 = sum(OG_in_all_3, na.rm = TRUE),
    pct_in_all_3 = 100*mean(OG_in_all_3, na.rm = TRUE),
    n_1to1to1 = sum(OG_1to1to1, na.rm = TRUE),
    pct_1to1to1 = 100*mean(OG_1to1to1, na.rm = TRUE)
  ) %>% kable(format = "markdown")

sus_up_12hr %>% group_by(species) %>%
    summarise(
    n_genes = n(),
    n_DE = sum(is_DE, na.rm = TRUE),
    pct_DE = 100*mean(is_DE, na.rm = TRUE),
    Mfuzz_cluster = paste0(unique(Mfuzz_cluster, na.rm = TRUE)),
    n_high_conf = sum(Mfuzz_highconf, na.rm = TRUE),
    pct_high_conf = 100*mean(Mfuzz_highconf, na.rm = TRUE),
    n_Modules = length(unique(na.omit(WGCNA_module))),
    largest_module = names(table(WGCNA_module))[which.max(table(WGCNA_module))],
    n_hubs = sum(is_hub, na.rm = TRUE),
    has_HSF1 = sum(has_HSF1, na.rm = TRUE),
    has_FOXO3 = sum(has_FOXO3, na.rm = TRUE),
    has_NFE2L2 = sum(has_NFE2L2, na.rm = TRUE),
    .groups = 'drop'
  ) %>% kable(format = "markdown")
```

### Gradual Down genes

```{r}
grad_down <- cross_species %>% filter(Mfuzz_pattern == "Gradual Down")

# ortholog summary
grad_down %>% group_by(species) %>%
    summarise(
    n_genes = n(),
    n_in_OG = sum(!is.na(OG)),
    pct_in_OG = 100*mean(!is.na(OG)),
    n_in_all_3 = sum(OG_in_all_3, na.rm = TRUE),
    pct_in_all_3 = 100*mean(OG_in_all_3, na.rm = TRUE),
    n_1to1to1 = sum(OG_1to1to1, na.rm = TRUE),
    pct_1to1to1 = 100*mean(OG_1to1to1, na.rm = TRUE)
  ) %>% kable(format = "markdown")

grad_down %>% group_by(species) %>%
    summarise(
    n_genes = n(),
    n_DE = sum(is_DE, na.rm = TRUE),
    pct_DE = 100*mean(is_DE, na.rm = TRUE),
    Mfuzz_cluster = paste0(unique(Mfuzz_cluster, na.rm = TRUE)),
    n_high_conf = sum(Mfuzz_highconf, na.rm = TRUE),
    pct_high_conf = 100*mean(Mfuzz_highconf, na.rm = TRUE),
    n_Modules = length(unique(na.omit(WGCNA_module))),
    largest_module = names(table(WGCNA_module))[which.max(table(WGCNA_module))],
    n_hubs = sum(is_hub, na.rm = TRUE),
    has_HSF1 = sum(has_HSF1, na.rm = TRUE),
    has_FOXO3 = sum(has_FOXO3, na.rm = TRUE),
    has_NFE2L2 = sum(has_NFE2L2, na.rm = TRUE),
    .groups = 'drop'
  ) %>% kable(format = "markdown")
```

## 9. Manually-curated heat stress gene families

```{r}
for (species in species_list){
  HeatStressGenes <- read_csv(paste0(annot_dir,"/heatstress/HeatStressGenes_", species ,".csv")) %>%
    dplyr::select(-1) %>%
    dplyr::rename(query = paste0(species,"_gene")) %>%
    dplyr::select(query,everything())
  
  # rename columns
  HeatStressGenes <- HeatStressGenes %>%
    dplyr::rename(ref_species=species) %>% 
    dplyr::rename(gene_sym=gene_id) %>% 
    dplyr::rename(gene_id=query)
  
  # keep only expressed genes
  HeatStressGenes <- HeatStressGenes %>% filter(gene_id %in% all_master$gene_id)

  assign(paste0("heatstress_",species),HeatStressGenes)
  
  HeatStressGenes_unique <- HeatStressGenes %>% group_by(gene_id) %>%
  summarize(gene_sym = paste(unique(gene_sym), collapse = ","),
            response_type = paste(unique(response_type), collapse = ","),
            category = paste(unique(category), collapse = ",")
            ) 
  
  assign(paste0("heatstress_unique_",species),HeatStressGenes_unique)
  rm(HeatStressGenes)
  rm(HeatStressGenes_unique)
}

Pacuta_expanded_sp <- heatstress_unique_Pacuta %>% select(gene_id, gene_sym) %>%
  left_join(orthologs %>% select(gene_id,OG), by="gene_id") %>% 
  filter(!is.na(OG))  %>%  select(-gene_id) %>% left_join(cross_species, by="OG")  %>% distinct()

Mcap_expanded_sp <- heatstress_unique_Mcap %>% select(gene_id, gene_sym) %>%
  left_join(orthologs %>% select(gene_id,OG), by="gene_id") %>% 
  filter(!is.na(OG))  %>%  select(-gene_id) %>% left_join(cross_species, by="OG")  %>% distinct()

Pcomp_expanded_sp <- heatstress_unique_Pcomp %>% select(gene_id, gene_sym) %>%
  left_join(orthologs %>% select(gene_id,OG), by="gene_id") %>% 
  filter(!is.na(OG))  %>%  select(-gene_id) %>% left_join(cross_species, by="OG")  %>% distinct()

unique(Pacuta_expanded_sp$gene_id.x)[!(unique(Pacuta_expanded_sp$gene_id.x) %in% unique(Pcomp_expanded_sp$gene_id.y))]

all_heat <- bind_rows(Pacuta_expanded_sp,Mcap_expanded_sp,Pcomp_expanded_sp) %>% distinct()

all_heat %>% group_by(species,gene_sym) %>%
  summarize(n=n())  %>% pivot_wider(names_from = species, values_from = n) %>% kable(format = "markdown")

summary <- all_heat %>% group_by(species,gene_sym) %>%
  summarize(n=n(),
            n_DE = sum(is_DE, na.rm = TRUE),
            pct_DE = 100*mean(is_DE, na.rm = TRUE),
    n_in_clusters = sum(Mfuzz_highconf, na.rm = TRUE),
    n_hubs = sum(is_hub, na.rm = TRUE)) 

summary %>% head() %>% kable(format = "markdown")

core_conserved_response %>%
  filter(gene_Pacuta %in% all_heat$gene_id |
         gene_Mcap %in% all_heat$gene_id |
         gene_Pcomp %in% all_heat$gene_id) %>% kable(format = "markdown")

core_divergent_response %>%
  filter(gene_Pacuta %in% all_heat$gene_id |
         gene_Mcap %in% all_heat$gene_id |
         gene_Pcomp %in% all_heat$gene_id) %>% kable(format = "markdown")

core_divergent_heat <- core_divergent_response %>%
  filter(gene_Pacuta %in% all_heat$gene_id |
         gene_Mcap %in% all_heat$gene_id |
         gene_Pcomp %in% all_heat$gene_id) %>% pull(OG) 

all_heat %>% filter(OG %in% core_divergent_heat) %>% kable(format = "markdown")
```

None of these genes are in the 81 1:1:1 orthogroups that are DE in all three species and share an Mfuzz pattern. However, five of them are 1:1:1 orthogroups DE in all three species and show different patterns! Cool!!

### heat heatmaps

```{r}
for (species in species_list){
  input_dir <- file.path("../../output_RNA/counts_filt_norm", species)
  vst_counts <- read.csv(file.path(input_dir, "vst_expression_matrix.csv"), row.names = 1) 
  assign(paste0(species,"_vst"),vst_counts)
  rm(vst_counts)
}
```


```{r}
top_100_Pcomp <- all_heat %>% filter(species =="Pcomp") %>% arrange(ImpulseDE2_padj) %>% filter(gene_id %in% rownames(Pcomp_vst)) %>% head(100) 
heat_samples <- Pcomp_vst %>% select(contains("_H")) %>% colnames()
T0_samples <- Pcomp_vst %>% select(contains("_H")) %>% select(contains("_R0_"))  %>% colnames()
T
vst_heat_top <- Pcomp_vst[top_100_Pcomp$gene_id,heat_samples]
vst_heat_top$T0mean <- rowMeans(vst_heat_top[,T0_samples])
vst_heat_norm <- vst_heat_top - vst_heat_top$T0mean
vst_heat_norm <- vst_heat_norm %>% select(-T0mean)

vst_max <- max(abs(vst_heat_norm), na.rm = TRUE)
scale_colors <- colorRampPalette(c("#3c58a7","#3c58a7", "white", "#ee2e25","#ee2e25"))(100)
legend_breaks <- seq(-vst_max, vst_max, length.out = 101)

heatmap_up <- pheatmap(vst_heat_norm, 
         cluster_rows = FALSE, 
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE, 
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         breaks = legend_breaks,
         col = scale_colors,
         border_color = NA)

save_ggplot(heatmap_up, "ImpulseDE2_heatmap_UP", width = 7, height = 10, units = "in", dpi = 300)

heatmap_up
```

### ortholog heatmaps

```{r}
for (species in species_list){
  input_dir <- file.path("../../output_RNA/counts_filt_norm", species)
  vst_counts <- read.csv(file.path(input_dir, "vst_expression_matrix.csv"), row.names = 1) 
  assign(paste0(species,"_vst"),vst_counts)
  
  samples <- colnames(vst_counts)
  
  meta <- data.frame(
  sample = samples, 
  species = str_split(samples, "_", simplify = TRUE)[,1], #extract first part of sample name to get species
  time = str_replace(str_split(samples, "_", simplify = TRUE)[,2],"R", ""), #extract "R##" part to get timepoint then remove R
  replicate = str_split(samples, "_", simplify = TRUE)[,3], #extract "R##" part to get timepoint then remove R
  treatment = str_replace(str_split(samples, "_", simplify = TRUE)[,3],"\\d", "")
)
  rownames(meta) <- meta$sample
  # make time and treatment factors
  meta$time <- factor(meta$time, levels = as.character(sort(unique(as.numeric(meta$time)))))
  meta$treatment <- factor(meta$treatment)
  
  assign(paste0(species,"_meta"),meta)
  rm(vst_counts)
  rm(samples)
  rm(meta)
}
```

```{r}
Pcomp_1to1 <- complete_1to1_ogs %>% filter(species =="Pcomp") %>% arrange(OG) %>% filter(gene_id %in% rownames(Pcomp_vst))# %>% head(100) 

Pcomp_1to1_conserved <- complete_1to1_ogs %>% arrange(ImpulseDE2_padj) %>% filter(species =="Pcomp") %>% filter(gene_id %in% core_conserved_response$gene_Pcomp) %>% filter(gene_id %in% rownames(Pcomp_vst))# %>% head(100) 

Pcomp_1to1_divergent <- complete_1to1_ogs %>% arrange(ImpulseDE2_padj) %>% filter(species =="Pcomp") %>% filter(gene_id %in% core_divergent_response$gene_Pcomp) %>% filter(gene_id %in% rownames(Pcomp_vst))# %>% head(100) 

Pcomp_1to1 <- bind_rows(Pcomp_1to1_conserved,Pcomp_1to1_divergent)

heat_samples <- Pcomp_vst %>% select(contains("_H")) %>% colnames()
T0_samples <- Pcomp_vst %>% select(contains("_H")) %>% select(contains("_R0_"))  %>% colnames()
row_annot <- Pcomp_1to1 %>% column_to_rownames(var="gene_id") %>% select(ImpulseDE2_response_type) %>% dplyr::rename(response_type = ImpulseDE2_response_type)
col_annot <- Pcomp_meta[heat_samples,] %>% select(time)

vst_heat_selected <- Pcomp_vst[Pcomp_1to1$gene_id,heat_samples]
vst_heat_selected$T0mean <- rowMeans(vst_heat_selected[,T0_samples])
vst_heat_norm <- vst_heat_selected - vst_heat_selected$T0mean
vst_heat_norm <- vst_heat_norm %>% select(-T0mean)

vst_max <- max(abs(vst_heat_norm), na.rm = TRUE)
scale_colors <- colorRampPalette(c("#3c58a7","#3c58a7", "white", "#ee2e25","#ee2e25"))(100)
legend_breaks <- seq(-vst_max, vst_max, length.out = 101)

Pcomp_coreOG1to1 <- pheatmap(vst_heat_norm, 
         cluster_rows = FALSE, 
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE, 
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         breaks = legend_breaks,
         col = scale_colors,
         border_color = NA,
         gaps_col = cumsum(table(col_annot$time))[-length(table(col_annot$time))],
         annotation_row = row_annot,
         annotation_col = col_annot,
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors,
                                  "response_type" = c("Other" = "gray","Monotonous" = "#D8AF39FF", "Transient" = "#278B9AFF")))

save_ggplot(Pcomp_coreOG1to1, "Pcomp_1to1to1_coreDE_orthologs", width = 7, height = 10, units = "in", dpi = 300)

Pcomp_coreOG1to1
```

```{r}
Pacuta_1to1 <- complete_1to1_ogs %>% filter(species =="Pacuta") %>% arrange(OG) %>% filter(gene_id %in% rownames(Pacuta_vst))# %>% head(100) 

Pacuta_1to1_conserved <- complete_1to1_ogs %>% arrange(ImpulseDE2_padj) %>% filter(species =="Pacuta") %>% filter(gene_id %in% core_conserved_response$gene_Pacuta) %>% filter(gene_id %in% rownames(Pacuta_vst))# %>% head(100) 

Pacuta_1to1_divergent <- complete_1to1_ogs %>% arrange(ImpulseDE2_padj) %>% filter(species =="Pacuta") %>% filter(gene_id %in% core_divergent_response$gene_Pacuta) %>% filter(gene_id %in% rownames(Pacuta_vst))# %>% head(100) 

Pacuta_1to1 <- bind_rows(Pacuta_1to1_conserved,Pacuta_1to1_divergent)

heat_samples <- Pacuta_vst %>% select(contains("_H")) %>% colnames()
T0_samples <- Pacuta_vst %>% select(contains("_H")) %>% select(contains("_R0_"))  %>% colnames()
row_annot <- Pacuta_1to1 %>% column_to_rownames(var="gene_id") %>% select(ImpulseDE2_response_type) %>% dplyr::rename(response_type = ImpulseDE2_response_type)
col_annot <- Pacuta_meta[heat_samples,] %>% select(time)

vst_heat_selected <- Pacuta_vst[Pacuta_1to1$gene_id,heat_samples]
vst_heat_selected$T0mean <- rowMeans(vst_heat_selected[,T0_samples])
vst_heat_norm <- vst_heat_selected - vst_heat_selected$T0mean
vst_heat_norm <- vst_heat_norm %>% select(-T0mean)

vst_max <- max(abs(vst_heat_norm), na.rm = TRUE)
scale_colors <- colorRampPalette(c("#3c58a7","#3c58a7", "white", "#ee2e25","#ee2e25"))(100)
legend_breaks <- seq(-vst_max, vst_max, length.out = 101)

Pacuta_coreOG1to1 <- pheatmap(vst_heat_norm, 
         cluster_rows = FALSE, 
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE, 
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         breaks = legend_breaks,
         col = scale_colors,
         border_color = NA,
         gaps_col = cumsum(table(col_annot$time))[-length(table(col_annot$time))],
         annotation_row = row_annot,
         annotation_col = col_annot,
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors,
                                  "response_type" = c("Other" = "gray","Monotonous" = "#D8AF39FF", "Transient" = "#278B9AFF")))

save_ggplot(Pacuta_coreOG1to1, "Pacuta_1to1to1_coreDE_orthologs", width = 7, height = 10, units = "in", dpi = 300)

Pacuta_coreOG1to1
```

```{r}
Mcap_1to1 <- complete_1to1_ogs %>% filter(species =="Mcap") %>% arrange(OG) %>% filter(gene_id %in% rownames(Mcap_vst))# %>% head(100) 

Mcap_1to1_conserved <- complete_1to1_ogs %>% arrange(ImpulseDE2_padj) %>% filter(species =="Mcap") %>% filter(gene_id %in% core_conserved_response$gene_Mcap) %>% filter(gene_id %in% rownames(Mcap_vst))# %>% head(100) 

Mcap_1to1_divergent <- complete_1to1_ogs %>% arrange(ImpulseDE2_padj) %>% filter(species =="Mcap") %>% filter(gene_id %in% core_divergent_response$gene_Mcap) %>% filter(gene_id %in% rownames(Mcap_vst))# %>% head(100) 

Mcap_1to1 <- bind_rows(Mcap_1to1_conserved,Mcap_1to1_divergent)

heat_samples <- Mcap_vst %>% select(contains("_H")) %>% colnames()
T0_samples <- Mcap_vst %>% select(contains("_H")) %>% select(contains("_R0_"))  %>% colnames()
row_annot <- Mcap_1to1 %>% column_to_rownames(var="gene_id") %>% select(ImpulseDE2_response_type) %>% dplyr::rename(response_type = ImpulseDE2_response_type)
col_annot <- Mcap_meta[heat_samples,] %>% select(time)

vst_heat_selected <- Mcap_vst[Mcap_1to1$gene_id,heat_samples]
vst_heat_selected$T0mean <- rowMeans(vst_heat_selected[,T0_samples])
vst_heat_norm <- vst_heat_selected - vst_heat_selected$T0mean
vst_heat_norm <- vst_heat_norm %>% select(-T0mean)

vst_max <- max(abs(vst_heat_norm), na.rm = TRUE)
scale_colors <- colorRampPalette(c("#3c58a7","#3c58a7", "white", "#ee2e25","#ee2e25"))(100)
legend_breaks <- seq(-vst_max, vst_max, length.out = 101)

Mcap_coreOG1to1 <- pheatmap(vst_heat_norm, 
         cluster_rows = FALSE, 
         show_rownames = FALSE,
         show_colnames = FALSE,
         cluster_cols = FALSE, 
         annotation_names_row = FALSE,
         annotation_names_col = FALSE,
         breaks = legend_breaks,
         col = scale_colors,
         border_color = NA,
         gaps_col = cumsum(table(col_annot$time))[-length(table(col_annot$time))],
         annotation_row = row_annot,
         annotation_col = col_annot,
         annotation_colors = list("treatment" = treat_colors,
                                  "time" = time_colors,
                                  "response_type" = c("Other" = "gray","Monotonous" = "#D8AF39FF", "Transient" = "#278B9AFF")))

save_ggplot(Mcap_coreOG1to1, "Mcap_1to1to1_coreDE_orthologs", width = 7, height = 10, units = "in", dpi = 300)

Mcap_coreOG1to1
```

```{r}
library(patchwork)
library(ggplotify)

patch_heatmaps <- as.ggplot(Pacuta_coreOG1to1) | as.ggplot(Mcap_coreOG1to1) | as.ggplot(Pcomp_coreOG1to1)

patch_heatmaps +  plot_layout(guides = 'collect')
```



