---
title: "Cross Species Analysis"
author: "Zoe Dellaert"
date: "`r Sys.Date()`"
output:
  github_document:
    toc: true
    html_preview: false
---

# Analysis of Time Series bulk RNA-seq data: Cross-Species Analysis

## Introduction

This analysis integrates expression data across three coral species (*Pocillopora acuta*, *Montipora capitata*, and *Porites compressa*) using orthologous relationships identified using the program [Broccoli](https://github.com/rderelle/Broccoli.git). In my annotation [repository](https://github.com/zdellaert/HI_genome_annotations.git) (read about this is my [README.md](https://github.com/zdellaert/TimeSeries/blob/main/4-multi-species/code/RNA/README.md)), I identified orthologous genes between these genomes, which I will use to compare expression patterns across species.

## Inputs

| Source | File |
|--------|------|
| Cross-Species Orthologs | `HI_genome_annotations/annotation/orthologs_3sp.csv` |
| Per-species | `output_RNA/analysis_integration/{species}/master_gene_table.csv` |

---

## 1. Load packages and functions

```{r setup}
# set up file paths so that Rmd outputs can be viewed using github markdown
knitr::opts_knit$set(base.dir = normalizePath("../../output_RNA/reports/"), base.url = "./")

knitr::opts_chunk$set(echo = TRUE, message = FALSE, fig.width = 10, fig.height = 8,
                      fig.path = "06_CrossSpecies_files/figure-gfm/")

#load packages
library(tidyverse)
library(knitr)

#load in parameters and functions
source("species_parameters.R")
source("utils.R")

sessionInfo() #provides list of loaded packages and version of R
```

## 2. Define directories

```{r directories}
# set up necessary output directories if they don't exist
outdir <- file.path("../../output_RNA/cross_species")
outdir_plots <- file.path(outdir,"plots")
if (!dir.exists(outdir)) dir.create(outdir, recursive = TRUE)
if (!dir.exists(outdir_plots)) dir.create(outdir_plots, recursive = TRUE)

reportdir <- "../../output_RNA/reports/06_CrossSpecies_files/figure-gfm/"
if (!dir.exists(reportdir)) dir.create(reportdir, recursive = TRUE)
```

## 3. Load in ortholog data

```{r}
orthologs <- read_csv(file.path(annot_dir,"orthologs_3sp.csv"))

orthologs %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    n_in_all_3 = sum(OG_in_all_3),
    pct_in_all_3 = 100*mean(OG_in_all_3),
    n_1to1to1 = sum(OG_1to1to1),
    pct_1to1to1 = 100*mean(OG_1to1to1),
    .groups = 'drop'
  ) %>% kable(format = "markdown")
```



## 4. Load in integrated expression results for all three species

```{r}
for (species in species_list){
  integration_dir <- file.path("../../output_RNA/analysis_integration", species)
  master_table <- read.csv(file.path(integration_dir, "master_gene_table.csv")) %>% mutate(species=species)
  
  assign(paste0(species,"_master"),master_table)
  rm(master_table)
}

tables <- paste0(species_list,"_master")

# Combine
all_master <- bind_rows(Pacuta_master,Mcap_master,Pcomp_master)

cat("Genes per species:\n")
print(all_master %>% count(species))
```

## 5. Join orthologs with expression data

```{r}
cross_species <- all_master %>%
  left_join(orthologs, by = c("gene_id", "species"))

write_csv(cross_species, file.path(outdir, "cross_species_integrated_results.csv"))
```

### Species-level summaries of orthogroups:

- n_genes = how many genes for this species are represented in the cross_species dataset (should be the same number as in master_gene_table.csv for each species since we are only adding orthogroup data for genes we have expression data for)
- n_in_OG = of these genes, how many are in an orthogroup
- n_in_all_3 = of these genes, how many are in an orthogroup that has genes from all three species in it
- n_1to1to1 = of these genes, how many are in an orthogroup that has exactly one gene from all three species in it (though, all of these might not be in this table if one of the species had this gene filtered out during expression filtering)

```{r}
# quick summary
cross_species %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    n_in_OG = sum(!is.na(OG)),
    pct_in_OG = 100*mean(!is.na(OG)),
    n_in_all_3 = sum(OG_in_all_3, na.rm = TRUE),
    pct_in_all_3 = 100*mean(OG_in_all_3, na.rm = TRUE),
    n_1to1to1 = sum(OG_1to1to1, na.rm = TRUE),
    pct_1to1to1 = 100*mean(OG_1to1to1, na.rm = TRUE)
  ) %>% kable(format = "markdown")
```

## 6. Filter to complete 1:1:1 ortholog groups

Get list of one-to-one-to-one orthogroups for which all three species are represented in the *filtered* data

```{r}
complete_1to1_ogs <- cross_species %>%
  filter(OG_1to1to1) %>%
  group_by(OG) %>%
  filter(n_distinct(species) == 3) %>%
  ungroup()

write_csv(complete_1to1_ogs, file.path(outdir, "filtered_orthogroups_1to1to1.csv"))
```

### Species-level summaries of 1:1:1 orthogroups:

- n_genes = how many genes for this species are represented in the 1:1:1 ortholog dataset (post-expression filtering)
- n_DE = of these genes, how many are DE by ImpulseDE2
- n_in_clusters = of these genes, how many showed membership > 0.5 in their assigned Mfuzz cluster
- n_hubs = of these genes, how many are WGCNA hub genes

```{r}
# quick summary
complete_1to1_ogs %>%
  group_by(species) %>%
  summarise(
    n_genes = n(),
    n_DE = sum(is_DE, na.rm = TRUE),
    pct_DE = 100*mean(is_DE, na.rm = TRUE),
    n_in_clusters = sum(Mfuzz_highconf, na.rm = TRUE),
    n_hubs = sum(is_hub, na.rm = TRUE),
    .groups = 'drop'
  ) %>% kable(format = "markdown")
```
