---
title: "Daily_measurements"
author: "PutnamLab"
date: "7/19/2022"
output:
  html_document: default
  pdf_document: default
---

This script plots daily measurements from the CBLS Wet lab for the Putnam lab. Run this script daily after daily measurements are taken and inputted into the datasheet. 

Script updated 11/14/23 by ZD. 
Script updated 11/16/23 by JA - added upper and lower threshold lines on plots. 

Script updated 1/2/24 by JA. Added probe set to daily df so that pH is calculated based on the probe set. The data is then merged back together for the plots.

Script updated 3/11/24 by DMB. Added total alkalinity plots and upper and lower threshold lines for TA.

Script updated 9/14/24 by ZD to set Probe Set #2 as probe set for QT.

Script updated 1/30/25 by ZD to add Calcium DMs.
Script updated 3/6/25 by ZD to add Alkalinity dkH measurements.

## Load packages 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(lubridate) # used for converting 8 digit date into datetime format for R
library(RColorBrewer)
library(rmarkdown)
library(tinytex)
library(ggrepel)

## If seacarb needs to be downloaded:
#packageurl <- "https://cran.r-project.org/src/contrib/Archive/seacarb/seacarb_3.2.tar.gz"
#install.packages(packageurl, repos = NULL, type = "source")
#install.packages("seacarb")
library(seacarb) 
```

## Load data 
```{r}
## Read in data
daily <- read.csv("../../../../CBLS_Wetlab/Data/Daily_measurements_tracking.csv")
head(daily)
tail(daily) # check to make sure data from today is there

## Swap blue_tank1 and 2 for BT 1/2
daily$Tank_ID <- ifelse(daily$Tank_ID == "blue_tank1" | daily$Tank_ID == "blue_tank2", "BT 1/2", daily$Tank_ID)

## Swap blue_tank3 and 4 for BT 3/4
daily$Tank_ID <- ifelse(daily$Tank_ID == "blue_tank3" | daily$Tank_ID == "blue_tank4", "BT 3/4", daily$Tank_ID)

## Swap quarantine_tank for QT
daily$Tank_ID <- ifelse(daily$Tank_ID == "quarantine_tank", "QT", daily$Tank_ID)

# Load TA measurements data
TA <- read.csv("../../../../CBLS_Wetlab/Data/CBLS_Aquarium_TA.csv")

# Rename the "SampleID" column to "Tank_ID"
TA.mod <- TA %>%
  rename(Tank_ID = SampleID)

# Filter rows for specific Tank_ID values ("header_tank", "BT 1/2", "BT 3/4")
selected_tanks <- c("header_tank", "BT 1/2", "BT 3/4", "QT")
filtered_data <- TA.mod %>%
  filter(Tank_ID %in% selected_tanks)

# Select only the desired columns
filtered_data_subset <- filtered_data %>%
  select(Date, Tank_ID, TA)
```

Set dates as characters (needs to be done for merging with the tris calibration file)
```{r}
daily$Date <- as.character(daily$Date)
filtered_data_subset$Date <- as.character(filtered_data_subset$Date)
daily$tris.date <- as.character(daily$tris.date)
```

Combine TA data frame and daily measurments
```{r}
daily <- merge(daily, filtered_data_subset, by = c("Date", "Tank_ID"), all.x = TRUE)
```

Probe Set 1 (for the header, all corals, and for the non-quarantine tank)
e.g., BT 3/4 and header tank and salt mixing tank
```{r}
daily.probe1 <- daily %>%
  filter(Probe.Set == "Probe1" | Probe.Set == "") 

range(na.omit(daily.probe1$Temperature_C))
range(na.omit(daily.probe1$pH_mv))
range(na.omit(daily.probe1$Salinity_psu))


daily.probe2 <- daily %>%
  filter(Probe.Set == "Probe2") 

range(na.omit(daily.probe2$Temperature_C))
range(na.omit(daily.probe2$pH_mv))
range(na.omit(daily.probe2$Salinity_psu))

```

## Calculate total pH from Probe Set 1 

Calculate the calibration curve from the Tris calibration and calculate pH on the total scale from pH.mV. 
```{r}
pHcalib<-read_csv("../../../../CBLS_Wetlab/Data/Tris_Calibration.csv")
pHcalib$tris.date<-as.character(pHcalib$tris.date)

pHSlope.probe1 <- pHcalib %>%
  group_by(tris.date) %>%
  nest() %>%
  mutate(fitpH = map(data, ~ lm(mVTris ~ TTris, data = .x))) %>%
  mutate(tidy_fit = map(fitpH, broom::tidy)) %>%
  unnest(tidy_fit) %>%
  select(tris.date, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  left_join(daily.probe1, ., by = "tris.date") %>%
  mutate(mVTris = Temperature_C * TTris + `(Intercept)`)

# range(pHSlope.probe1$Temperature_C)
# pHSlope.probe1 <- pHSlope.probe1 %>% filter(!is.na(Temperature_C)) 
# range(pHSlope.probe1$Temperature_C)
# 
# range(pHSlope.probe1$pH_mv)
# pHSlope.probe1 <- pHSlope.probe1 %>% filter(!is.na(pH_mv)) 
# range(pHSlope.probe1$pH_mv)
# 
# range(pHSlope.probe1$Salinity_psu)
# pHSlope.probe1 <- pHSlope.probe1 %>% filter(!is.na(Salinity_psu)) 
# range(pHSlope.probe1$Salinity_psu)
# 
# range(pHSlope.probe1$pH_mv)
# pHSlope.probe1 <- pHSlope.probe1 %>% filter(!is.na(pH_mv)) 
# range(pHSlope.probe1$pH_mv)
# 
# range(pHSlope.probe1$mVTris)
# pHSlope.probe1 <- pHSlope.probe1 %>% filter(!is.na(mVTris)) 
# range(pHSlope.probe1$mVTris)

pHSlope.probe1 <- pHSlope.probe1 %>%
  mutate(pH.total = seacarb::pH(Ex = pH_mv, Etris = mVTris, S=Salinity_psu, T=Temperature_C))
```

## Calculate total pH from Probe Set 2 
Probe Set 2 (for the header, all corals, and for the non-quarantine tank)
e.g., BT 1/2 while HI corals were islotaed and quarantine tank

Calculate the calibration curve from the Tris calibration and calculate pH on the total scale from pH.mV. 
```{r}
pHcalib2<-read_csv("../../../../CBLS_Wetlab/Data/Tris_Calibration_ProbeSet_2.csv")
pHcalib2$tris.date<-as.character(pHcalib2$tris.date)

pHSlope.probe2 <- pHcalib2 %>%
  group_by(tris.date) %>%
  nest() %>%
  mutate(fitpH = map(data, ~ lm(mVTris ~ TTris, data = .x))) %>%
  mutate(tidy_fit = map(fitpH, broom::tidy)) %>%
  unnest(tidy_fit) %>%
  select(tris.date, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate) %>%
  left_join(daily.probe2, ., by = "tris.date") %>%
  mutate(mVTris = Temperature_C * TTris + `(Intercept)`)

range(pHSlope.probe2$Temperature_C)
pHSlope.probe2 <- pHSlope.probe2 %>% filter(!is.na(Temperature_C))
range(pHSlope.probe2$Temperature_C)

range(pHSlope.probe2$pH_mv)
pHSlope.probe2 <- pHSlope.probe2 %>% filter(!is.na(pH_mv))
range(pHSlope.probe2$pH_mv)

range(pHSlope.probe2$Salinity_psu)
pHSlope.probe2 <- pHSlope.probe2 %>% filter(!is.na(Salinity_psu))
range(pHSlope.probe2$Salinity_psu)

range(pHSlope.probe2$pH_mv)
pHSlope.probe2 <- pHSlope.probe2 %>% filter(!is.na(pH_mv))
range(pHSlope.probe2$pH_mv)

range(pHSlope.probe2$mVTris)
pHSlope.probe2 <- pHSlope.probe2 %>% filter(!is.na(mVTris))
range(pHSlope.probe2$mVTris)

pHSlope.probe2 <- pHSlope.probe2 %>%
  mutate(pH.total = seacarb::pH(Ex = pH_mv, Etris = mVTris, S=Salinity_psu, T=Temperature_C))

```

Join Probe 1 and Probe 2 Sets
```{r}
pHSlope <- rbind(pHSlope.probe1, pHSlope.probe2)
```

Convert date to ymd for plotting 
```{r}
pHSlope <- pHSlope %>%
  filter(!Tank_ID == "quarantine_tank")

pHSlope$Date <- ymd(pHSlope$Date) # convert 8 digit date into datetime format

pHSlope <- pHSlope %>% relocate("pH.total", .after = Salinity_psu) %>%
  relocate(pH_mv, .after = pH.total)

head(pHSlope)

nutrients <- pHSlope %>% select(Date, Tank_ID, Time, Date.Time, Observer.Initials, Temperature_C,Phosphorus,Nitrite,Ammonia,Probe.Set, notes,P_Dosing_ml,N_Dosing_ml,P_N_Dosing_Time) %>% filter((!is.na(Phosphorus))|(!is.na(Nitrite))|(!is.na(Ammonia))|(!is.na(P_Dosing_ml))|(!is.na(N_Dosing_ml)))
write.csv(nutrients,"../../../../CBLS_Wetlab/Data/Nutrient_summary.csv")

# Move "TA" after "pH.total"

pHSlope <- pHSlope %>%
  select(Date, Tank_ID, Time, Date.Time, Observer.Initials, Temperature_C, Salinity_psu, pH.total,Calcium_ppm, TA_DKH, TA, pH_mv, Orion_Temp, light2, light3, light4, tris.date, Probe.Set, notes, `(Intercept)`, TTris, mVTris)
```

## Change to long format

Change data format to long format 
```{r}
pHSlope.long <-pHSlope %>% pivot_longer(cols=Temperature_C:TA,
  names_to = "metric",
  values_to = "value")

nutrients.long <-nutrients %>% pivot_longer(cols=Temperature_C:Ammonia,
  names_to = "metric",
  values_to = "value") %>% filter(!is.na(value))
```

Filter by relevant dates if needed

## Plot pre-Time Series data

```{r}
hlines_data <- list(
  data.frame(yintercept = 25, metric = "Temperature_C"), # lower threshold for temperature in C°
  data.frame(yintercept = 26, metric = "Temperature_C"), # upper threshold for temperature in C°
  data.frame(yintercept = 34, metric = "Salinity_psu"), # lower threshold for salinity in psu
  data.frame(yintercept = 35, metric = "Salinity_psu"), # upper threshold for salinity in psu
  data.frame(yintercept = 8.05, metric = "pH.total"), # lower threshold for total pH
  data.frame(yintercept = 8.2, metric = "pH.total"), # upper threshold for total pH
  data.frame(yintercept = 2000, metric = "TA"), # lower threshold for TA
  data.frame(yintercept = 2500, metric = "TA") # upper threshold for TA
    )

facet_labels <- unique(pHSlope.long$metric)
names(facet_labels) = facet_labels
facet_labels <- replace(facet_labels, which(facet_labels == "pH.total"), "pH")
facet_labels <- replace(facet_labels, which(facet_labels == "Salinity_psu"), "Salinity (psu)")
facet_labels <- replace(facet_labels, which(facet_labels == "Temperature_C"), "Temperature (ºC)")
facet_labels <- replace(facet_labels, which(facet_labels == "TA"), "Total Alkalinity")

pre_TimeSeries_data <- pHSlope.long %>%
  filter(Date <= "2025-07-31" & Date > "2024-10-01") %>% 
  filter(Tank_ID == "BT 1/2") %>% 
  filter(metric %in% c("Temperature_C","Salinity_psu","pH.total","TA"))

pre_TimeSeries_data %>%
  group_by(metric) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    sem  = sd(value, na.rm = TRUE) / sqrt(n()),
    n    = n(),
    .groups = "drop"
  )

daily_tank<- pre_TimeSeries_data %>%
  ggplot(aes(x=Date, y=value))+
  geom_point(size=2, color="gray50")+
  xlab("Date")+
  facet_grid(factor(metric,c("Temperature_C","Salinity_psu","pH.total","TA")) ~ ., scales = "free", labeller = as_labeller(facet_labels))+
  geom_hline(data = hlines_data[[1]], aes(yintercept = yintercept), linetype = "dashed") +    
  geom_hline(data = hlines_data[[2]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[3]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[4]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[5]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[6]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[7]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[8]], aes(yintercept = yintercept), linetype = "dashed") +
  theme_bw() +
  theme(text = element_text(size = 14)); daily_tank

# Save plot 
ggsave("Daily_Measurements_pre_TimeSeries.png", daily_tank, width = 20, height = 15, units = c("in"))

monthly_tank <- pre_TimeSeries_data %>%
  mutate(Month = floor_date(Date, "month")) %>%
  group_by(Month, metric) %>%
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    sem = sd(value, na.rm = TRUE) / sqrt(sum(!is.na(value))),
    .groups = "drop"
  )

monthly_plot <- ggplot(monthly_tank, aes(x = Month, y = mean_value)) +
  geom_errorbar(aes(ymin = mean_value - sem, ymax = mean_value + sem),
    width = 12,linewidth = 0.4,color = "gray50") +
  #geom_line(color = "gray30",linewidth = 0.6)+
  geom_point(size = 1.2,
    shape = 21,
    fill = "white",
    color = "gray20",
    stroke = 0.6) +
  xlab("Date") +
  ylab(NULL) +
  facet_grid(
    factor(metric, c("Temperature_C","Salinity_psu","pH.total","TA")) ~ .,
    scales = "free",labeller = as_labeller(facet_labels)) +
  geom_hline(
    data = bind_rows(hlines_data),
    aes(yintercept = yintercept),
    linetype = "dashed",
    linewidth = 0.4,color = "gray70") + 
  theme_bw(base_size = 12) +
  theme(
   # panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),

    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),

    axis.text = element_text(color = "black"),
    axis.title.x = element_text(margin = margin(t = 8)),

    panel.spacing = unit(1, "lines")
  )


monthly_plot

ggsave("Daily_Measurements_pre_TimeSeries_monthly.png", monthly_plot, width = 10, height = 8, units = c("in"))
```

```{r}
pre_TimeSeries_data <- pHSlope.long %>%
  filter(Date <= "2025-07-31" & Date > "2025-06-01") %>% 
  filter(Tank_ID == "BT 1/2") %>% 
  filter(metric %in% c("Temperature_C","Salinity_psu","pH.total","TA"))

pre_TimeSeries_data %>%
  group_by(metric) %>%
  summarise(
    mean = mean(value, na.rm = TRUE),
    sem  = sd(value, na.rm = TRUE) / sqrt(n()),
    n    = n(),
    .groups = "drop"
  )

daily_tank<- pre_TimeSeries_data %>%
  ggplot(aes(x=Date, y=value))+
  geom_point(size=2, color="gray50")+
  xlab("Date")+
  facet_grid(factor(metric,c("Temperature_C","Salinity_psu","pH.total","TA")) ~ ., scales = "free", labeller = as_labeller(facet_labels))+
  geom_hline(data = hlines_data[[1]], aes(yintercept = yintercept), linetype = "dashed") +    
  geom_hline(data = hlines_data[[2]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[3]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[4]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[5]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[6]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[7]], aes(yintercept = yintercept), linetype = "dashed") +
  geom_hline(data = hlines_data[[8]], aes(yintercept = yintercept), linetype = "dashed") +
  theme_bw() +
  theme(text = element_text(size = 14)); daily_tank

# Save plot 
ggsave("Daily_Measurements_JunJul_pre_TimeSeries.png", daily_tank, width = 10, height = 8, units = c("in"))

weekly_tank <- pre_TimeSeries_data %>%
  mutate(Week = floor_date(Date, "week")) %>%
  group_by(Week, metric) %>%
  summarise(
    mean_value = mean(value, na.rm = TRUE),
    sem = sd(value, na.rm = TRUE) / sqrt(sum(!is.na(value))),
    .groups = "drop"
  )

weekly_plot <- ggplot(weekly_tank, aes(x = Week, y = mean_value)) +
  geom_errorbar(aes(ymin = mean_value - sem, ymax = mean_value + sem),
    width = 3,linewidth = 0.4,color = "gray50") +
  #geom_line(color = "gray30",linewidth = 0.6)+
  geom_point(size = 1.2,
    shape = 21,
    fill = "white",
    color = "gray20",
    stroke = 0.6) +
  xlab("Date") +
  ylab(NULL) +
  facet_grid(
    factor(metric, c("Temperature_C","Salinity_psu","pH.total","TA")) ~ .,
    scales = "free",labeller = as_labeller(facet_labels)) +
  geom_hline(
    data = bind_rows(hlines_data),
    aes(yintercept = yintercept),
    linetype = "dashed",
    linewidth = 0.4,color = "gray70") + 
  theme_bw(base_size = 12) +
  theme(
   # panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.3),

    strip.background = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),

    axis.text = element_text(color = "black"),
    axis.title.x = element_text(margin = margin(t = 8)),

    panel.spacing = unit(1, "lines")
  )


weekly_plot

ggsave("Daily_Measurements_pre_TimeSeries_JunJul_weekly.png", weekly_plot, width = 10, height = 8, units = c("in"))
```

